Noradrenergic infraslow rhythm during sleep is the critical link between heart-rate dynamics and memory consolidation

Version 1

Abstract

Recent work shows that the brain’s arousal system remains active during sleep, with rhythmic locus coeruleus (LC) activity shaping sleep architecture and supporting memory consolidation. The LC releases norepinephrine (NE) in infraslow (∼0.02 Hz) bouts that gate NREM sleep spindles. Here, we demonstrate that heart rate (HR) fluctuations during NREM are tightly phase-locked to these NE rhythms, identifying the LC as a key driver of very-low-frequency HR variability (VLF-HRV), an understudied autonomic signal. Using optogenetics, transient LC inhibition blunts HR slowing, whereas LC activation produces rapid HR acceleration, directly linking LC output to cardiac control during sleep. We further show that infraslow HR variability is a cross-species marker of spindle-dependent memory processing. In mice, the amplitude of HR decelerations during NREM predicts spindle activity and subsequent memory performance. Remarkably, human sleepers show the same pattern: stronger VLF-HR fluctuations during NREM correspond to increased spindle expression and better overnight memory retention. These findings reveal a mechanistic pathway through which LC activity modulates autonomic physiology during sleep and identify infraslow HR variability as a non-invasive marker of brainstem function and memory-promoting sleep. Because LC degeneration occurs early in neurodegenerative disease, sleep-derived HR metrics may provide a scalable indicator of emerging neuromodulatory dysfunction.

Introduction

Heart rate variability (HRV) is a well-established, noninvasive biomarker of autonomic nervous system function and cardiovascular health. Reduced HRV is associated with increased morbidity and mortality and is frequently observed in neurological disorders such as Alzheimer’s disease, where it may reflect central autonomic network dysfunction (1, 2). Yet, despite its widespread use, decoding of heart rate signals and the central drivers of HRV remain elusive. One potential orchestrator of brain-body communication is the locus coeruleus-norepinephrine (LC-NE) system; LC is a brainstem nucleus that is part of the brain’s arousal systems. A recent discovery of an infraslow phasic pattern (∼30-50 s) of LC-NE activity during non-rapid eye movement (NREM) sleep has linked LC-NE activity with a wide range of critical cortical functions, including memory consolidation, as well as glymphatic clearance (3–5). The ascending phase of NE fluctuations is often coupled to brief periods of intrinsic arousal known as micro-arousals (MAs) (6). In contrast, the declining phase of NE is associated with an increase in sleep spindles, which are discrete oscillatory events lasting 1 to 2 seconds in the sigma band (8–15 Hz), which are implicated in memory consolidation (6–8). These LC-generated infraslow phasic fluctuations of NE lead to a microstructural organization of NREM sleep shifting between arousal and spindle-rich segments (3). Additionally, infraslow NE activity is involved in organization of ripple-dependent memory consolidation (9) and elicit vasomotion-dependent cerebrospinal flow (10), indicating a broader functional role of infraslow phasic NE fluctuations in sleep restoration. With age, LC-NE function deteriorates along with restorative NREM sleep and memory performance (11, 12). Yet, the LC brain structure and physiological markers are difficult to access in humans. Thus, there is a need to explore more approachable biomarkers for sleep-dependent memory consolidation. marker of Alzheimer’s disease. Furthermore, maintaining LC neural density prevents neurodegeneration (13). Given the reported reduction in HRV in aging and Alzheimer’s disease, suppressed

Sleep spindles, along with their role in memory consolidation, also show tight coupling with the peripheral autonomic nervous system (14, 15). Specifically, phasic bursts in heart rate (HR) during NREM sleep are preceded by surges in delta power (1-4 Hz) and sleep spindles, and this temporal coupling has been linked to memory improvements across sleep (14, 15). Evidence suggests that HR fluctuate at similar infraslow frequencies (4, 16) and in humans, phasic HR fluctuations correlate with infraslow changes in pupil diameter (17), a proxy for changes in NE levels (18, 19). These findings suggest a coupling between the central LC-NE system and HR, which is further supported by findings that HR increases are associated with MAs during NREM sleep (17, 20). However, the extent to which phasic NE infraslow fluctuations – occurring on the same timescale as HR fluctuations – impact the peripheral autonomic nervous system is not well understood. Given the noradrenergic role in spindle organization, could the peripheral autonomous systems be a read-out of the central LC-NE system and thus be a biomarker of memory consolidation and LC dysfunction?

In this paper, we used real-time fluorescent imaging of NE and HR detection in mice to investigate if phasic infraslow LC-NE activity during NREM sleep is reflected in HR and whether these brain-heart changes are functionally linked. Furthermore, we explored if HR could serve as a marker for spindle gating and memory consolidation in both mice and humans.

Results

Norepinephrine and heart rate correlate during NREM sleep

Separate areas of research has demonstrated that both NE and HR signals fluctuate every 30-50s. Here, we used causal experimental methods to explore whether these independent infraslow fluctuations may be coupled and how they were impacted by different sleep transitions (3, 4, 20–22).

Toward this goal, we expressed the fluorescent NE biosensor GRAB-NE2m(3.1) in the medial prefrontal cortex (mPFC) of adult C57BL/6 mice and recorded extracellular NE using fiber photometry simultaneously with EEG and EMG recordings covering multiple sleep–wake episodes (Fig. 1a). R-peaks of the QRS complex of the heart were extracted from the EMG (for details on HR/RR extraction, see Methods and Supplementary Fig. 1a-c) and the inter-beat-interval (RR) was calculated (23) (Supplementary Fig. 1d-e). A shorter RR interval reflects faster HR and vice-versa. HR during the NREM period was on average 493 bpm (SD ± 59) while REM HR was 519 bpm (SD ± 67) both of which was consistent with the expected 450-600 bpm, based on the literature (24–26).

Norepinephrine and heart rate correlate during NREM sleep transitions.

a. Left: Experimental setup visualizing the fiber photometric acquisition of extracellular norepinephrine (NE) levels in the medial prefrontal cortex, EEG, and EMG recordings. Right: Example trace of NE, EEG, EMG, and RR intervals (RR) during NREM sleep (blue shade), microarousals (MAs) <15 s (yellow shade), and wake (gray shade). RR ranged from 0.0980 to 0.1439 (417-612 BPM). b. Power spectral density (PSD) of RR and NE from 0-0.1 Hz. c. Cross correlation between NE and RR during transitions from NREM sleep to NREM sleep (blue), short MAs (beige), long MAs (orange), and wake (pink). These color codes were used for the rest of the figure. d. Left: mean trace with standard error of the mean (SEM) of NE with yellow background shade for parts of the trace that are quantified through area under the curve (AUC) estimations. Right: AUC for time 0 to 25 s after the NE trough with AUC of −25 to 0 as baseline correction. e. Left: mean trace with SEM of RR during sleep transitions. RR ranged from 0.105 to 0.13 (462-571 BPM). Right: RR AUC for time 0 to 25 after the NE trough with AUC of −25 to 0 as baseline correction. f. Correlation between RR AUC and NE AUC across all sleep transitions. g. Mean and SEM of EEG power bands surrounding the NE trough event marker during sleep transitions. h. AUC for time 0 to 25 after the NE trough for EEG power bands with AUC of −25 to 0 as baseline correction. i. Correlations between the AUC of RR and the EEG power bands. n=7, NREM events = 245, NREM to short MA events = 174, NREM to long MA events = 75, NREM to wake events = 66. *p < 0.05., **p < 0.01., and ***p < 0.001. All frequency domain analysis figures have visualizations based on weighted estimates (see methods). For a more detailed overview of the statistics, see section 1c-i in ‘Statistics’ under Supplementary Materials.

From the simultaneous recording of NE and EMG-derived RR, example traces suggested that there was an inverse correlation between NE and RR, where RR would decrease (rapid burst in HR) around the same time NE would increase (Fig. 1a). To highlight the potential temporal relationship between NE and RR, we conducted a frequency-domain analysis of the infraslow fluctuations in both signals (Fig. 1b, Supplementary Fig. 1e-f).

To explore the NE-RR relationship across sleep-wake transitions, we examined four arousal levels following a NE trough: continued NREM (lowest arousal), NREM to short MA (movement ≤ 4 s) transitions, NREM to long MA (movement 5 ≤ 15 s), and NREM to wake (movement > 15 s, highest arousal). Here we saw a negative relationship between NE and RR in cross correlations across arousal states (Fig. 1c). Additionally, we found that the level of NE-ascent corresponded to the extent of arousal in the sleep transitions (Fig. 1d). Similarly, when comparing the RR, we found a corresponding drop in values (faster HR) across arousal transitions (Fig. 1e). When correlating the NE and RR changes across the sleep transitions, a negative correlation was found (Fig. 1f), indicating that phasic fluctuations in NE and peripheral autonomic activity were highly correlated.

To appropriately categorize the different arousal events, we calculated the power of delta (1 - 4 Hz), theta (4 - 8 Hz), sigma (8 - 15 Hz), beta (15 - 30 Hz), gamma (80 - 100 Hz) bands from the spectrum of EEG signals surrounding NE throughs (Fig. 1g). Power in all bands, except gamma, decreased as the extent of arousal increased (Fig. 1h). Inversely, as arousal increased, gamma also increased (20). Lastly, changes in band power were correlated with the decline in RR across all sleep transitions. All EEG bands showed a positive correlation with RR, except for gamma, which was negatively correlated (Fig. 1i). Thus, these results indicate that NE-locked RR changes predicted EEG-defined arousals.

In conclusion, changes in cardiac rhythm are tightly coupled with the infraslow phasic NE fluctuations during NREM sleep, and HR accelerations may serve as a proxy for estimating temporal arousability dynamics during sleep. Interestingly, the systematic periodicity of this heart-brain coupling falls within the very low frequency (VLF) band (0-0.15 Hz for mice (27)), previously noted in the heart rate variability literature (28, 29).

Closed-loop optogenetic activation of the locus coeruleus during NREM sleep increases heart rate

After finding the inverse correlation between NE and RR in natural sleep transitions, we sought to determine the causal relationship between them. To this end, we designed a stimulation paradigm to modulate NE oscillatory frequency during NREM sleep. Specifically, we optogenetically stimulated the LC bilaterally in TH-Cre transgenic mice through Cre-dependent Channelrhodopsin (ChR2) activation. We employed a closed-loop paradigm, where LC stimulations (2 s 20 Hz (10 ms) blue laser pulses with a light intensity of 5 mW) were triggered when NE levels fell below increasing thresholds (Fig. 2a-b, Methods).

Closed-loop optogenetic activation of locus coeruleus during NREM sleep increases heart rate.

a. Experimental overview. Top: Channelrhodopsin (ChR2) was used to enable Locus Coeruleus (LC) stimulation through light activation. NE release was recorded with fiber photometry during simultaneous EEG/EMG recordings. Bottom: LC stimulation was activated when the real-time ΔF/F (%) calculations of NE levels fell below the pre-specified thresholds of −15, −10, −5, 0, and 5 ΔF/F (%) difference of the mean ΔF/F (%) from the last two minutes. b. Example trace of NE, EEG, EMG, and RR intervals (RR) from the first threshold (-15) period during NREM sleep (blue shade), microarousals (MAs) <15 s (yellow shade), and laser stimulation (light pink shade). RR ranged from 0.0940 to 0.1310 (458-638 BPM). c.-d. Mean traces with standard error of the mean (SEM) of NE (left) and RR (right) surrounding the onset of a laser burst with beige (pre-stimulation) and green (post-stimulation) background shade for all five thresholds: −15 (beige), −10 (light pink), −5 (pink), 0 (dark pink), and 5 (wine red) ΔF/F (%) difference. These color codes remain consistent throughout the figure. RR ranged from 0.1190 to 0.1370 (438 - 504 BPM). e.-d. Distributions of area under the curve (AUC) of NE (e.) and RR (d.) either 20-0 s before stimulation (left) or 0-20 s after stimulation (right) for all five thresholds. Baseline estimates were subtracted from the observed values at 20-40 s before and after stimulation respectively. g. Pre-stimulation AUC minus the post-stimulation AUC for NE (left) and RR (right) for all thresholds. h. Left: correlation of the pre-stimulation AUC minus the post-stimulation AUC for NE and RR across all thresholds. Right: correlation of the pre-stimulation AUC minus the post-stimulation AUC for NE and RR across thresholds −15, −10, and −5. i. From left to right: 1) Power spectral density (PSD) for RR between 0-0.2 Hz across all thresholds. 2) AUC of the RR PSD across all thresholds in the very low frequency (VLF) range. 3) Frequency at the peak of the PSD across thresholds for RR in the VLF range. 4) Power value at the peak of the PSD across thresholds for RR in the VLF range. Datapoint darkness is proportional to bout length. n=10 (6 ChR2). Threshold −15 events = 108 (97 ChR2), Threshold −10 events = 260 (205 ChR2), Threshold −5 events = 777 (481 ChR2), Threshold 0 events = 1444 (934 ChR2), and Threshold 5 events = 1148 (771 ChR2). Number of ChR2 bouts for frequency domain analysis were, from baseline to highest threshold: 212, 44, 41, 30, 37, and 43. *p < 0.05., **p < 0.01., and ***p < 0.001. All frequency domain analysis figures have visualizations based on weighted estimates (see Methods). For a more detailed overview of the statistics, see section 2e-i in ‘Statistics’ under Supplementary Materials.

Mean traces of NE and RR were aligned to LC stimulation onset (Fig. 2c-d). These traces suggested that as the stimulation became more frequent through higher NE thresholds, the NE baseline increased, while the RR baseline decreased (faster HR). Additionally, the magnitude of pre-stimulation NE descent and post-stimulation NE ascent was reduced as the thresholds increased, indicating less pronounced NE dynamics as LC stimulation became more frequent (Fig. 2e). A reversed pattern was observed for the RRs (Fig. 2f). These changes were consistent with subcortical arousal responses through power reductions for sigma and beta (Supplementary Fig. 2a-c) (30). When exploring the shifts in NE and RR surrounding the LC stimulations, the negative association between NE and RR plateaued beyond the 3rd stimulation threshold (also known as ‘5’, Fig. 2g-h). These findings indicate a causal relationship in which LC-mediated NE modulation affects HR; however, this relationship breaks down at extreme LC stimulation frequencies. No such effects were present in the control condition (Supplementary fig. 3a-f).

To understand if there was a mechanistic link between the PSD overlap between NE and RR observed in Fig. 1c, the PSD was calculated for RR across stimulation thresholds (Fig. 2i). As LC stimulation became more frequent (higher thresholds), the peak frequency of the VLF range increased, indicating that the VLF band might be driven by LC activations (Fig. 2i). Surprisingly, the overall amount of power for the PSD also increased without changes to the peak power level, suggesting that a wider range of fluctuations, and thus a wider frequency distribution, emerge as LC stimulation increases (Fig. 2i).

In conclusion, we demonstrate a causal relationship between LC and HR, where LC activations drive HR bursts.

Heart rate decelerations in response to sustained high norepinephrine levels

After observing the increase in AUC without an increase in peak power for the frequency distribution of the RRs during rapid LC stimulations, we investigated the underlying mechanism of this phenomenon. Although the mean trace of the RRs at higher LC stimulation thresholds did not reveal such dynamics, the variability of RRs was elevated compared to lower thresholds (Fig. 3a-b). Mean HR also increased as LC stimulation became more frequent, and this correlated with the variability of the RRs (Fig. 3c-d). Thus, HR fluctuations became more frequent with more stimulation of LC. This was also evident from individual raw RR traces (Fig. 3e).

Heart rate decelerations in response to sustained high cortical norepinephrine levels

a. Mean RR intervals (RRs) surrounded by raw RR traces for the five thresholds: −15 (beige), −10 (light pink), −5 (pink), 0 (dark pink), and 5 (wine red) ΔF/F (%) difference. Color codes in this figure remain consistent. b. Standard deviation (SD) of the RRs +/- 30s around the LC stimulation (left). Sample entropy of RRs 10 s after LC stimulation (right). c. Mean RRs +/- 30s around the LC stimulation (left). Mean heart rate (HR) measured in beats per minute (BPM) +/- 30s around the LC stimulation (right). d. Correlation between HR and SD of RRs across stimulation thresholds. e. RR example traces surrounding laser onset. f. Mean RR traces surrounding HR decelerations. g. RR value at HR decelerations. h. HR at HR decelerations. i. RR amplitude leading up to HR decelerations. j. Correlation between HR and RR value at HR decelerations across thresholds. k. Norepinephrine (NE) mean-trace surrounding HR decelerations across thresholds (left). NE mean-trace surrounding HR decelerations at thresholds 0 and 5 (right). l. Sigma power mean-trace surrounding HR decelerations across thresholds (left). Sigma power mean-trace surrounding HR decelerations at thresholds 0 and 5 (right). m. NE amplitude leading up to HR decelerations. n. NE mean values leading up to HR decelerations. o. Correlation between mean NE values and RR values at HR decelerations across thresholds. p. Sigma power amplitude leading up to HR decelerations. q. Correlation between sigma power amplitude and RR values at HR decelerations across thresholds. Color codes for laser thresholds: −15 (beige), −10 (light pink), −5 (pink), 0 (dark pink), and 5 (wine red). n=6. Threshold −15 events = 97 ChR2, Threshold - 10 events = 205, Threshold −5 events = 481 ChR2, Threshold 0 events = 934 ChR2, and Threshold 5 events = 771 ChR2. *p < 0.05., **p < 0.01., and ***p < 0.001. For a more detailed overview of the statistics, see section 3b-q in ‘Statistics’ under Supplementary Materials.

To quantify the HR decelerations, we took the maximal RR value 2-7 s after LC stimulation and baseline corrected the value to obtain their amplitude. We found that the amplitude only differed from 0 in the higher stimulation thresholds (0 and 5) and correlated with mean HR (Fig. 3 f-j). Thus, it seems that at high LC stimulation, where HR is elevated, HR decelerations occur. This may reflect compensatory decelerations in response to sustained elevated sympathetic activity and thus HR (31). To understand the potential mechanism behind the HR decelerations, we explored NE and sigma power surrounding the HR decelerations (Fig. 3k-l). As LC stimulation frequency increased, general NE levels elevated, while infraslow NE fluctuations shrank (Fig. 3m-n). Notably, larger HR decelerations correlated with a higher preceding mean NE value, further supporting the interpretation that HR decelerations occurred when NE levels were high and sustained over time (Fig. 3o). Lastly, we observed that as LC activations increased, the sigma power amplitude in the seconds leading up to the HR decelerations shifted from negative to positive, indicating that HR decelerations were associated with sigma power increases (Fig. 3p). The sigma amplitude furthermore positively correlated with the RR value at the HR peak (Fig. 3q). This indicates that HR decelerations, despite high NE levels, are still associated with neuronal synchronization in the sigma frequency range, since NE can still display subtle fast declines (Fig. 3k).

In conclusion, frequent LC activations do not lead to sustained elevation in HR; rather, the heart exhibits compensatory HR decelerations, leading to large dynamic HR shifts following sigma changes.

Heart rate responses to locus coeruleus suppression reflect memory consolidation

To further understand the causal relationship between NE and RR, we expressed a Cre-dependent inhibitory opsin, Archaerhodopsin (Arch), in the LC of TH-Cre transgenic mice, with YFP serving as a control. We measured NE levels through an optic fiber implanted in mPFC in addition to EEG/EMG measurements (Fig. 4a-b). When exploring the NE and RR response to 2 min LC suppression during NREM sleep, we observed a response in NE decline, while HR didn’t deviate (Fig. 4c-e). Surprisingly, we still observed a negative correlation between the two measures, indicating that, while HR generally did not respond strongly to LC suppression, the variability in RR responses was dependent on NE changes to the suppression (Fig. 4e). To better understand how LC suppression impacted the NE and RR oscillations, a frequency-domain analysis was conducted for NE and RR for YFP and Arch mice (Fig. 4f). When exploring the PSD for NE and RR (VLF), the analysis revealed no shifts in infraslow dynamics between the Arch or YFP condition for either NE or RR (Fig. 4f-g), suggesting little impact of LC inhibition on the phasic fluctuations.

HR modulation to LC suppression improves memory consolidation.

a. Experimental overview: Archaerhodopsin (Arch) or Yellow fluorescent protein (YFP) was injected into the locus coeruleus (LC) while GRAB-NE2m(3.1) was expressed in the medial prefrontal cortex (mPFC). Arch was stimulated through green light activation, leading to decreased norepinephrine (NE) release in the mPFC. NE release was recorded with fiber photometry during simultaneous EEG/EMG recordings. Mean HR across the recordings for the Arch group was 525 BPM (+/- 60 SD) or 0.1143 (+/- 0.0131 SD) s between beats and for the YFP group 517 BPM (+/- 61 SD) or 0.1161 (+/- 0.0137 SD) s between beats. b. Example trace of NE, EEG, EMG, and RR intervals (RR) during NREM sleep (blue shade), microarousals (MAs) <15 s (yellow shade), REM sleep (green shade), and wake (gray shade). RR ranged from 0.0949 to 0.1453 (413-632 BPM). c.-d. Mean trace with standard error of the mean (SEM) of NE (left) and RR (right) surrounding the laser onset event marker with purple shade to mark the area quantified through area under the curve (AUC) calculations for Arch-(blue) and YFP- (green) related animals. RR ranged from 0.1109 to 0.1220 (492-541 BPM). e. AUC of the NE trace (left) and RR trace (middle) under LC suppression with the two measures correlated (right) for Arch- (blue) and YFP- (green) related animals. f. Power spectral density (PSD) of NE (left) and RR (right) for YFP mice (green) and Arch mice (blue). g. From left to right: 1-2) AUC of NE PSD and RR PSD in the very low frequency (VLF) range (blue shade), 3-4) frequency at the peak of the NE and RR PSD, and 5-6) power at the peak of the NE and RR PSD for YFP mice (green) and Arch mice (blue). Datapoint darkness is proportional to bout length. h. Mean trace with SEM of theta power (left), sigma power (middle), and beta power (right) surrounding the laser onset event marker. Purple shade marks the area quantified through AUC calculations for YFP (green) and Arch (blue) animals. i. AUC for theta (left), sigma (middle), and beta (right) during laser activation for Arch- (blue) and YFP- (green) related animals along with their correlations with RR AUC for the Arch condition. j. Left: Overview of the Novel Object Recognition task. Mice were habituated to the environment before neural recordings and were then tested in the same environment with one object being replaced with a novel one. Right: Novel-to-familiar ratios correlated with RR AUC, and RR AUC correlated with NE AUC for Arch- (blue) and YFP- (green) related mice. n=8 (4 Arch). Arch events = 48, YFP events = 38. PSD bouts: Arch = 66, YFP = 43. *p < 0.05., **p < 0.01., and ***p < 0.001. All frequency domain analysis figures have visualizations based on weighted estimates (see methods). For a more detailed overview of the statistics, see section 4e-j in ‘Statistics’ under Supplementary Materials.

Lastly, we were interested in understanding the relationship between HR during LC suppression and memory consolidation. As sleep spindles are important for memory consolidation and have been found to increase in density during NE decline (3), we wanted to understand if changes in HR would correlate with post-sleep performance and thus if the HR/NE coupling during this phase is important for memory consolidation. We confirmed a sigma power upregulation and general neuronal synchronization during LC suppression. Furthermore, we found a correlation between RR responses to LC suppression and sigma power, suggesting that a stronger HR reduction response is linked to higher spindle power (Fig. 4h-i). To test whether HR changes were associated with memory outcomes, we performed a novel object recognition task. The mice were habituated to an environment with two objects prior to the LC suppression paradigm and were subsequently re-introduced to the environment after the recording, but now with one known object exchanged for a new one. By measuring approaches to the new vs. old object, the novel-to-familiar ratio (NFR) was extracted and used as a behavioral measure of memory consolidation (Fig. 4j). Interestingly, the extent of RR increases following laser onset positively correlated to memory performance. Additionally, RR and NE changes during LC suppression were negatively correlated. This indicates not only that a strong decline in HR is associated with better post-sleep performance, but also that NE and HR have a coupled response in cases where memory consolidation was at its highest.

Sigma power surges prior to heart rate burst predict next day memory performance

To further understand the cross-species relationship between shifts in autonomic activity and memory consolidation, we explored sigma power dynamics surrounding phasic HR accelerations, also known as HR bursts (HRBs, increases in HR larger than 2 or 2.2 SD of the mean in windows of 180 or 24 s for humans or mice respectively) (14, 15, 32–34), in both humans and mice. In mice, sigma power has been previously shown to inversely correlate with NE fluctuations and predict memory performance (3). As expected, NE increases in mice preceded phasic HRBs and peaks at the time of the HRB (Fig. 5b-c). Additionally, the amplitude of the NE increases predicted the strength of the HRBs (Fig. 5d). We showed that sigma power dropped directly prior to HRBs, consistent with our expectation given the negative correlation between sigma power and NE in prior work (3) (Fig. 5e-g). A similar analysis was performed in a human dataset, where participants underwent memory testing and EEG/ECG recordings overnight. While the relationship between sigma power and RR intervals were conserved across species, we observed a temporal shift in the human data (Fig. 5h). Here, the surge in sigma power preceding the HRBs was confined to the 5-s interval prior to the HRB (Fig. 5i-j), while the sigma for mice peaked in the 15-10 s range prior to the event. This suggests that sigma dynamics in mice might be slightly slower and less powerful than in humans. When analyzing the relationship between the pre-HRB sigma power and the slowing of the heart following the HRB in humans, we observe a positive correlation, suggesting the importance of the pre-HRB sigma power for HR baseline recovery (Fig. 5k). Lastly, we measured improvements in hippocampal-dependent, episodic memory through the word-pair association (WPA) task before and after a night of sleep (Fig. 5l). Sleep-dependent memory improvement (post-pre-difference) was associated with pre-HRB sigma power (baseline-corrected, p = 0.05), suggesting that surges in spindle activity linked to HRBs were important for memory consolidation (Fig. 5m).

Pre ‘Heart Rate Burst’ Sigma surges linked to improved memory.

a. Example trace showing norepinephrine (NE), sigma power, and RR-interval (RR) during NREM sleep (blue shade) and heart rate burst (HRB) events (gray lines). b.-d. Mean trace with standard error of the mean (SEM) of RR (left), NE (right), and sigma power (below) surrounding the HRB events. e-g. Correlation between the NE and RR amplitude (left), Sigma power and RR amplitude (middle), and Sigma and NE (right) amplitude. h. Example trace of human RR and sigma power during NREM stage 2 sleep (pink shade) during HRB events (gray lines). i. Mean trace of RR intervals surrounding HRB events. Purple shade represents the time during which the RRrecovery amplitude was measured. j. Mean trace of sigma power intervals surrounding HRB events. Purple shade represents the sigma power interval quantified in following sub-figures. k. Correlation between sigma power pre-HRB and RR amplitude following the HRB l. Overview of the word-paired association (WPA) memory task. m. Correlation between memory improvement and baseline corrected pre-HRB sigma power. Mice: n=7, HRB events = 920. n. Schematic diagram showing the relationship between phasic locus coeruleus and heart rate events, which are followed by norepinephrine descends and sigma power upregulations correlating with memory performance. Hypothesis figure: Humans: n=28, HRB events = 4753. *p < 0.05., **p < 0.01., and ***p < 0.001. For a more detailed overview of the statistics, see section 5d-m in ‘Statistics’ under Supplementary Materials.

In conclusion, mice and humans displayed similar autonomic-central coupling (HR and spindle dynamics surrounding HRBs) indicative of a generally conserved link between the central and autonomic arousal systems likely mediated by NE. Furthermore, the coupling between HR biomarkers and memory consolidation is associated with the coupling of these autonomic-central events, with sigma power preceding HRBs directly associated with memory consolidation improvement (Fig. 5n).

Discussion

Variability in HR is a noninvasive biomarker of autonomic nervous system function and deficits are frequently seen in aging and Alzheimer’s disease. In this study, we leveraged real-time measurements of mPFC NE levels and HR measurements from EMG recordings in mice to investigate the causal link between the two variables with high temporal resolution in freely moving sleeping mice, with similar inspection in humans. We found that HR follows the infraslow phasic fluctuation of LC-NE and that HR accelerations (or HRBs) can be elicited by direct LC activations, implicating HR as a read-out of LC-NE phasic activity. Furthermore, we demonstrated that HR is a marker of sleep quality and memory consolidation in both mice and humans (Fig. 5n), suggesting that HR can be used as a non-invasive biomarker in humans for assessing vigilance states and sleep-dependent cognitive processes.

Heart rate is modulated by infraslow norepinephrine fluctuations during NREM sleep

Utilizing HR detection and fiber photometry, we found that NE and HR correlate across NREM sleep transitions. To determine causality, we utilized TH-Cre mice with ChR2/YFP injected in the LC to activate the nucleus. We found that LC stimulations were associated with increases in HR, implying that NE levels mediate HR changes.

These findings build on Osorio-Forero et al. (35) as well as other studies that reported similar results (21, 22, 36, 37). While all these studies link HR to NE, many rely on plasma levels of NE, which, while linked to central NE (38), has low temporal resolution, making causal interpretations harder. With the use of biosensors and fiber photometry, we have a very reliable estimate of the temporal dynamics of NE changes.

The periodicity of phasic fluctuations of LC and HR overlaps and corresponds with previous findings from HR studies, namely those centered on exploring HRV. HRV is a well-established metric within the health sciences literature (28), with greater HRV consistently associated with a broad spectrum of positive health and cognitive outcomes, many of which are linked to frontal lobe functioning (39). HRV is conventionally analyzed across three primary frequency bands: high frequency (HF), low frequency (LF), and very low frequency (VLF) (28). While HF HRV and LF HRV and their relation to functioning and been characterized (28, 40, 41), the physiological mechanisms underlying VLF HRV remain poorly understood (42–44). The frequency range for VLF in humans has been reported to be ≤0.04 Hz (28, 29) but have in mice been extended to ≤0.15 Hz (27) due to the faster HR dynamics.

Here, we show that VLF HRV power overlaps with NE infraslow phasic fluctuations (0-0.1 Hz) in rodents, and recent evidence from pupil response in humans suggests that this overlap is conserved cross-species (17). Additionally, stimulating the LC pushed VLF HRV towards higher frequencies, suggesting that one potential driving force of VLF HRV is LC-mediated NE infraslow oscillations. Indeed, increased efferent sympathetic activity has been linked to increases in VLF power (41, 45). Prior findings demonstrate that the LC is linked to the autonomic nervous system through direct projections to the preganglionic cells in the sympathetic nervous system, leading to projections to the heart and more general peripheral sympathetic control (46–48). Additionally, stimulation of certain LC projections decreases parasympathetic cardiac vagal activity (49); thus, we expected to see a slowing of HR during LC suppression due to parasympathetic disinhibition. Interestingly, we find no consistent modulation in HR during LC suppression, suggesting that the LC-HR connection is more strongly associated with sympathetic activity as opposed to parasympathetic inhibition. Previous research had indicated that LF power may also represent sympathetic activity before this position was challenged (50–52), with LF and HF ratio (LF/HF) thought to reflect balance between sympathetic and parasympathetic activities (i.e., the sympatho-vagal balance) (53, 54). Our findings implicate the VLF/HF ratio as a complementary measure of sympatho-vagal balance, considering the connections of the VLF to the sympathetic branch. Further research is needed to characterize this potentially informative autonomic metric. The current data suggest that the power and frequency of LC-mediated infraslow phasic NE fluctuations may control VLF HRV either directly through sympathetic activation or indirectly by inhibition of cardiac vagal activity. Furthermore, heart rate and HRV may be a promising highly accessible read-out of LC activity, which has important clinical implications.

Heart rate as a cross-species memory marker

Spindle density, implicated in sleep-dependent memory consolidation, increases during periods of NE decline (3). Specifically, the infraslow descent of NE creates spindle-rich periods. Now we report a correlation between memory consolidation and coupled HR-NE decrease during LC suppression, suggesting that HR can be used as a biomarker of memory consolidation in sleep. This finding was extended to humans, where the sigma power surges prior to heart rate bursts (HRBs) predicted sleep-dependent memory improvement. In mice, we saw that the HRBs co-occurred with a peak in NE, which had been on the rise prior to HR changes. Since NE changes cause HR alterations, this suggests that HRBs are triggered by increases in NE signaling in mice. Thus, a similar NE increase may exist in humans (17).

Our findings also suggest that the pre-HRB surge in sigma power could play a role in memory consolidation. While sigma power overall is linked to improved memory, the sigma measure correlated to the memory estimates were baseline corrected, meaning that the timing of the sigma power relative to the HRB might be important for memory consolidation. These findings replicate past findings (14) in daytime sleep and further generalize them to overnight sleep and thus overnight memory changes.

Interestingly, when comparing mouse data to human data, we see temporal differences in the sigma power leading up to the HRB. The mouse sigma peaks around 10 s earlier and has more oscillatory dynamics as opposed to the stark rise in sigma power pre-HRB in humans. This pattern has previously been reported (16, 17) and has been attributed to species-specific differences in anatomy and physiology in the neural coupling to the heart (16). These species-specific differences could be in intrinsic cardiac neuron (ICN) properties and cardiac neural circuitry. A comparative study shows that mouse ICNs exhibit more phasic firing and direct cholinergic projections to the sinoatrial node, enabling faster and more oscillatory heart-brain coupling, while human ICNs are predominantly tonic, supporting slower, more sustained modulation (55). These anatomical and physiological differences likely contribute to the earlier and more dynamic sigma peak observed in mice compared to humans and further research efforts to map cross-species heart-brain interactions may reveal critical differences that can be used in translational techniques.

In summary, these findings imply that HR serves as a surrogate marker for NE in humans as well as a biomarker for memory consolidating processes during sleep. Loss of LC neurons occurs with general aging and is an early marker of Alzheimer’s disease (11, 12). Furthermore, maintaining LC neural density prevents neurodegeneration (13). Given the reported reduction in HRV in aging and Alzheimer’s disease (1, 2), suppressed LC responses might be responsible for the blunted autonomic cardiac modulation during sleep. The link to HRV could be further investigated by neuroimaging studies quantifying LC volume (atrophy) and determine whether the extent of LC loss could be a possible predictor of future cognitive decline through its relationship to sleep spindles. Indeed, evidence suggests that the LC activity during sleep is compromised (e.g., increased arousals, fragmented sleep, elevated cortisol) and linked to memory failure in both aging (56, 57) and Alzheimer’s disease (58).

Locus-coeruleus-mediated sympathetic control drives compensatory heart rate decelerations

A surprising finding was that diminishing the power of NE fluctuations by repetitive LC activation did not lead to less power in the HR fluctuations; rather, the opposite happened. This was caused by large decelerations in HR in conditions with repetitive LC activation and high HR. We propose that the deceleration in HR during repeated LC stimulation may be explained by a dynamic balance between heightened sympathetic control due to increased LC activity and compensatory baroreflex responses. This is likely occurring as NE activates alpha-1 receptors on the smooth muscle in blood vessels, leading to vasoconstriction and increased blood pressure, triggering the baroreflex (59, 60). This baroreflex signals to the nucleus of the solitary tract, which activates inhibitory neurons. These neurons suppress the activity of the rostral ventrolateral medulla, ultimately reducing sympathetic output and lowering the HR (61, 62). An interesting finding was that the heart rate decelerations at high NE levels were associated with increases in neuronal synchronization (i.e., increased sigma power), which seemed to be closely linked to HR decelerations and smaller NE fluctuations. It is possible that the LC receives autoinhibitory feedback from the baroreflex responses, but that this is not reflected in our cortical NE recordings.

Overall these findings suggest that infraslow phasic fluctuations of LC-NE can drive large HR fluctuations consistent with the VLF range. However, when NE levels reach artificially high and sustained levels, compensatory mechanisms such as baroreceptor reflex activation may intervene. This adaptive feedback could serve as a safety valve, reducing the coupling between central NE activity and cardiac oscillations to prevent excessive or destabilizing cardiovascular responses.

Conclusion

We show that LC-mediated NE release causes detectable changes in HR. HR appears to be more sensitive to stimulation than inhibition of the LC, indicating a potential role for NE in modulating cardiac function. These results may point toward a contribution of sympathetic activity, though further investigation is needed to determine the precise autonomic mechanisms involved. This provides an explanation of the mechanism behind slow fluctuations of HR that overlaps with the frequency of NE infraslow fluctuations. Lastly, we conclude that HR changes correlate with sigma and reflect memory consolidation processes in mice and humans. This research opens doors to unexplored possibilities of using HR as a non-invasive, cost-effective biomarker not only for sleep quality and LC-NE function, but potentially also for conditions of cognitive decline associated with sleep disturbances.

Methods

Mice

This study included female and male mice, which were group-housed according to sex with unlimited access to food and water. The housing environment had a 12-hour light/dark cycle at 21°C with 40-60% humidity. The data for all mouse-based experiments were acquired in connection to past publications (3). For experiments associated with sleep transitions, seven C57BL/6 (wild-type) male mice were used. The heterozygous TH-Cre transgenic mice used for optogenetic experiments were bred on a C57BL/6 background. Ten mice were included in the LC activation experiments, four of which were controls (M = 6), and nine were involved in the LC suppression experiments . Of the nine mice, four were put in the control condition and one Arch mouse was excluded because it took over one-third of the recording period to enter NREM (M = 3). The number of mice included in the analysis was lower than in previous publications due to the exclusion of animals with poor EMG signal quality. All experiments received approval from the Danish Animal Experiments Inspectorate and were supervised by the University of Copenhagen’s Institutional Animal Care and Use Committee (IACUC), in accordance with the European Communities Council Directive of 22 September 2010 (2010/63/EU) on the protection of animals used for scientific purposes.

Surgery

Mice underwent surgery at 7-15 weeks of age. General anesthesia was induced with 5% isoflurane and maintained at 1-3% isoflurane. Additionally, 0.05 mg/kg of preoperative buprenorphine was administered for general analgesia, and 0.03 mg/kg of lidocaine was injected at the incision site. A stereotactic frame was used to ensure stability and alignment throughout the surgery. The fur was trimmed from the upper part of the skull, and a cut was made between the ears. The skull was cleaned with ethanol, and an electric drill (Tech2000, RAM Microtorque) was used to make 3-5 burr holes in the skull following the stereotactic coordinates relative to the bregma.

In the mPFC burr hole, AAV9-hSyn-GRAB-NE2m(3.1) (see Table 1 for supplier information) was injected under the neuronal hSyn promoter (stereotactic coordinates: A/P +1.7 mm, M/L −0.3 mm, and D/V −2.00, −2.25, −2.50, and −2.75 mm, with 125 nL of virus injected at each level). Additionally, for mice included in the optogenetic experiments, an AAV5/2 (Adeno-associated virus 5/2) encoding floxed Arch (ssAAV5/2-shortCAG-dlox-Arch3.0-eGFP-dlox-WPRE-SV40p(A)) or ChR2 (ssAAV5/2-hEF1a-dlox-hChR2(H134R)-eYFP(rev)-dlox-WPRE-hGHp(A)) was injected bilaterally into the LC of the transgenic TH-Cre mice (stereotactic coordinates: A/P −5.5 mm, M/L ±0.9 mm, and D/V −3.75 mm, with 300 nL of virus injected). YFP (AAV5-EF1a-DIO-eYFP) was injected in control animals. For both virus injections, the needle was kept in place for 7 minutes after the injection (speed: 100 nL/min) before slow withdrawal.

For EMG acquisition, two wires from W3 Wires International were placed into the musculus trapezius. For EEG acquisition, two 0.8 mm stainless steel screws from NeuroTek with low impedance were screwed into holes above the cerebellum (reference point) and the frontal cortex. Contralateral to the EEG screws, a mono fiber-optic cannula (400 μm, 0.48 NA, Doric Lenses), attached to a metal ferrule (2.5 mm diameter), was placed in the mPFC at D/V −2.50 mm. For mice that had received Arch, ChR2 or YFP injections in the LC, dual fiberoptic cannulas (200 μm, 0.22 NA, Doric Lenses) were implanted bilaterally above the LC at D/V −3.65 mm. Connectors linked to the EEG screws and EMG wires, along with the cannulas, were fixed to the head with SuperBond dental cement.

Mice received 5 mg/kg of carprofen (s.c.) prior to waking up and two days after as post-op care. They were given at least two weeks post-operation for recovery and virus expression before experimentation. The surgeries were conducted following established institutional guidelines. Viral expression and injection sites were validated through immunostaining of perfused brain slices from the experimental animals (see Kjaerby et al. (3) for more information).

Fiber photometry

Two light-emitting diodes (LEDs) with wavelengths of 465 nm and 405 nm (Doric Lenses, Tucker Davis Technologies) were connected to a minicube (Doric Lenses) using attenuator patch cords (400-μm core, NA = 0.48, Doric Lenses). The fiber-optic patch cords were secured to the fiber implants on the animals’ heads through zirconia sleeves. The minicube, equipped with dichroic mirrors and cleanup filters, enabled the detection of fluorophores by matching the relevant excitation and emission spectra. The LEDs were controlled by LED drivers (Thorlabs, Doric, Tucker Davis Technologies) and interfaced with an RZ-5 or RZ10-X real-time processor (Tucker-Davis Technologies). In the mPFC, both the 465 nm and 405 nm excitation lights were transmitted through a single patch cord to activate GRAB-NE2m(3.1) fluorescence, with the 405 nm light serving as an isosbestic control wavelength to adjust for bleaching and movement-related signal fluctuations. The 465 nm and 405 nm excitations were modulated sinusoidally at 531 Hz and 211 Hz, respectively. The two modulated signals generated by the LEDs were each retrieved using standard synchronous demodulation techniques, which were implemented on the RZ-5/RZ10-X real-time processor at a sampling rate of 1000 Hz. Signal processing was managed using Synapse software (Tucker-Davis Technologies), which also synchronized the fluorescent signals with video and EEG/EMG signals via TTL pulses. The resulting data were then exported for further analysis in MATLAB R2023b (MathWorks).

To calculate ΔF/F, the 405 nm signal was fitted to the 465 nm signal using a first-degree polynomial regression (MATLAB polyfit function). The slope and intercept obtained from the regression were used to generate a scaled 405 nm control signal. This fitted control signal was then smoothed using a moving average filter (MeanFilter) with an order of 1000 to reduce noise. The ΔF/F was calculated by normalizing the 465 nm signal against the smoothed control signal, subtracting the fitted 405 nm signal from the 465 nm signal, and dividing by the fitted 405 nm signal. The resulting normalized signal was expressed as a percentage to represent the change in fluorescence relative to baseline levels.

Locus coeruleus optogenetics

To induce LC excitation, two-second stimulations consisting of 10 ms pulses at 20 Hz (resulting in a 40 ms break between pulses) at a 5-mW intensity were applied bilaterally to the LC using a 430-490 nm laser (Shanghai Laser & Optics Century Co., Ltd., model: BL473-200FC). A two-hour baseline without stimulation was implemented before and after the five-hour stimulation paradigm. The simulations were conducted in a closed loop, where stimulation was triggered if the smoothed, real-time calculated NE ΔF/F (%) crossed the threshold based on an average ΔF/F (%) value of the last two minutes. The threshold was initially set to −15 ΔF/F (%) but increased by 5 every hour until reaching the final level of +5 ΔF/F (%).

Throughout the baseline and stimulation paradigm, fiber photometry recordings from the mPFC, along with EMG and EEG, were acquired. Recordings took place during the light phase to encourage natural sleep behavior.

To induce LC inhibition, two-minute Arch stimulations at 5 mW intensity, followed by a four-minute break, were applied bilaterally to the LC using a 532 nm laser (Changchun New Industries Optoelectronics Tech. Co., Ltd., model: 16030476). Simultaneously, GRAB-NE2m(3.1) recordings were made at the mPFC site, along with the previously described EEG and EMG measurements. The two-hour recordings took place during the light phase when the mice were expected to exhibit natural sleep/wake behavior undisturbed. The mice were recorded with an infrared camera.

Novel object recognition task

The NOR task was performed on 12-20 weeks old mice involved in the LC suppression experiments in an open-field arena (PVC, custom-made, 40 x 42 x 40 cm). The mice were habituated to the environment for 10 minutes as part of the encoding phase, where the mice familiarized themselves with two objects. Following this, the animals underwent the LC suppression paradigm and were then given five minutes in the NOR task arena for the recall phase to determine the NFR. In the recall phase, one of the two known objects had been exchanged for an unknown object and the ratio of approaches to the novel vs. familiar object was recorded. Natural sleep was encouraged between the encoding and recall phases by performing the task during the light phase in an environment where the mice had been habituated for over 12 hours prior to the experiment. The mice were tracked using Ethovision XT 11.5 (Noldus).

EMG signal preprocessing

The R-peak detection algorithm was applied to EMG recordings, that were priorly subjected to filtration in frequency domain using in-house pipeline in MATLAB. First, all the signals were subjected to removal of frequencies associated with amplitude current using a bandstop infinite impulse response (IRR) Butterworth 4th order filter between frequencies of 59 and 61 Hz. In case of recordings acquired in US, the cut-off frequencies for the notch filter were of 60 to 64 Hz, and the filter order of 2 was used due to different frequency span of the nuisance signal. Subsequently, a high-pass IIR Butterworth filter with 0.6 steepness was applied to the frequencies <2.5Hz to gradually remove fluctuation of the EMG signal baseline from the recordings. This was followed with low-pass filtration >80Hz, using the same filter but of 0.9 steepness. The steepness’s of the filters were chosen to minimally induce artefacts at the border conditions, and either gradually (high-pass) or rapidly (low-pass) remove the signals that affect detecting R peaks due to either increased fluctuations of the baseline (i.e., insufficient grounding or slow motion) or high frequency noise (relatively high spectral power of muscle signals). In a few recordings, both low-pass and high-pass frequencies were adapted to be slightly above 4Hz (but not higher than 5 Hz) and 45Hz respectively, to better remove the signal baseline fluctuations or higher amplitude high-frequencies due to presence of artefacts significantly affecting the R peaks detection. This was, however, applied in recordings of low signal-to-noise ratio in the frequency range of HR signals, and was carefully evaluated for appropriateness in all cases, so that the bandwidth of the signals within the range of heart pulsations was preserved. For that purpose, at each step of the signal preprocessing, periodograms of the filtered and unfiltered signals were computed and compared to evaluate the effects of the applied filtrations. After filtration, all signals were saved in the .mat format and were used for further R peaks detection.

Heart rate detection

R-peak detection in mice was based on preprocessed EMG recordings to obtain HR without needing additionally invasive procedures. EMG is acquired in most sleep studies to aid sleep scoring, and the method has previously been validated in Lecci et al. (16) and subsequently used for other publications (4, 20). User settings for this function included adjusting the movement threshold and peak prominence to ensure optimal detection for different traces. The MATLAB function first determines the movement threshold based on the overall mean and SD of the trace. It then runs peak detection in windows of two seconds of data at a time. The threshold for the peaks is based on the mean and SD within each window to ensure peak detection even in cases of power loss. If movement is detected within a section, data that falls 100 ms before and 250 ms after the movement is not considered for peak detection due to noise and is also excluded from the peak threshold calculation within the window.

When all windows are analyzed, the timestamps at which R-peaks have been found are processed through a diff function to determine the time between peaks. The original time vector is maintained to ensure that the times of the RR are preserved for event extraction. To reduce any artifacts in the data, RR that suggested a HR faster than 800 BPM or slower than 400 BPM were removed, as previous findings suggest that the HR typically falls within 450-650 BPM throughout a 24-hour cycle (24–26). Lastly, the data and associated time vector were interpolated to 64 Hz using the p-chip method, as in Chen et al. (63), since this method was found to have the lowest error when analyzing the frequency domain (64). Interpolation was performed to obtain a consistent sampling rate, but also to get RR interval estimates during periods of movement based on the values leading up to and after the movement periods.

EEG power

For the mouse data, the power within each EEG band was determined by applying the spectrogram function to extract the short-time Fourier transformation of the EEG in one-second windows for frequencies between 0 and 100 Hz. To obtain frequency power bands, the absolute log of the short-time Fourier-transformed data was taken, and the mean values within each frequency band (slow oscillations: 0.5 - 1 Hz, delta: 1 - 4 Hz, theta: 4 - 8 Hz, sigma: 8 - 15 Hz, beta: 15 - 30 Hz, gamma (high): 80 - 100 Hz) were extracted.

Mouse sleep scoring

The mice were recorded in chambers using ViewPoint Behavior Technology, with EEG and EMG electrodes connected to a commutator (Plastics One, Bilaney, SL12C) through cables. Habituation in the recording environment lasted at least 12 hours before recordings, during which the EEG and EMG cables were connected. The mice were attached to fiber optic implants immediately before the recording, which lasted for 2–9 hours during their light phase. The EEG and EMG signals were amplified using a model 3500, 16-channel AC amplifier from National Instruments. A bandpass filter was applied to the EEG at 1-100 Hz and to the EMG at 10-100 Hz. Both signals had an added notch filter at 50 Hz to reduce power line noise. The EMG and EEG were digitized through a Multifunction I/O DAQ device from National Instruments (USB-6343) and sampled at 512 Hz. An infrared camera from FLIR Systems was used to record the mice to aid in scoring vigilance states.

Sleep scoring for mice was performed manually through visual inspection of the EEG and EMG traces, along with the video, to divide the traces into 5-second and, later, 1-second windows. Scoring was performed using SleepScore software from ViewPoint Behavior Technology. Vigilance states included ’wake’ (high-amplitude EMG and high-frequency, low-amplitude EEG), ’NREM sleep’ (low EMG amplitude and low-frequency, high-amplitude EEG), and ’REM sleep’ (low EMG amplitude and high-frequency, low-amplitude EEG). Later in the analysis pipeline, wake periods between 5-15 s were classified as long MAs, and wake periods below 5 s were scored as short microarousals.

Event marker selection

All time-domain analyses were based on the timestamps of event markers in the data. To determine the NE trough timestamps during the sleep transition analysis, similar approaches were used for the different stages. Across all stages, the NE trace was inverted, and peak detection was performed on the data. For the NE troughs during NREM, a peak prominence criterion of 0.7 to 3 was set. For NREM to MA (short and long), only data from 30 s prior to the transition were considered, and the peak prominence criterion was set to 0.8 to 1.8. Additionally, there had to be at least 15 s between troughs, with the one closest to the transition being favored. Lastly, for transitions to wake, data from 15 s prior to the scored transitions were considered. The peak prominence criterion was set to 0.1 to 0.4, and the peak height was set to a minimum of the mean of the trace minus five. After automatic trough detection, manual verification was performed across all transitions to ensure optimal trough detection.

The event marker for the data presented in Fig. 2 and 3a-b was the onset of the stimulation burst. As the laser had an online measure of NE and turned on for 2 s at 20 Hz, the event marker was defined as the point when the laser turned on after being off for over 100 ms. These timestamps were divided into groups based on when the threshold shifted. For the rest of Fig. 3, the RR peak value 2-7 s after onset of stimulation burst was utilized as the event marker for HR decelerations. The event marker for the data presented in Fig. 4 was set to when the laser turned on. For both LC suppression and excitation data, the exclusion criteria required that the animal be in NREM for 30 s before and after the laser turning on to be included.

For figure 5, the event marker was HRBs. These events have previously been reported in humans, where they are defined as when the RR intervals drop below 2 SD of the mean in a 3 min window (14, 15). This event marker was for mice adjusted to be 2.2 SD from the mean in 24 s windows. This adjustment was done based on differences in dynamics. The time window was adjusted based on the differences in mean HR between humans and mice.

Data analysis

Fiber photometry, EMG, and EEG data were extracted from all mice, and RR was calculated for each recording using the R-peak detection algorithm. For analysis pertaining to the sleep transition data, the EEG data were excluded from further analysis for two animals in the sleep transitions analysis due to noise, resulting in 245 events for NREM to NREM transitions (191 with EEG), 174 events for NREM to short MA transitions (133 with EEG), 75 events for NREM to long MAs (54 with EEG), and 66 events for NREM to wake transitions (42 with EEG) across the seven mice.

For the LC activation-related analysis, 108 events were found for Threshold −15 (11 of these being YFP), 260 events for Threshold −10 (55 of these YFP), 777 events for Threshold −5 (296 of these YFP), 1,444 events for Threshold 0 (510 of these YFP), and 1,148 events for Threshold 5 (377 of these YFP) across ten animals (four being YFP). For the LC suppression experiment, 48 events were found for the four Arch mice and 38 for the four YFP mice. All data preprocessing was performed using MATLAB. For the mouse data related to HRB events, 1654 events were found across the 7 mice also analyzed in connection with sleep transitions.

For all four experiments, the event variables were generated, and data from RR, NE, and the EEG frequency bands were extracted from 60 s before and after each event. Each data type was averaged across all events from all mice within each relevant group for each experiment, leading to the standard error being based on the number of events rather than the number of animals.

Across all experiments and data types, the AUC was calculated as a measure of change between the different conditions. For the sleep transitions experiments, the AUC from −25 to 0 s (zero being the event time) was used as a baseline and was subtracted from the AUC from 0 to 25 s after the event. For the ChR2/YFP experiment, three types of AUC measures were used: the pre-stimulation AUC was defined as the data from -40 to −20 s before the event (baseline) subtracted from the data from −20 to 0 s before the event. Similarly, the post-stimulation AUC was defined as the data from 20 to 40 s before the event (baseline) subtracted from the data from 0 to 20 s after the event. The pre-stimulation minus the post-stimulation AUC was defined as the difference between the two measures outlined above. Lastly, the AUC for the Arch/YFP-related experiments were defined as -40 to −5 s before the event marker (baseline) subtracted from 5 to 40 s after the event marker.

The data presented in Fig. 3 was based on the same +/- 60 s data files surrounding the LC stimulation onset as seen in Fig. 2, though only ChR2 animals were included in this analysis, as it focused on HR changes in response to stimulation. In Fig. 3a, mean traces across thresholds were shown and 97 raw traces were randomly extracted for each threshold to visualize variability, as this was the lowest number of total events present within a threshold. In Fig 3b, the RR, BPM and RR SD estimate was based on the full 2 min data file, while the sample entropy (m = 2) only included the 10 s after laser onset (65). The following analysis was based on the HR deceleration event marker. Here, the RR amplitude was calculated as the RR at HR decelerations minus the mean RR 8-10 s prior. Similarly, the NE amplitude was determined by subtracting the lowest NE value 5-10 s before the HR decelerations from the NE value at the HR decelerations. The mean NE values also reflected the NE values in the 10 s leading up to the HR decelerations. Lastly, the sigma amplitude was determined as the mean sigma 2-3 s before the HR decelerations subtracted from the sigma value at the HR decelerations.

For the analysis related to HRB events, amplitudes were analyzed rather than AUC to use similar measures as in the human data. For the mouse data, the RR amplitude was defined as the RR value at the HRB minus the mean across the 15 to 10 s leading up to the event. The NE amplitude was calculated as the peak value in the 2.5 s interval around the HRB event minus the minimum in the 15 to 10 s interval preceding the event. Lastly, the sigma amplitude was defined as the smallest observation in the 2.5 s interval around the HRB event minus the largest value found between 15 to 5 s range preceding the event.

The frequency-domain analysis was based on the non-parametric Welch approach, and similar procedures were used on NREM data across all experiments. The data underwent detrending with a 5th degree polynomial fitted to the raw data before analysis as in Ferreira et al. (66). Across experiments, frequencies between 0-0.1 Hz with a frequency resolution of 0.0002 Hz were analyzed for NE, and frequencies between 0-0.4 Hz with a resolution of 0.002 Hz were calculated for RR. Across all frequency-domain analyses, consecutive periods of minimally 300 s of NREM sleep (with MAs) were included for RR/NE, as this was the minimal NREM period duration suggested to capture VLF (67). However, as the ability to interpret VLF power within short term recordings has previously been questioned (68), we also extracted NREM sleep periods of minimum 11 minutes (allowing 10 cycles of VLF 0-0.015 Hz) to visualize an overlap in the frequency distribution (Supplementary Fig. 1f). Given the large similarities between the frequency distributions, we chose to utilize the minimum of 300 s NREM as the data selection criteria, as it is uncommon for mice to remain in NREM sleep for over 10 min, making frequency-domain analyses across different experimental paradigms challenging. Each bout of data was run through Welch separately, and its AUC, peak power value, and frequency at peak power were extracted along with information about the length of the period (bout length). Bout length was subsequently used to weigh the final mean/SEM estimates, which was utilized both for statistical analysis and visualization generation. SEM was based on the number of bouts. The PSD presented in Fig. 1 and Supplementary Fig. 1 were based on the same data from the mice used to analyze sleep transitions and held 144 bouts. For the LC activation data, the PSD was made separately for ChR2/YFP and for each stimulation threshold including the baseline before and after stimulation. The number of bouts for the ChR2 data (from baseline to threshold 5 were 212, 44, 41, 30, 37, and 43. For YFP, these numbers were 147, 24, 28, 21, 27, and 28. For the LC suppression data, the PSD was split by Arch/YFP, which consisted of 66 and 43 bouts respectively.

Statistical analysis

Linear mixed-effects models (LMMs) were utilized across all figures to estimate the effects using R. One exception was made for models related to memory estimates, as only one estimate was available per night per subject. Thus, a standard linear model was employed instead. The choice of applying LMMs to the data was made, as the model type allowed for event-based analysis by adding a random intercept for subject ID. By doing so, a hierarchy was created in the data to ensure that issues with independence did not occur (69). When continuous predictors were present in the model, the predictor was added as a random slope and compared to the base model (with only a random intercept) to determine which model was best using the Akaike and Bayesian information criteria. If the random slope model had a lower Akaike/Bayesian information criterion, converged, and did not have a singular fit, the statistics were summarized based on that model.

Models correlating sleep transitions with an array of dependent variables were built by setting the four sleep transitions as independent variable factors, which were re-leveled to ensure that the lowest level of arousal (NREM to NREM) was set as the baseline. For models associated with Fig. 2-3, the thresholds were similarly set as a factor, with −15 as the baseline threshold, except for the frequency-domain analysis, where the baseline was set as the intercept. The thresholds and conditions (ChR2 or YFP) were both set as predictors in Fig. 2 and had an interaction effect to explore the differences between the two conditions at different thresholds. In Fig. 3, the thresholds were the main predictor utilized, as only ChR2 data was included. Lastly, models for Fig. 4 had a similar predictor setup with condition (YFP or Arch) as the predictor. All LMMs had SubjectID as a random intercept.

After fitting the LMMs, they were assessed for any violations of the fixed effects assumptions through visual inspection of the residual distributions. While LMMs have been found to be robust to assumption violations (70), bootstrapping of 1000 samples was performed to obtain robust estimates of the model in cases where one or more assumptions had been violated. For models where autocorrelation of residuals was present, block bootstrapping was chosen over normal bootstrapping. The block bootstrapping resampled the data in groups based on subject ID, which was the recommended solution to issues with autocorrelations of residuals. Additionally, when relevant to the analysis, weights were added to the LMMs and factored into the bootstrapping to ensure the bout length was appropriately weighed in the model.

For models with a continuous independent variable, the marginal and conditional R squared values were reported in the figure and was also the measure that was bootstrapped (71, 72). In cases where the predictor had more than two levels or there were multiple predictors, the estimates of the pairwise comparisons were bootstrapped. Lastly, in cases where only a single predictor with two levels was present, the difference between model estimates for the two levels was bootstrapped. All pairwise comparisons were adjusted for multiple comparisons using Turkey’s method.

Participants

28 human participants (Mage=20.46 ± 2.73 years, 17 females) were included in the analysis of HRB. For information regarding demographics, see the supplementary table. Participants had no history of psychological, neurological or chronic illnesses and were previously included in another publication (73). All participants provided informed consent approved by both the Western Institutional Review Board and the Human Research Review Board at the University of California, Riverside. Individuals were excluded from participation if they exhibited irregular sleep/wake cycles, had diagnosed sleep disorders, personal or familial histories of psychopathology, substance abuse or dependence, a history of epilepsy or loss of consciousness exceeding two minutes, current psychotropic medication use, or cardiac or respiratory illnesses potentially influencing cerebral metabolism. Eligibility was determined during an in-person psychiatric assessment conducted by trained research personnel. Additionally, participants underwent a medical evaluation, including a detailed medical history review and physical examination by a staff physician, to confirm overall physical health. Participants were instructed to avoid caffeine, alcohol, and all stimulants for at least 24 hours prior to and during the experimental day. Compliance with consistent sleep patterns for one week prior to each experimental session was verified using sleep diaries completed daily and wrist-based activity monitors (Actiwatch Spectrum, Philips Respironics). Participants were required to achieve a minimum of 7 hours of sleep per night during this period, including the night immediately before the study day. Monetary compensation and/or course credits were provided to participants in acknowledgment of their time and participation.

Word-paired association task

Human participants underwent the word-paired association task, which consisted of an encoding, criterion and testing phase. In the encoding phase, participants were presented with 60 word-pairs one at a time for 1 s with an inter-trial interval of 1s. During the criterion phase, the participants were probed with one of the words from a pair and were subsequently presented with the missing word for 2 s followed by a 100 ms inter-trial interval. This process continued until the participants obtained 70% accuracy overall. Both the encoding and criterion phase occurred in the morning on the first day. Lastly, the testing phase was conducted three times: shortly after the criterion phase, in the evening before participants slept in the lab and the following morning. At each test, participants were tested on their memory for 20 of the half word-pairs. Participants were shown one word and asked to recall the missing word. No feedback was provided during the testing phase and there was a 1 s inter-trial interval between pairs. Performance was measured as the difference score between the first test and the final test approximately 24 hours later.

Heart rate detection in humans

In humans, ECG data were recorded using a modified Lead II Einthoven configuration at a sampling rate of 1,000 Hz. Heart rate variability (HRV) analysis was conducted on the R-wave intervals across the entire sleep-wake period using Kubios HRV Analysis Software 2.2 (Biosignal Analysis and Medical Imaging Group, University of Kuopio), following Task Force guidelines(74). The R-wave peaks within the ECG QRS complex were automatically detected by Kubios software and subsequently verified visually by trained technicians. Any incorrectly identified R-peaks were manually adjusted, and missing beats were interpolated using cubic spline interpolation. After computing inter-beat intervals, a third-order polynomial detrending filter was applied to eliminate trend-related components in the data. Artifacts were further corrected using Kubios software’s built-in automatic medium-level artifact correction filter.

Human sleep recordings

EEG data were collected using a 32-channel electrode cap (EASEYCAP GmbH) equipped with Ag/AgCl electrodes arranged according to the international 10-20 system (75). Among these electrodes, 22 were utilized for scalp EEG recordings, while the remaining electrodes were dedicated to ECG, electromyogram (EMG), electrooculogram (EOG), ground, an online common reference channel located at FCz (retained post-referencing), and mastoid references (A1 and A2). EEG signals were recorded at a sampling rate of 1,000 Hz and subsequently referenced offline to the contralateral mastoid electrode (A1 and A2).

Data preprocessing was performed using BrainVision Analyzer 2.0 software (BrainProducts). Eight scalp electrodes (F3, F4, C3, C4, P3, P4, O1, O2), along with EMG and EOG recordings, were selected for scoring nighttime sleep stages. EEG and EOG signals were filtered with high-pass and low-pass settings of 0.3 Hz and 35 Hz, respectively. Raw data were visually staged in 30-second epochs into Wake, Stage 1, Stage 2, Slow Wave Sleep (SWS), and REM sleep, according to the criteria of Rechtschaffen and Kales, using the custom MATLAB-based scoring toolbox, HUME. Following staging, epochs containing artifacts and arousals were visually identified and removed prior to spectral analysis.

Human data analysis

HRBs were used as the event marker for human data and was defined as when the RR intervals drop below 2 SD of the mean in a 3 min window (14, 15). This resulted in 4753 events across the 28 subjects during stage 2 NREM sleep. For these events, the RRRecovery amplitude reflected the mean RR values 5 to 10 s after the HRB minus the RR value at the event-time. The sigma power (12-15 Hz) trace was extracted and its value in the 5 to 0 s preceding the event was used to correlate to the RRRecovery. To correlate the sigma values to memory estimates, we followed the same procedure as past papers(14). Sigma power was extracted in +/- 20 s around the HRB and then segmented into four-time bins: 10 to 5 s preceding the event, 5 to 0 s preceding the event, 0 to 5 s following the event, and 5 to 10 s following the event. For each participant, sigma power was averaged across all events within each bin, resulting in one mean value per bin. To correlate sigma power before the HRB and memory performance, baseline sigma power was subtracted from the sigma power 5 s before the HRB, following the equation below. This baseline represents mean sigma power during NREM sleep outside of 20 s windows surrounding event extracted data. All statistical tests followed the same principles outlines in the ‘Statistical Analysis’ section.

eLife Assessment

This valuable study provides evidence that locus coeruleus activity is coordinated with heart rate during sleep, confirming previous work in mice and humans, with a possible role for sleep-dependent memory consolidation. The claims are supported by solid evidence, although the underlying mechanisms and the predictive value of the correlative dataset would benefit from additional controls. This work will be of interest to neuroscientists focusing on sleep, memory, and autonomic functions.

Reviewer #1 (Public review):

Summary:

This study examined whether infraslow fluctuations in noradrenaline and in heart rate are coupled and how they are affected by sleep transitions. The authors used the fluorescent NA biosensor GRAB-NE2m in the medial prefrontal cortex of mice to record extracellular NA while also recording EEG and EMG during sleep-wake episodes. They also analyzed previously published human data to reproduce relationships they found between sigma power and RR intervals in mice.

Strengths:

This is an impressive study with significant strengths, as it involves a rich set of data that includes not only observations of associations between heart rate and noradrenergic dynamics but also optogenetic manipulation of the locus coeruleus. Human data is presented to show parallels in the association between sigma power during sleep and phasic heart-rate bursts.

Weaknesses:

(1) Language could be clearer and more precise. As detailed below, in both the introduction and the discussion, the way the hypotheses and study objectives are described could use some revision to be more precise and accurate.

1A) In the introduction on p. 4: The overarching question is framed as "could the peripheral autonomous systems be a read-out of the central LC-NE system and thus be a biomarker of memory consolidation and LC dysfunction?" This gives the impression that the LC function would be the main influence on peripheral autonomous systems. There are, of course, many influences on peripheral autonomous systems, so it would be advisable for the authors to be more specific here about what signal(s) in particular would be predicted to be sensitive markers of LC function.

1B) In the discussion on p. 12: "In this study, we leveraged real-time measurements of mPFC NE levels and HR measurements from EMG recordings in mice to investigate the causal link between the two variables with high temporal resolution in freely moving sleeping mice, with similar inspection in humans." To test the causal link between mPFC NA levels and HR measures, the study would manipulate NA levels just in the mPFC and not elsewhere in the brain. However, in this study, the manipulation occurred in the LC, and so there would be broad cortical changes in NA levels. Thus, it could be that LC activity causes HR changes via a non-PFC pathway.

(2) Comparisons with the control condition need further development.

2A) While the authors did include a key YFP control condition, in the main text no direct statistical comparison between the closed-loop optogenetic stimulation (ChR2) condition and the YFP control condition was reported. (It was reported in Supplementary Figure 2c-d.) Instead, in the main text, the authors only reported that the effects of stimulation were significant in the closed-loop condition and not in the control. However, that is not the same as demonstrating that the two conditions significantly differed from each other, and it is the direct test that is important for the conclusions, so it seems important to include this result in the main presentation.

2B) In addition, the authors should address the issue that the pre-stimulation NE was consistently significantly lower in the YFP condition than in the ChR2 condition (see Supplementary Figure 2c), which is a potential confound.

2C) Direct comparison of the strengths of correlations shown in Figure 2h vs. Supplementary Figure 2f should be included. Currently, we see relatively weak correlations in both ChR2 and YFP conditions, and it is not clear if the relationships differ in the control. It seems they are still present in the control condition but weaker, which would contradict the apparently broad claim on p. 7 that "No such effects were present in the control condition" (it is not entirely clear whether this claim refers to all effects discussed in the figure or just a subset - this language should be clarified).

2D) Did the YFP controls vs. ChR2 animals show any differences in the number of NA states that triggered stimulation in the closed-loop system? With ChR2 animals, stimulation changes NA, which could change future triggering. In YFP animals, nothing changes NA (other than natural fluctuations), so the dynamics of stimulation timing could diverge between groups in a way that complicates interpretation. Specifically, if ChR2 stimulation raises NA and prevents future threshold crossings, ChR2 animals may end up receiving fewer subsequent stimulations than YFP animals (or a different temporal clustering). If the number or pattern of stimulation differed in two groups, it would be important to have a yoked control where matched animals get the same stimulation pattern but not triggered by their own NA.

(3) Some more discussion/explanation of the rationale for the closed-loop approach and how it influences how we should interpret the results could be useful. For instance, currently, it is not clear whether LC stimulation needs to be timed after an NA dip to yield the effects seen.

(4) The section on heart rate decelerations is hard to follow. In particular, I was not sure how to interpret Figure 3f-j. For Figure 3f, what does the middle line represent? The laser onset or the max RR value after laser onset? What is the baseline that is used to correct the values to obtain amplitudes? If it is the whole period before the maximal RR value or the laser onset, wouldn't baseline values differ significantly across conditions and so potentially account for differences seen between conditions in the reported HR decelerations? Larger HR decelerations may be seen in conditions with higher HR simply as a regression to the mean phenomenon.

(5) The findings regarding LC suppression could be further clarified.

5A) Page 8: "observed a response in NE decline" - please be more precise. Did NE decline more or less?

5B) It would be helpful to also show the correlation between NE and RR in the control (YFP) condition and whether there were any differences between YFP and Arch conditions (Figure 4e).

5C) This sentence took me multiple readings to understand - it would be helpful to rewrite to make it clearer: "indicating that, while HR generally did not respond strongly to LC suppression, the variability in RR responses was dependent on NE changes to the suppression (Figure 4e)."

5D) The two colors in Figure 4 are similar and hard to distinguish.

5E) The correlations shown in Figure 4j seem to be driven by just two of the cases. Are the effects significant when outliers are removed?

5D) Page 10: Were there any differences in memory performance between the Arch and YFP conditions?

5E) Page 10: "We found a correlation between RR responses to LC suppression and sigma power, suggesting that a stronger HR reduction response is linked to higher spindle power." It should be noted in the text that the correlation was not specific to sigma (it was also seen for theta and beta, Figure 4i).

(6) It is not clear which of the sigma power and RR interval findings do/do not exactly line up between the mice and humans. It could be helpful to have a table comparing them. For instance, was the finding in humans that pre-HRB sigma power was positively associated with slowing in heart rate after the HRB also seen in mice? Was there evidence in mice (as seen in the human sample) that sleep-dependent memory improvement was associated with pre-HRB sigma power?

(7) Page 18: It is not clear if the sex of mice was balanced across controls and optogenetics groups.

Reviewer #2 (Public review):

Summary:

The major part of this study reproduces previously published findings in both mice and humans and provides incremental analyses on these findings. In essence, the work reaffirms the presence of coordinated infraslow fluctuations in sigma power and heart rate during NREM sleep. It further confirms previous findings that coordination depends on noradrenaline-releasing neurons in the locus coeruleus. Also supporting previously published work in mice and humans, the authors describe a link between the strength of these infraslow fluctuations and memory consolidation in mice and humans.

Strengths:

The authors successfully replicate key previously reported phenomena across both mice and humans. Confirmatory studies and demonstrations of reproducibility are essential for progress in neuroscience. To maximize their value, such studies should clearly acknowledge their confirmatory nature and carefully situate what, in their view, are novel results, going beyond existing literature.

Weaknesses:

The authors' interpretation of their data needs to be revised. Many of their claims regarding the mechanistic basis of their findings and the predictive value of their correlative datasets are not supported by the available evidence.

In the present manuscript, several citations of literature on the work they reproduce lack precision or completeness, which reduces transparency and obscures how the reported findings relate to previously established results.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This study examined whether infraslow fluctuations in noradrenaline and in heart rate are coupled and how they are affected by sleep transitions. The authors used the fluorescent NA biosensor GRAB-NE2m in the medial prefrontal cortex of mice to record extracellular NA while also recording EEG and EMG during sleep-wake episodes. They also analyzed previously published human data to reproduce relationships they found between sigma power and RR intervals in mice.

Strengths:

This is an impressive study with significant strengths, as it involves a rich set of data that includes not only observations of associations between heart rate and noradrenergic dynamics but also optogenetic manipulation of the locus coeruleus. Human data is presented to show parallels in the association between sigma power during sleep and phasic heart-rate bursts.

We thank Reviewer #1 for their thoughtful, detailed, and constructive evaluation of our manuscript. We appreciate their recognition of the strengths of the study, particularly the integration of noradrenergic recordings, optogenetic manipulation, and cross-species analyses. We are especially grateful for the reviewer’s careful attention to clarity, experimental interpretation, and control comparisons. The comments have helped us sharpen the framing of our hypotheses, clarify causal claims, improve statistical reporting, and better explain our closed-loop approach and heart rate analyses. We have addressed each point in detail below and believe that the revisions substantially strengthen the manuscript.

Weaknesses:

(1) Language could be clearer and more precise. As detailed below, in both the introduction and the discussion, the way the hypotheses and study objectives are described could use some revision to be more precise and accurate.

Thank you for this helpful comment. We have sharpened the description of the study objectives, hypotheses, and interpretation of the findings to better distinguish between what was directly tested, what was inferred, and what remains speculative. We revised the language throughout these sections to improve clarity, accuracy, and overall readability

(1A) In the introduction on p. 4: The overarching question is framed as "could the peripheral autonomous systems be a read-out of the central LC-NE system and thus be a biomarker of memory consolidation and LC dysfunction?" This gives the impression that the LC function would be the main influence on peripheral autonomous systems. There are, of course, many influences on peripheral autonomous systems, so it would be advisable for the authors to be more specific here about what signal(s) in particular would be predicted to be sensitive markers of LC function.

Thank you for this important point. We agree that heart rate reflects the integrated output of multiple autonomic mechanisms and should not be interpreted as being exclusively driven by LC activity. Cardiac dynamics arise from the balance between sympathetic and parasympathetic influences, which themselves are regulated by several central and peripheral systems. In addition, recent work shows that the infraslow oscillations observed during NREM sleep are not restricted to norepinephrine alone but also involve other neuromodulatory systems, including acetylcholine and serotonin (e.g., Teng et al., PNAS 2025, Kjaerby et al., iScience, 2026). Our intention was therefore not to imply that the LC is the sole driver of peripheral autonomic dynamics. We have revised our overarching questions to make them more specific: Please see new text below:

(Introduction, page 4/5). “The sympathetic and parasympathetic autonomic nervous system are involved in HRV, which is conventionally analyzed across three primary frequency bands: high frequency (HF) HRV, low frequency (LF) HRV, and very low frequency (VLF) HRV (Berntson et al., 1997). Due to the frequency overlap with VLF HRV, we wondered if central infraslow NE dynamics could be linked to this poorly understood HRV indicator. Furthermore, are infraslow NE fluctuations directly reflected by HRV under different physiological states or does LC–HR coupling scale differently with LC output? Specifically, if infraslow NE oscillations display faster frequencies - as occurs during sleep fragmentation - will cardiac dynamics exhibit corresponding changes? Conversely, given that stronger infraslow NE dynamics correlate with memory consolidation through their regulation of sleep spindles, could the peripheral autonomic signatures provide an accessible cross-species biomarker of spindle-dependent memory consolidation? Addressing these questions could help bridge mechanistic insights into LC-mediated sleep regulation with established HRV metrics used in human physiology.”

(1B) In the discussion on p. 12: "In this study, we leveraged real-time measurements of mPFC NE levels and HR measurements from EMG recordings in mice to investigate the causal link between the two variables with high temporal resolution in freely moving sleeping mice, with similar inspection in humans." To test the causal link between mPFC NA levels and HR measures, the study would manipulate NA levels just in the mPFC and not elsewhere in the brain. However, in this study, the manipulation occurred in the LC, and so there would be broad cortical changes in NA levels. Thus, it could be that LC activity causes HR changes via a non-PFC pathway.

We thank the reviewer for this important comment. Indeed, mPFC NE is merely a readout of LC activations and we expect that NE in other brain regions would show the same patterns. Indeed, mPFC NE is not expected to provide any causal link to heart rate. We have revised added a sentence to the results section and also changed the initial summary part of the discussion to reflect this better.

(Results, page 6). “mPFC was selected as a representative cortical readout of LC-mediated norepinephrine dynamics, as infraslow NE fluctuations are coordinated across widespread brain regions.”

(Discussion, page 14/15). “Variability in HR is a non-invasive biomarker of autonomic nervous system function and is frequently disrupted in ageing and Alzheimer’s disease. Here, by combining real-time measurements of mPFC NE dynamics with simultaneous HR recordings in freely sleeping mice, we demonstrate that HR closely tracks the infraslow phasic activity of the LC–NE system.”

(2) Comparisons with the control condition need further development.

(2A) While the authors did include a key YFP control condition, in the main text no direct statistical comparison between the closed-loop optogenetic stimulation (ChR2) condition and the YFP control condition was reported. (It was reported in Supplementary Figure 2c-d.) Instead, in the main text, the authors only reported that the effects of stimulation were significant in the closed-loop condition and not in the control. However, that is not the same as demonstrating that the two conditions significantly differed from each other, and it is the direct test that is important for the conclusions, so it seems important to include this result in the main presentation.

We thank the reviewer for this important point and agree that direct statistical comparisons between ChR2 and YFP conditions are important for interpretation. These comparisons were performed and are shown in Supplementary Figure 2c–d, but we acknowledge that this was not sufficiently emphasized in the main text. We are now more clearly referring to this comparison in Result section:

(Results, page 9). “The magnitude of pre-stimulation NE descent and post-stimulation NE ascent was reduced as the thresholds increased, indicating less pronounced NE dynamics as LC stimulation became more frequent (Fig. 2e, for direct comparison with YFP control, see Suppl. Fig. 2c-d).”

Our rationale for prioritizing the within-animal threshold comparisons in the main figure was that the central experimental question concerned how progressive shifts in infraslow NE oscillatory frequency influence the NE–HR relationship. Because variability in viral expression levels (both NE sensor expression and LC opsin expression) introduces substantial between-animal variability, we considered within-animal comparisons across threshold conditions to provide the most informative representation of how changes in LC-driven NE dynamics alter cardiac responses. That said, we agree that highlighting the direct ChR2 versus YFP comparison is important for the overall interpretation. As mentioned, the reference to Suppl. Fig. 2c– d, where the between-group analyses are visualized are now clearly referred to.

(2B) In addition, the authors should address the issue that the pre-stimulation NE was consistently significantly lower in the YFP condition than in the ChR2 condition (see Supplementary Figure 2c), which is a potential confound.

We thank the reviewer for bringing up this important point. We agree that differences in stimulation timing between ChR2 and YFP animals could complicate interpretation in a closed-loop design and appreciate the opportunity to clarify this aspect of the experiment.

In the ChR2 condition, animals were exposed to repeated optogenetic LC activation designed to mimic progressively faster infraslow NE dynamics. Such repeated stimulation is expected to produce a gradual elevation in tonic NE levels across the recording session, which explains the higher pre-stimulation baseline relative to YFP controls. We acknowledge that elevated tonic NE levels could introduce additional physiological effects. For example, higher NE tone would be expected to increase α2mediated autoinhibitory feedback on LC neurons and presynaptic NE release. Within the LC itself, we expect that optogenetic stimulation would largely override such effects due to the strong Na+-mediated depolarization induced by ChR2 activation. However, NE release in downstream regions such as the mPFC may be influenced to some extent by elevated tonic noradrenergic tone. Importantly, such feedback mechanisms would likely also occur under physiological conditions characterized by elevated LC activity, such as stress or sleep fragmentation. Because the goal of our stimulation paradigm was to model progressively faster infraslow NE dynamics under physiologically relevant conditions, we believe this feature of the manipulation may in fact increase the translational relevance of the model. We specifically address this elevation in Figure 3, where we show that very rapid stimulation regimes are accompanied by signs of compensatory cardiovascular regulation, likely reflecting baroreceptor-mediated responses to sustained increases in heart rate.

We agree that the precise contribution of elevated tonic NE to the overall manipulation cannot be fully disentangled in the present study. We therefore avoid overinterpreting these effects and have instead added text to the manuscript acknowledging this consideration without extensive speculation.

(Results, page 9) “Since pre-stimulation NE baseline levels progressively became higher in the ChR2 condition compared with YFP controls (Fig. 2c+e, Supplementary Fig. 3a-f), it demonstrates that they arise from stimulation-dependent modulation of noradrenergic tone rather than nonspecific signal drift. As a result, the pre-stimulation state at higher thresholds differed between ChR2 and control conditions, which should be considered when interpreting the immediate effects of laser stimulation across thresholds.”

(2C) Direct comparison of the strengths of correlations shown in Figure 2h vs. Supplementary Figure 2f should be included. Currently, we see relatively weak correlations in both ChR2 and YFP conditions, and it is not clear if the relationships differ in the control. It seems they are still present in the control condition but weaker which would contradict the apparently broad claim on p. 7 that "No such effects were present in the control condition" (it is not entirely clear whether this claim refers to all effects discussed in the figure or just a subset - this language should be clarified).

We thank the reviewer for this important comment and agree that the original wording could be interpreted as implying a complete absence of an NE–RR relationship in the YFP condition. To address this concern, we directly compared the strength of the NE– RR relationship between ChR2 and YFP animals. Please see Supplementary Figure 3h.

Using a linear mixed-effects model that accounted for repeated measurements within animals, we found that the slope of the NE–RR relationship was significantly steeper in ChR2 animals than in YFP controls (all thresholds: slope difference = 1.915, p < 0.0001; thresholds −15, −10, and −5 only: slope difference = 2.367, p < 0.0001). Consistent with this result, comparison of Pearson correlations using Fisher's r-to-z transformation also indicated significantly stronger coupling in ChR2 animals than in YFP controls (see Statistics in the Supplementary File).

These analyses demonstrate that an inverse relationship between NE and RR is present under physiological conditions in YFP animals, but that optogenetic LC activation substantially strengthens this coupling. We have revised the corresponding text to clarify this:

(Results, Page 9): “An inverse NE–RR relationship was present in both ChR2 (Fig. 2h) and YFP (Suppl. Fig. 3f) groups but was significantly stronger in ChR2 animals than in YFP controls (Suppl. Fig. 3h).”

(2D) Did the YFP controls vs. ChR2 animals show any differences in the number of NA states that triggered stimulation in the closed-loop system? With ChR2 animals, stimulation changes NA, which could change future triggering. In YFP animals, nothing changes NA (other than natural fluctuations), so the dynamics of stimulation timing could diverge between groups in a way that complicates interpretation. Specifically, if ChR2 stimulation raises NA and prevents future threshold crossings, ChR2 animals may end up receiving fewer subsequent stimulations than YFP animals (or a different temporal clustering). If the number or pattern of stimulation differed in two groups, it would be important to have a yoked control where matched animals get the same stimulation pattern but not triggered by their own NA.

We thank the reviewer for this important point. We agree that differences in stimulation timing between ChR2 and YFP animals could complicate interpretation in a closed-loop design.

Importantly, stimulation triggering was based on relative declines in NE fluorescence calculated against a rolling 2-minute baseline, rather than absolute NE levels. Thus, as tonic NE levels gradually increased in ChR2 animals, the threshold adapted accordingly, reducing the likelihood that elevated baseline NE alone would prevent future triggering. Instead, stimulation continued to occur when NE declined relative to the recent baseline, thereby preserving the infraslow closed-loop structure.

The number of stimulation events across thresholds is already reported in the manuscript (Methods, p. 30 and corresponding figure legends), but we have now indicated more clearly in the result section where to find the information:

(Results, Page 9). “Mean traces of NE and RR were aligned to LC stimulation onset (Fig. 2c-d, for number of laser stimulations see Fig. 2 legend or Methods).”

(Methods, Page 31). “For the LC activation-related analysis, 108 events were found for Threshold -15 (11 of these being YFP), 260 events for Threshold -10 (55 of these YFP), 777 events for Threshold -5 (296 of these YFP), 1,444 events for Threshold 0 (510 of these YFP), and 1,148 events for Threshold 5 (377 of these YFP) across ten animals (four being YFP).”

While the total number of events was lower in YFP animals, this is expected in part due to the smaller group size (4 YFP vs. 6 ChR2 animals included in this analysis). Furthermore, because ChR2 stimulation increased NE levels by design, more pronounced subsequent declines in NE may have modestly facilitated additional threshold crossings.

Nevertheless, the overall temporal structure of the stimulation paradigm remained comparable across groups, and YFP animals underwent the same closed-loop stimulation protocol. Our primary comparison was mainly based on shifts in within-animal NE oscillatory frequency and how this change would impact the connection to HR. Thus, we believe the present control condition appropriately addresses the central question of whether optogenetic LC activation are able to conduct a continuum of NE oscillations.

(3) Some more discussion/explanation of the rationale for the closed-loop approach and how it influences how we should interpret the results could be useful. For instance, currently, it is not clear whether LC stimulation needs to be timed after an NA dip to yield the effects seen.

We thank the reviewer for pointing this out. The rationale for the closed-loop LC stimulation approach was to test whether the relationship between infraslow LC–NE dynamics and heart rate is maintained only under physiological infraslow conditions or whether it breaks down when the rhythm becomes progressively faster, as occurs during sleep fragmentation and other high-arousal states. Specifically, we asked whether heart rate continues to track LC–NE fluctuations as the infraslow rhythm shifts toward higher frequencies, thereby assessing its utility as a potential biomarker of disrupted restorative sleep.

To address this while preserving the intrinsic temporal structure of infraslow LC activity, we implemented a closed-loop strategy in which stimulations were triggered following defined declines in the NE signal, using a rolling preceding 2-minute window as baseline. This allowed LC activation to occur during the descending phase of the endogenous infraslow cycle, maintaining its physiological phase structure while systematically increasing its effective frequency. By progressively relaxing the decline threshold, stimulations were triggered earlier in the cycle, thereby compressing the infraslow period in a controlled manner.

Importantly, the intention was not to test whether LC stimulation specifically needs to occur after an NE dip to elicit the observed effects. Rather, triggering stimulation during the decay phase provided a way to accelerate the infraslow rhythm without disrupting sleep through indiscriminate stimulation. This enabled us to examine whether the coupling between LC–NE dynamics and heart rate remains stable under increasingly rapid infraslow regimes. Our results indicate that this relationship weakens at higher infraslow stimulation frequencies, suggesting that heart rate reliably reflects physiological LC–NE oscillations but becomes less tightly coupled when the rhythm is compressed beyond its normal range.

We have clarified this rationale in the revised manuscript by adding the below section in the result section.

(Result, page 8/9). “This approach enabled controlled compression of the infraslow NE cycle by triggering LC activation during the descending phase of the endogenous NE signal, thereby increasing the effective oscillatory frequency while preserving the temporal structure of physiological LC–NE dynamics. This strategy allowed us to test whether heart-rate responses continue to track LC-driven NE fluctuations as the infraslow rhythm becomes progressively faster.”

(4) The section on heart rate decelerations is hard to follow. In particular, I was not sure how to interpret Figure 3f-j. For Figure 3f, what does the middle line represent? The laser onset or the max RR value after laser onset? What is the baseline that is used to correct the values to obtain amplitudes? If it is the whole period before the maximal RR value or the laser onset, wouldn't baseline values differ significantly across conditions and so potentially account for differences seen between conditions in the reported HR decelerations? Larger HR decelerations may be seen in conditions with higher HR simply as a regression to the mean phenomenon.

We thank the reviewer for this feedback and agree that additional clarification of Figure 3f–j is warranted.

For Figure 3f, the central line represents the peak RR value (maximal heart-rate deceleration) identified within the 2–7 s window following laser onset, rather than the laser onset itself. We realize this was not sufficiently clear and have revised the figure and corresponding Results text to clarify this point.

Regarding baseline correction, RR amplitudes were calculated as the difference between the RR at peak deceleration and the mean RR during the 8–10 s period preceding the RR peak, as described in the manuscript. Thus, the baseline was defined locally for each event and was not based on the entire pre-laser period or stimulation onset. We chose this approach to account for shifts in baseline heart rate across conditions and to capture the relative magnitude of the deceleration response rather than absolute RR values.

We appreciate the reviewer’s point regarding potential regression-to-the-mean effects, particularly in conditions with higher baseline heart rates. This is an important consideration. However, because the amplitude measure was baseline-corrected on an event-by-event basis, we believe the reported differences are unlikely to be explained solely by higher pre-stimulation heart rate. At the same time, our findings clearly show that elevated baseline heart rate influence the dynamic range of deceleration responses. Our findings show that under physiological conditions with elevated HR, larger heart-rate fluctuations would also contribute to increased HRV, which is often interpreted positively, despite potentially reflecting fragmented or dysregulated sleep states in this context.

To improve readability, we have revised the figure and associated text.

(Results, page 10). “To quantify the HR decelerations that happened after LC activation, we took the maximal RR value (so slowest HR) 2-7 s after LC stimulation and baseline corrected the value to the mean RR during the 8–10 s period preceding the RR peak to obtain their amplitude.”

(5) The findings regarding LC suppression could be further clarified.

(5A) Page 8: "observed a response in NE decline" - please be more precise. Did NE decline more or less?

We thank the reviewer for this suggestion and agree that the original wording was imprecise. To clarify the direction and nature of the response, we have revised the text to state:

(Results, page 11). “...we observed a gradual NE decline sustained throughout the laser period that was not observed in the YFP condition…”

(5B) It would be helpful to also show the correlation between NE and RR in the control (YFP) condition and whether there were any differences between YFP and Arch conditions (Figure 4e).

We thank the reviewer for this suggestion. We have now added the corresponding YFP correlation to Figure 4e. In the YFP group, the relationship between NE and RR showed a similar negative trend but did not reach statistical significance (p = 0.053). To directly assess whether the NE–RR relationship differed between Arch and YFP animals, we performed both a linear mixed-effects analysis and a Fisher r-to-z comparison.

Neither analysis revealed a significant difference between groups. The linear mixed-effects model showed no significant Group × NE interaction (slope difference = 0.994, p = 0.51), indicating that the NE–RR coupling was not altered by LC suppression. Similarly, Fisher's r-to-z comparison found no significant difference between the correlations (p = 0.42).

We believe this result is consistent with the relatively modest nature of the LC suppression paradigm. While Arch stimulation produced a clear reduction in NE levels, it did not induce a large shift in the overall NE–RR relationship. Instead, the data suggest that heart-rate responses remain coupled to noradrenergic fluctuations under both physiological conditions and during mild LC suppression. We have added the YFP data and clarified this interpretation in the revised manuscript.

(Results, page 12): “A similar negative relationship was observed in YFP controls (Fig. 4e), and the strength of the NE–RR association did not differ significantly between Arch and YFP animals (Suppl. File, Statistics), suggesting that LC suppression did not substantially alter the underlying coupling between these measures.”

(5C) This sentence took me multiple readings to understand - it would be helpful to rewrite to make it clearer: "indicating that, while HR generally did not respond strongly to LC suppression, the variability in RR responses was dependent on NE changes to the suppression (Figure 4e)."

We agree with the reviewer that this phrasing is hard to understand and we have optimized for better clarity. Please see new version below:

(Results, page 11/12). “Notably, despite the absence of a robust group-level HR effect, NE and RR responses remained negatively correlated across trials (Fig. 4e), indicating that HR dynamics continued to track the magnitude of noradrenergic suppression at the individual-response level”.

(5D) The two colors in Figure 4 are similar and hard to distinguish.

We agree that the colors are hard to separate and have altered them to make them easier to separate.

(5E) The correlations shown in Figure 4j seem to be driven by just two of the cases. Are the effects significant when outliers are removed?

We thank the reviewer for raising this point. To assess whether the observed correlation in Figure 4j was disproportionately driven by a small number of data points, we performed a formal outlier analysis using the ROUT method (Q = 1%). This analysis did not identify any statistical outliers in the dataset. Therefore, we did not have an objective basis for excluding any observations from the analysis.

(5F) Page 10: Were there any differences in memory performance between the Arch and YFP conditions?

We thank the reviewer for this question. The memory experiments were based on a previously published dataset (Kjaerby, Andersen et al., 2022), in which the primary objective was to assess the effect of LC suppression on sleep spindle dynamics and memory consolidation. In the present study, we performed an additional analysis by extracting heart-rate (RR) measures from these recordings to evaluate whether cardiac responses could serve as a biomarker of LC-mediated noradrenergic regulation.

However, due to technical limitations (EMG recording often suffers from noise) in extracting reliable RR signals from all animals in this dataset, the number of subjects available for this secondary analysis was reduced. All these considerations are described in Methods/Mice. As a result, we were not sufficiently powered to perform a direct statistical comparison of memory performance between Arch and YFP groups based on RR measures alone. Instead, we examined whether RR responses to LC suppression predicted behavioral performance across animals. When pooling Arch and YFP conditions, we observed a correlation between the magnitude of the RR response and subsequent memory performance, suggesting that heart-rate dynamics reflect noradrenergic modulation relevant for memory consolidation.

To avoid overinterpretation, we have therefore limited our conclusions to reporting this association rather than making direct group-level comparisons between Arch and YFP animals, and we have clarified this point in the revised manuscript.

(Results, page 13): “Interestingly, across pooled Arch and YFP animals, larger RR increases following LC suppression were associated with better subsequent memory performance. Additionally, RR and NE responses to LC suppression were negatively correlated indicating that animals showing stronger NE reductions also exhibited larger RR changes. Together, these findings suggest that heart-rate dynamics covary with noradrenergic responses during sleep and may reflect physiological processes relevant for sleep-dependent memory consolidation.”

(5G) Page 10: "We found a correlation between RR responses to LC suppression and sigma power, suggesting that a stronger HR reduction response is linked to higher spindle power." It should be noted in the text that the correlation was not specific to sigma (it was also seen for theta and beta, Figure 4i).

We agree with reviewer that this should be highlighted. We have changed the sentence:

(Results, page 12). “Furthermore, we found a correlation between RR responses to LC suppression and sigma power, suggesting that a stronger HR reduction response is linked to higher spindle power; similar correlations were also observed in the theta and beta frequency ranges (Fig. 4h-i).”

(6) It is not clear which of the sigma power and RR interval findings do/do not exactly line up between the mice and humans. It could be helpful to have a table comparing them. For instance, was the finding in humans that pre-HRB sigma power was positively associated with slowing in heart rate after the HRB also seen in mice? Was there evidence in mice (as seen in the human sample) that sleep-dependent memory improvement was associated with pre-HRB sigma power?

We thank the reviewer for this thoughtful comment and agree that the cross-species comparisons could be communicated more clearly. Our intention was not to imply exact one-to-one correspondence between all mouse and human findings, but rather to examine whether central–autonomic coupling surrounding phasic heart-rate events shows conserved features across species while acknowledging species-specific physiological differences.

Importantly, the mouse and human analyses were designed to address related but not identical questions. In mice, we leveraged optogenetic LC suppression to probe a more causal relationship between noradrenergic activity, heart-rate slowing, spindle-related dynamics, and memory consolidation. Previous work using this dataset demonstrated that 2-minute LC suppression robustly enhances spindle density and that spindle enhancement correlates with improved memory performance. In the present study, we therefore asked whether heart-rate slowing covaries with this LC-mediated spindle/memory relationship, supporting HR as a potential biomarker of these restorative processes.

By contrast, causal manipulation of LC activity is not feasible in humans. Instead, we focused on naturally occurring HRBs as putative downstream signatures of phasic LC– NE activity, motivated by our mouse findings that NE increases precede HR accelerations. We observed conserved coupling between sigma activity and HR dynamics across species, although the temporal profile differed, with sigma activity occurring closer to the HRB in humans than in mice. These temporal differences may reflect species-specific differences in cardiac and sleep physiology.

Regarding the reviewer’s specific questions, the positive relationship between pre-HRB sigma power and post-HRB heart-rate slowing was tested in humans, where this metric showed the strongest relationship to behavioral outcome. We did not directly test the same measure in mice because, unlike humans, HR recovery following HRBs did not show a pronounced baseline shift (Fig. 5b), limiting the interpretability of this comparison. Similarly, we did not directly correlate pre-HRB sigma power with memory performance in mice, as the more causal LC suppression paradigm already demonstrated a spindle–memory relationship in this species and was the focus of our mechanistic analysis.

To reduce confusion, we have revised the Results conclusion to make a clearer overview:

(Results, page 14): “In conclusion, mice and humans displayed evidence of conserved autonomic-central coupling, reflected in coordinated HR and sigma power dynamics surrounding phasic cardiac events.

However, the temporal relationship between these events differed across species, likely reflecting differences in sleep and cardiovascular physiology. In mice, causal manipulation of LC activity demonstrated that heart-rate dynamics covary with LC-mediated noradrenergic and spindle-related processes linked to memory consolidation. In humans, sigma power preceding HR bursts was associated with both post-HRB heart-rate slowing and sleep-dependent memory improvement, suggesting that autonomic–central coupling surrounding HR events may provide a translational marker of restorative sleep processes (Fig. 5n).”

(7) Page 18: It is not clear if the sex of mice was balanced across controls and optogenetics groups.

We thank the reviewer for this important comment and agree that the sex distribution should be reported more clearly. These experiments relied on the availability of animals from the heterozygous TH-Cre transgenic line, and given the relatively small cohort sizes, perfect balancing across sex and experimental groups was not always feasible.

For the LC activation experiments, the sex distribution was: YFP: 2 male / 2 female; ChR2: 4 male / 2 female. For the LC suppression experiments, the distribution was: YFP: 4 female; Arch: 3 female / 1 male.

Although the groups were not perfectly sex balanced, we had no strong reason to expect robust sex-dependent differences in the physiological effects of these optogenetic manipulations, particularly given the relatively strong and acute nature of the intervention. At the same time, we acknowledge that the present study was not powered to assess sex as a biological variable, and subtle sex-dependent effects therefore cannot be excluded. To improve transparency, we have now clarified the sex distribution in the Methods/Mice section.

Reviewer #2 (Public review):

Summary:

The major part of this study reproduces previously published findings in both mice and humans and provides incremental analyses on these findings. In essence, the work reaffirms the presence of coordinated infraslow fluctuations in sigma power and heart rate during NREM sleep. It further confirms previous findings that coordination depends on noradrenaline-releasing neurons in the locus coeruleus. Also supporting previously published work in mice and humans, the authors describe a link between the strength of these infraslow fluctuations and memory consolidation in mice and humans.

Strengths:

The authors successfully replicate key previously reported phenomena across both mice and humans. Confirmatory studies and demonstrations of reproducibility are essential for progress in neuroscience. To maximize their value, such studies should clearly acknowledge their confirmatory nature and carefully situate what, in their view, are novel results, going beyond existing literature.

Weaknesses:

The authors' interpretation of their data needs to be revised. Many of their claims regarding the mechanistic basis of their findings and the predictive value of their correlative datasets are not supported by the available evidence.

In the present manuscript, several citations of literature on the work they reproduce lack precision or completeness, which reduces transparency and obscures how the reported findings relate to previously established results.

We thank Reviewer 2 for the thoughtful comment regarding positioning of our findings relative to the literature, and caution in mechanistic interpretation. In response, we have revised the Introduction, Results, and Discussion to more clearly acknowledge foundational studies in this area and to better clarify how the present work extends beyond them.

We agree that prior work has demonstrated infraslow coupling between sigma activity, norepinephrine (NE) dynamics, and heart rate (HR), and has established a role for the locus coeruleus (LC) in coordinating these oscillations. However, cardiac measures in these studies were typically treated as secondary observations rather than as primary experimental targets. A central goal of the present study was therefore to provide a systematic and mechanistically grounded characterization of NE-mediated HR dynamics during sleep across multiple timescales, including infraslow oscillations, sleep–wake transitions, and causal manipulations of LC activity.

Importantly, we also aimed to relate infraslow HR fluctuations to the very-low-frequency (VLF) component of heart rate variability (HRV), which remains comparatively under-characterized and mechanistically unresolved in the clinical HRV literature. By linking LC activity, NE dynamics, and HR fluctuations across behavioral states, our findings provide a biologically grounded framework that may help explain this component of HRV.

A second major objective of the study was translational. Because direct LC recordings are not feasible in humans, we asked whether cardiac dynamics alone could reflect the infraslow, memory-consolidating potential of sleep and thus serve as a noninvasive biomarker. By directly manipulating LC activity and demonstrating corresponding changes in HR dynamics, our results strengthen the mechanistic rationale for using HRV—particularly its VLF component—as an accessible proxy of LC-dependent sleep physiology.

We therefore respectfully disagree with the suggestion that the present study does not provide novel insight. Rather, the revised manuscript now more clearly emphasizes that our contribution lies in (i) systematically characterizing NE-dependent HR dynamics across sleep states, (ii) linking these dynamics to the poorly understood VLF component of HRV, and (iii) establishing a causal and translational framework for using cardiac measures as markers of LC-mediated sleep processes.

We hope the reviewer finds that the revised Introduction and Discussion better highlight both the existing literature and the specific advances provided by the present work.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

I have been convinced by discussions about the replicability crisis that it should be a standard practice to share data upon publication in a publicly accessible online repository such as Open Science Framework or OpenNEURO. Making data available can increase the impact of the research study. Simply stating "data available upon request" as done in the current draft is not sufficient as, unfortunately, when data are not shared upon publication in a public repository, it can be impossible to gain access by request from the researchers - the majority of requests from other researchers to obtain data are not complied with (e.g., Vanpaemel, Vermorgen, Deriemaecker, & Storms, 2015; Wicherts, Bakker, & Molenaar, 2011).

We fully agree with the reviewer about the importance of data sharing for transparency, reproducibility, and maximizing the impact of research. Consistent with these principles, we will make the human dataset publicly available on the Open Science Framework upon publication and have added the link to this repository in the manuscript under Data Availability (https://osf.io/g6emj/). With respect to the mouse dataset, we respectfully note that this dataset is currently the subject of multiple planned and ongoing analyses that extend beyond the scope of the present manuscript. Releasing these data publicly at this stage could compromise these efforts and lead to potential misinterpretation prior to completion of the full analytic pipeline. For this reason, we believe that it would not be appropriate to share the mouse data in a public repository at this time. However, we remain committed to transparency and will make the mouse data available upon reasonable request during this period, with the intention of publicly releasing the dataset once the planned analyses are complete.

(1) p. 4: There is some orphan text at the top of the page ("marker of Alzheimer's disease. Furthermore, maintaining LC neural density prevents neurodegeneration (13). Given the reported reduction in HRV in aging and Alzheimer's disease, suppressed").

Thank you. We have removed the orphan text.

(2) p. 22: What does NFR refer to?

Novel-to-familiar ratio. We have removed the abbreviation from the main text. It is now only used in the figure.

(3) I might have missed this information, but it was not clear to me how the epochs to be analyzed were selected, and for the mice, how much wake vs. sleep they included.

We thank the reviewer for this comment and apologize that the epoch selection criteria were not sufficiently clear. They were mentioned under Methods. For the optogenetic analyses in mice, epochs were selected based on NREM sleep including microarousals to ensure that physiological responses were evaluated within stable sleep conditions while allowing for natural brief interruptions.

For the LC activation (ChR2) experiments, stimulation epochs were included only if NREMinclMA began at least 30 s before laser onset and continued for at least 30 s after laser onset. For the LC suppression (Arch) experiments, the criterion was similarly ≥30 s of NREMinclMA prior to laser onset, but extending ≥60 s following laser onset to accommodate the longer suppression response profile.

Thus, analyses were intentionally restricted to sleep periods, and wakefulness was not included except where it emerged naturally as an outcome of the manipulation or transition under investigation. We have now clarified these inclusion criteria in the Methods/ Event marker selection section to improve transparency.

(4) In the Supplement, Figure 3 appears before Figure 2.

Thanks. We have corrected it.

Signed, Mara Mather

Reviewer #2 (Recommendations for the authors):

The authors' interpretation of their data needs to be revised. Many of their claims regarding the mechanistic basis of their findings and the predictive value of their correlative datasets are not supported by the available evidence.

There are three major directions in which this study would need to be revised.

Part 1 - Literature citations of both mouse and human literature need to be revised, and citations placed in a manner that accurately reflects what has been previously done. In detail:

(1.1) The statement on p. 4 regarding "the extent to which phasic infraslow NE fluctuations ...is not well understood" disregards a previous publication in which closed-loop optogenetic stimulation of LC was already shown to regulate HR variations on the infraslow time scale (10.1016/j.cub.2021.09.041).

We thank the reviewer for this important point. The study by Osorio-Forero et al. (2021) was already cited as ref. 4 in our original manuscript and in the discussion, we specifically highlighted this study: ‘These findings build on Osorio-Forero et al. (35) as well as other studies’; however, we agree that our wording did not sufficiently emphasize its key finding that closed-loop optogenetic manipulation of LC activity can coordinate infraslow heart-rate fluctuations and spindle clustering during NREM sleep.

We have revised the relevant paragraph in the introduction to explicitly acknowledge that ref. 4 demonstrated causal coordination between LC activity, sleep spindle dynamics, and heart-rate fluctuations. We have also refined our statement of the knowledge gap to clarify which novel questions our study addresses. We believe these revisions more accurately position our work within the existing literature and clearly distinguish our contributions from prior studies.

We have updated the introduction in several places to address these comments:

(Introduction, page 4). “It has previously been reported that HR fluctuates at similar infraslow frequencies as NE fluctuations and sigma power in mice (Lecci et al., 2017; Osorio-Forero et al., 2021) and that phasic HR fluctuations correlate with infraslow changes in pupil diameter (Carro-Domínguez et al., 2025), a proxy for changes in NE levels (Murphy et al., 2014; Reimer et al., 2016). Importantly, optogenetic manipulation of the LC has demonstrated that infraslow LC activity coordinates sleep spindle clustering and heart rate fluctuations during NREM sleep (Osorio-Forero et al., 2021). Together, these findings support a functional coupling between the central LC–NE system and peripheral cardiac dynamics, further supported by findings that HR increases accompany MAs during NREM sleep (Carro-Domínguez et al., 2025;

Osorio-Forero et al., 2025).”

(Introduction, page 4/5). “Due to the frequency overlap with VLF HRV, we wondered if central infraslow NE dynamics could be linked to this poorly understood HRV indicator. Furthermore, are infraslow NE fluctuations directly reflected by heart-rate variability under different physiological states or does LC–HR coupling scales differently with LC output? Specifically, if infraslow NE oscillations display faster frequencies - as occurs during sleep fragmentation - will cardiac dynamics exhibit corresponding changes? Conversely, given that stronger infraslow NE dynamics correlate with memory consolidation through their regulation of sleep spindles, could the peripheral autonomic signatures provide an accessible cross-species biomarker of spindle-dependent memory consolidation? Addressing these questions could help bridge mechanistic insights into LC-mediated sleep regulation with established HRV metrics used in human physiology.”

(1.2) In this same published paper, optogenetic stimulation of LC was already used to "determine the causal relationship..." (p.6). However, the authors do not cite these data.

We thank the reviewer for this comment. As noted in our response to Comment 1.1, ref. 4 was already cited in the Introduction, and we have now revised that section to more explicitly emphasize this paper. In addition, we have modified the wording in the Results section.

(Results, page 8). “After finding the inverse correlation between NE and RR in natural sleep transitions, we next sought to further characterize the causal influence of LC activity on HR dynamics, using a closed-loop optogenetic approach to modulate NE oscillatory frequency during NREM sleep.”

(1.3) The authors' speculation about LC-induced sympathetic and parasympathetic actions is premature: this study does not provide pharmacological experiments in this direction. However, two published studies implied a parasympathetic mechanism linked to infraslow fluctuations of LC activity (10.1016/j.cub.2021.09.041, 10.1016/j.cub.2017.12.049). These findings should be appropriately cited. While HR decelerations may reflect compensatory autonomic responses, there is no direct evidence in the present study that these effects are sympathetically mediated. An alternative, and equally plausible, interpretation is enhanced parasympathetic activity. This distinction is particularly important given that the observed increases in mean HR during LC stimulation cannot distinguish between reduced parasympathetic activity and increased sympathetic drive.

We thank the reviewer for raising this important point and agree that the current study does not provide direct mechanistic evidence to disentangle sympathetic versus parasympathetic contributions to LC-mediated heart rate regulation. This was not the intention of our study; rather, the relevant discussion section was meant to provide mechanistic interpretations and hypotheses based on the observed physiology. To avoid overstating our conclusions, we have revised the wording to more clearly emphasize the speculative nature of these interpretations. Furthermore, we have added the suggested references demonstrating that muscarinic blockade reduces heart rate and pupil fluctuations during sleep, which support the possibility of a parasympathetic contribution to infraslow LC-related dynamics.

(Discussion, page 16/17). “Prior findings demonstrate that LC is linked to the autonomic nervous system. Stimulation of LC projections decrease parasympathetic cardiac vagal activity (Wang et al., 2014) and also influences sympathetic output through direct projections to the preganglionic cells in the sympathetic nervous system (Karemaker, 2017; Nygren and Olson, 1977; Samuels and Szabadi, 2008). While the mechanisms generating VLF HRV are not well defined (Armour, 2003; Shaffer et al., 2014; Wang et al., 2014), there is a clear parasympathetic component (Taylor et al., 1998). This combined with the ability of pharmacological blockage of the parasympathetic system to block infraslow oscillations of HR (Osorio-Forero et al., 2021) and pupil diameter during sleep (Yüzgeç et al., 2018), led us to expect a slowing of HR during LC suppression due to parasympathetic disinhibition.”

(1.4) It is not clear why prefrontal NE signals were associated with HR fluctuations. Literature evidence indicates other brain areas that are functionally more directly linked to autonomous fluctuations.

We thank the reviewer for raising this point. We do not speculate that mPFC NE activity is causally linked to heart rate fluctuations or that the mPFC directly mediates the observed autonomic dynamics. Instead, mPFC NE signaling was used as an experimentally accessible readout of LC activity. Importantly, accumulating evidence suggests that infraslow NE oscillations are globally coordinated phenomena that are expressed across multiple brain regions during sleep, making mPFC NE a valid proxy for LC-driven neuromodulatory state dynamics. We have clarified this in the result section:

(Results, page 6). “mPFC was selected as a cortical readout of LC-mediated norepinephrine dynamics, as infraslow NE fluctuations are coordinated across widespread brain regions.”

(1.5.a) The lack of effect of Arch-inhibition of LC on infraslow NE signals is concerning. Prior work showed that bilateral LC inhibition does affect NE signals and also infraslow sigma power fluctuations (10.1016/j.cub.2021.09.041, 10.1038/s41593-024-01822-0). This discrepancy should be explicitly acknowledged and discussed on p.9.

We thank the reviewer for highlighting this important point. To clarify, we did observe a robust effect of LC inhibition on noradrenergic signaling, with clear suppression of the NE signal following Arch-mediated LC inhibition (Fig. 4c), aligned to laser onset. In addition, LC suppression increased neuronal synchronization, including enhanced sigma power relative to YFP controls (Fig. 4h), consistent with previous reports showing that reduced LC activity promotes synchronized sleep-related oscillations. Thus, we do not interpret our findings as indicating an absence of LC suppression effects.

The apparent discrepancy relates specifically to the absence of a statistically significant group-level shift in infraslow NE or HRV power during NREM sleep, rather than the efficacy of the manipulation itself. We note that our inhibition paradigm was intentionally mild, consisting of repeated 2-minute suppression periods separated by 4-minute intervals, and was designed to introduce subtle shifts within physiological ranges rather than globally reorganize infraslow sleep structure. Accordingly, we consider the immediate NE and heart-rate responses to LC inhibition to be the most sensitive physiological readouts of LC-mediated regulation in this context.

We also note that the studies cited by the reviewer used different suppression paradigms, including more frequent manipulations. Thus, while our suppression scheme did not result in detectable infraslow reorganization, we do not believe they reflect an ineffective LC suppression as we demonstrated a clear NE reduction in response to time-locked LC suppression.

We have added a sentence to the result section to explain the lack of effect infraslow power:

(Results, page 12). “This likely reflects the relatively mild and intermittent LC suppression paradigm, which was designed to remain within physiological ranges and therefore did not globally reorganize infraslow sleep dynamics.”

(1.5.b) Additionally, the authors state in the Discussion that they "find no consistent modulation in HR during LC suppression, suggesting that the LC-HR connection is more strongly associated with sympathetic activity rather than parasympathetic inhibition". However, the results primarily demonstrate an absence of modulation in mean HR, while preserving a significant relationship between RR intervals and stimulation. This indicates a modulation of heart rate variability, even in the absence of changes in average HR. Notably, such variability-related effects may fall outside the VLF range and could instead involve higher-frequency components.

We thank the reviewer for this important clarification. We agree that our original wording may have conflated the absence of modulation in mean HR with the absence of autonomic modulation more generally. To address this point, we revised the Discussion to emphasize that LC activity may influence VLF HRV through sympathetic activation and/or indirect modulation of cardiac vagal activity, even in the absence of robust changes in average HR. We have softened our previous interpretation that the LC-HR relationship is primarily sympathetic in nature and instead discuss a more nuanced interaction between sympathetic and parasympathetic influences on HRV dynamics.

(Discussion, page 16/17). “Prior findings demonstrate that LC is linked to the autonomic nervous system. Stimulation of LC projections decrease parasympathetic cardiac vagal activity (Wang et al., 2014) direct projections to the preganglionic cells in the sympathetic nervous system (Karemaker, 2017; Nygren and Olson, 1977; Samuels and Szabadi, 2008). While the mechanisms generating VLF HRV are not well defined (Armour, 2003; Shaffer et al., 2014; Wang et al., 2014) there is a clear parasympathetic component (Taylor et al., 1998). This combined with the ability of pharmacological blockage of the parasympathetic system to block infraslow oscillations of HR (Osorio-Forero et al., 2021) and pupil diameter during sleep (Yüzgeç et al., 2018) led us to expect a slowing of HR during LC suppression due to parasympathetic disinhibition. Interestingly, we found no consistent modulation in HR during LC suppression, suggesting that the LC-HR connection may also somehow be driven by sympathetic outflow. Previous research had indicated that LF power may also represent sympathetic activity, but this interpretation has been challenged due to the mixed contribution of both autonomic branches (Houle and Billman, 1999; Japundzic et al., 1990; Reyes et al., 2013). LF and HF ratio (LF/HF) were traditionally thought to reflect balance between sympathetic and parasympathetic activities (i.e., the sympatho-vagal balance), though currently considered as an oversimplification of non-linear integration of autonomic signals (Billman, 2013; Pagani et al., 1986). Our findings implicate the VLF may offer a precise marker for central arousal states, given its overlaps with infraslow phasic fluctuations of LC-NE levels. Although parasympathetic activity appears important for the expression of VLF oscillations, growing evidence suggests that VLF dynamics reflect broader interactions between the heart and autonomic nervous system rather than simple sympathetic or parasympathetic control alone (Armour, 2003; Shaffer et al., 2014). Within this framework, infraslow LC–NE dynamics may represent one central contributor to these slow cardiac fluctuations during sleep either directly through sympathetic activation or indirectly by inhibition of cardiac vagal activity.”

(1.6) The relationship between sigma power fluctuations and HR is different in humans than in mice. This has been shown before (10.1126/sciadv.1602026, 10.1038/s41593025-02159-y). This work should be mentioned on p. 11.

We already acknowledge prior studies demonstrating species differences in the relationship between sigma power fluctuations and heart rate in the discussion section. However, to accommodate the reviewer’s comment and improve clarity for the reader, we have now also added the suggested references to the Results section, where the relationship between sigma power fluctuations and HR is first discussed.

(Results, page 14). “These temporal differences may reflect species-specific physiology differences in cardiac timescales, which has also been previously reported (Bergel et al., 2025; Carro-Domínguez et al., 2025; Lecci et al., 2017).”

(1.7) Correlations between the strength of infraslow sigma power fluctuations and memory consolidation have been published and should be discussed (10.1126/sciadv.1602026). It is surprising to see that correlations with learning in humans are done using pre-HRB sigma peaks rather than heart rate. This is a measure that is very close to the one used by Lecci et al.; this similarity should be clearly acknowledged. The way the data are currently presented limits this manuscript's novelty, also in its translational aspect.

We agree that the work by Lecci et al. (2017) established an important relationship between the association of infraslow sigma power fluctuations and memory consolidation, which is highly relevant to our findings.

Importantly, the underlying infraslow fluctuations in neuromodulatory tone are increasingly recognized as key regulators of sleep spindle dynamics (sigma power), including from our own previous work demonstrating that direct manipulation of locus coeruleus–norepinephrine infraslow rhythms alters spindle organization and sleep continuity. Thus, our findings are conceptually aligned with prior studies linking sigma fluctuations to memory consolidation.

However, we would like to clarify an important distinction in our translational approach. While Figure 4 demonstrates that heart rate dynamics during sleep can predict memory performance in mice, Figure 5 was designed to address the translational potential of these findings in humans, where direct neuromodulatory readouts are not readily accessible. Here, we deliberately focused on heart rate bursts and their associated sleep dynamics as a clinically tractable physiological measure.

We acknowledge that the pre-HRB sigma increase may appear conceptually similar to the measure used by Lecci et al.; however, our approach is not equivalent. Rather than selecting spindle or sigma peaks themselves, we aligned analyses to heart rate accelerations and examined the robust upregulation of sigma activity preceding these events. In this framework, sigma activity serves as a physiological readout linked to autonomic dynamics, rather than being the primary anchor of analysis. We chose this measure because heart rate bursts are influenced by multiple physiological factors, and the associated sigma dynamics provided the clearest and most robust relationship with memory outcomes in the human dataset.

We have revised the Discussion to more clearly acknowledge the similarity to prior work. We believe our findings extend prior observations by providing evidence that sleep-related heart rate fluctuations may serve as a non-invasive readout of the memory-preserving function of sleep.

(Discussion, page 18). “Previous work in humans demonstrated that the strength of infraslow sigma oscillations correlates with sleep-dependent memory consolidation in humans (Lecci et al., 2017).”

(1.8) Conclusions as to whether sigma fluctuations might be slightly slower and less powerful in mice are not justified. More work is required to determine which infraslow manifestations are most useful for cross-species comparisons. Moreover, little is currently known about LC activity in human sleep. A careful look into how pupil diameter correlates with sigma power should provide clues for further discussion (see Carro-Dominguez et al). A detailed study of infraslow fluctuations in human sleep should also be discussed https://doi.org/10.1101/2024.11.06.620875.

We thank the reviewer for this thoughtful comment. We agree that our original phrasing suggesting that sigma dynamics in mice may be “slower and less powerful” than in humans was overly interpretive. We have removed this sentence from the Results section. In addition, we have expanded the Discussion to more thoroughly integrate recent human literature on infraslow sleep dynamics.

(Discussion, page 19). “Importantly, infraslow fluctuations of sigma power in human sleep have received growing attention. Recent work demonstrates that the infraslow fluctuation of sigma power segments N2 sleep into functional phases associated with arousal and memory-related sleep markers (Dimitriades et al., 2024). Complementary findings using pupillometry show that pupil diameter fluctuates on similar infraslow timescales during NREM sleep and is inversely related to spindle clustering, providing indirect evidence that arousal-related noradrenergic dynamics shape human sleep microstructure (Carro-Domínguez et al., 2025). However, LC activity during human sleep remains inferred rather than directly measured, and systematic perturbation studies linking LC output to spindle–autonomic coupling in humans are currently lacking. Together, these observations underscore both the promise and the current limitations of cross-species comparisons of infraslow sleep dynamics.”

(1.9) Citation of literature should be as explicit as possible. Referring to "many studies rely on plasma levels..." while including some that actually did real-time fiber photometric measures is misleading.

We thank the reviewer for this suggestion and agree with the reviewer about the importance of accurately representing prior studies. We have now updated the discussion to clarify this.

(Discussion, page 15). “Many studies have linked HR to NE (Fawaz and Simaan, 1963; Sundaram et al., 1991; Tanoue et al., 2022; Watson et al., 1979). Many rely on plasma levels of NE, which, while linked to central NE (Gurguis and Uhde, 1998), has low temporal resolution, making causal interpretations harder. In recent years, the use of biosensors and fibre photometry allows for very reliable estimate of the temporal dynamics of NE changes making association to HR more precise (Osorio-Forero et al., 2021).”

(1.10) Regarding the discussion on the baroreflex: The emphasis is placed predominantly on sympathetically mediated effects. However, the description of the baroreflex loop is incomplete, as it overlooks the substantial contribution of parasympathetic modulation. In particular, heart rate adjustments within the baroreflex are primarily mediated by parasympathetic mechanisms.

We agree with reviewer that this important notion should be added. We have rephrased the discussion as below:

(Discussion, page 20). “These neurons suppress the activity of the rostral ventrolateral medulla, ultimately resulting in reflex parasympathetic activation with sympathetic inhibition lowering the HR (Aicher et al., 2000; Lanfranchi and Somers, 2002).”

(1.11) Regarding the interpretation of HRV analysis, the discussion places disproportionate weight on sympathetic modulation in the interpretation of HRV metrics. This framing is inconsistent with recent conceptual clarifications, including a recent Nature Reviews Cardiology article by Menuet et al. (10.1038/s41569-02501160-z), which cautions against simplistic low-frequency/high-frequency (LF/HF) interpretations of autonomic balance.

We thank the reviewer for this thoughtful comment. We agree that HRV frequency bands should not be interpreted as exclusive markers of specific autonomic branches. In the original manuscript, we cited Billman (2013), which challenges the validity of LF/HF as a measure of sympatho-vagal balance, to acknowledge these conceptual limitations. However, we recognize that some of our phrasing, particularly in the section discussing compensatory HR decelerations, may have implied branch-specific dominance.

We have updated the Discussion section as follows:

(Discussion, page 17). “Interestingly, we found no consistent modulation in HR during LC suppression, suggesting that the LC-HR connection may also somehow be driven by sympathetic outflow. Previous research had indicated that LF power may also represent sympathetic activity, but this interpretation has been challenged due to the mixed contribution of both autonomic branches (Houle and Billman, 1999; Japundzic et al., 1990; Reyes et al., 2013). LF and HF ratio (LF/HF) were traditionally thought to reflect balance between sympathetic and parasympathetic activities (i.e., the sympatho-vagal balance), though currently considered as an oversimplification of nonlinear integration of autonomic signals (Billman, 2013; Pagani et al., 1986). Our findings implicate the VLF may offer a precise marker for central arousal states, given its overlaps with infraslow phasic fluctuations of LC-NE levels. Although parasympathetic activity appears important for the expression of VLF oscillations, growing evidence suggests that VLF dynamics reflect broader interactions between the heart and autonomic nervous system rather than simple sympathetic or parasympathetic control alone (Armour, 2003; Shaffer et al., 2014). Within this framework, infraslow LC–NE dynamics may represent one central contributor to these slow cardiac fluctuations during sleep either directly through sympathetic activation or indirectly by inhibition of cardiac vagal activity.”

We have also replaced the title of the Discussion section title “Locus-coeruleus-mediated sympathetic control drives compensatory heart rate decelerations” with “Locus-coeruleus activation drives compensatory heart rate decelerations via autonomic feedback mechanisms”

(1.12) In particular, the manuscript attributes VLF power primarily to sympathetic activity, despite evidence that very-low-frequency RR-interval oscillations are strongly dependent on parasympathetic integrity. Notably, parasympathetic blockade has been shown to nearly abolish VLF oscillations in humans (Taylor et al., Circulation, 1998; doi:10.1161/01.CIR.98.6.547).

We thank the reviewer for highlighting this important point. It was not our intention to imply that VLF HRV is driven primarily by sympathetic activity or to disregard the important role of parasympathetic integrity in shaping VLF oscillations. Our intention was to present the multifactorial and incompletely resolved nature of the VLF component; however, if our wording can be interpreted otherwise, we agree that clarification is warranted.

In response, we have revised the manuscript to better reflect the current understanding of VLF physiology. Specifically, we now make clearer that, while VLF HRV remains less mechanistically defined than HF and LF HRV, substantial evidence supports a strong parasympathetic contribution to the expression of VLF oscillations. We have incorporated the reviewer-suggested references within this comment and others and clarified this point throughout the revised manuscript.

(Discussion, page 16). “While the mechanisms generating VLF HRV are not well defined (Armour, 2003; Shaffer et al., 2014; Wang et al., 2014) there is a clear parasympathetic component (Taylor et al., 1998).”

(Discussion, page 17). “Although parasympathetic activity appears important for the expression of VLF oscillations, growing evidence suggests that VLF dynamics reflect broader interactions between the heart and autonomic nervous system rather than simple sympathetic or parasympathetic control alone (Armour, 2003; Shaffer et al., 2014).”

(1.13) Moreover, based on work by Armour (2003) and Kember et al. (2000, 2001), the VLF rhythm is thought to emerge from stimulation of afferent sensory neurons within the heart, further arguing against a purely sympathetic interpretation. Together, these findings indicate that the discussion overemphasizes sympathetic mechanisms and underrepresents the contribution of parasympathetic and afferent cardiac pathways to HRV, particularly in the VLF range.

We thank the reviewer for this important point. Our response to this comment is largely aligned with our response to comment (1.12). In the revised Discussion, we now more clearly acknowledge that VLF oscillations likely arise from more complex cardioautonomic interactions than simple sympathetic and parasympathetic innervation alone. At the same time, we have intentionally avoided an extensive discussion of these mechanisms, as our experimental design does not directly address these pathways.

Part 2 - There are a number of conceptual issues that need more careful elaboration:

(2.1) A highly problematic point throughout this study is the choice of AUCs, notably of NE signals and RR intervals, rather than the signal amplitudes. It confuses the correlations to events of different durations, such as MAs and wakefulness. It is not possible to draw conclusions of the kind "cardiac rhythm are tightly coupled with the infraslow phasic NE ..." because such statements ignore that AUCs conflate amplitude and duration of a signal.

We thank the reviewer for this comment and agree that the distinction between amplitude- and duration-related signal features is important. We deliberately chose AUC measures because our intention was to capture the overall physiological response over time, including both the magnitude and temporal evolution of the signal, rather than relying solely on a single peak value. In this context, AUC provides information about the shape and sustained nature of NE and RR changes within a defined time window, which we considered particularly relevant for temporally dynamic responses.

At the same time, we appreciate the reviewer’s concern that AUC may conflate response amplitude and duration, particularly when comparing events of different lengths such as microarousals and wakefulness. To minimize this issue, the AUC windows were intentionally kept relatively short and fixed, thereby limiting the influence of prolonged wake episodes or differences in transition duration on the measure. Thus, our intention was not to quantify the total duration of awakenings, but rather the immediate physiological response profile surrounding the event.

To address this concern more directly, we compared amplitude- and AUC-based measures across all mice. The results are now shown in Suppl. Figure 1g. We found a strong correspondence between amplitude and AUC measurements for NE signals, indicating that the observed relationships are not dependent on the choice of metric. A similar, albeit weaker, relationship was observed for RR responses. This likely reflects physiological constraints on heart-rate dynamics, where the initial heart-rate acceleration is relatively similar across vigilance-state transitions (Figure 1e), while the duration of the response differs substantially. As a result, amplitude measures may underestimate differences between transitions, whereas AUC better captures the extent of the cardiac response. Importantly, direct comparison of NE and RR amplitudes still revealed a significant relationship, supporting the overall conclusion that cardiac and noradrenergic responses are coupled. However, this relationship was weaker than that observed using AUC measures, suggesting that incorporating temporal aspects of the response captures additional biologically relevant information.

In addition to the figure, we have revised the Results section to include these considerations:

(Results, page 7): “These comparisons were performed using area under the curve (AUC) estimates of NE and R-R responses. Importantly, comparing AUC and peak amplitude measures showed a similar overall relationship, and the coupling between NE and RR remained significant when only peak amplitudes were considered (Suppl. Fig. 1g), indicating that the findings are not solely driven by the response duration aspect of AUC. The somewhat stronger relationship observed with AUC-based measures may reflect rapid saturation of heart-rate responses across vigilance-state transitions, making response persistence an informative component of the physiological signal.”

(2.2) A next problematic aspect of the study is the analysis of noradrenergic signals and HR at transitions (e.g., from NREM sleep to sleep wakefulness or to microarousals). The abstract does not mention these data, leaving open how they fit into the paper's message. There are also two problems with it: a) Noradrenaline levels increase with wakefulness, as do many other neuromodulators. This is not novel. Furthermore, why use an AUC measure for 0-25 s when MAs last only 5 or 15 s? b) biosensor signal comparisons are difficult to make for state transitions, because blood flow changes and modifies the fluorescent signal.

We thank the reviewer for these comments and appreciate the opportunity to clarify the rationale and interpretation of these analyses.

Regarding the inclusion of vigilance-state transitions, our intention was not to claim novelty in the observation that NE levels increase during wakefulness. Rather, these analyses were included to provide an additional physiological context in which to examine the coupling between NE and heart-rate dynamics. Specifically, the transition analyses allowed us to determine whether graded changes in NE across sleep-to-wake transitions were mirrored by corresponding RR changes (Fig. 1d–e) and whether these responses covaried (Fig. 1f), thereby strengthening the overall conclusion that cardiac dynamics track noradrenergic signaling across naturally occurring sleep-state fluctuations.

Regarding the use of AUC measures, we deliberately chose this metric because it captures the overall physiological response over a defined time window, including both magnitude and temporal evolution, rather than relying solely on a peak value. In this context, AUC was intended to reflect differences in the overall response profile, for example that NE and RR responses during microarousals may return more rapidly toward baseline than during sustained wakefulness. To minimize the influence of differing event durations, the analysis window was kept fixed (0–25 s) across all transition types. Importantly, we directly compared AUC- and amplitude-based measures and found that they produced largely similar relationships (Suppl. Fig. 1g). The coupling between NE and RR responses remained significant when only peak amplitudes were considered, indicating that the observed relationship is not solely driven by response duration. However, the relationship was somewhat stronger with AUC-based measures, likely because heart-rate responses rapidly saturate across vigilance-state transitions, making response persistence an informative component of the physiological signal. As mentioned in the previous comment, we have added a Figure and new result text to highlight this.

Regarding the concern about blood-flow related artifacts in fluorescent biosensor signals, we agree that hemodynamic contamination is an important consideration for neuromodulator recordings and applies broadly to biosensor-based measurements, including analyses of infraslow fluctuations. To minimize this issue, ΔF/F calculations were performed using the isosbestic control channel, which serves to correct for movement- and hemodynamic-related signal fluctuations. Furthermore, hemodynamic artifacts typically occur rapidly at state transitions, whereas GRAB-NE signals display slower dynamics. If uncorrected hemodynamic contamination strongly influenced the signal, we would expect abrupt signal distortions tightly aligned to arousal onset, which was not evident in our recordings. While we cannot completely exclude residual hemodynamic influences, we do not believe they account for the graded NE responses observed across vigilance-state transitions or the corresponding relationship with RR dynamics.

(2.3) Wordings such as 'extent of arousal' to compare MAs and wakefulness are problematic. Microarousals and wakefulness are qualitatively different behaviorally, physiologically, and in terms of neuromodulatory conditions

We thank the reviewer for the helpful comment. Our intention was not to imply that microarousals and wakefulness are qualitatively identical states differing only in magnitude. Rather, we used arousal in the broader neurophysiological sense, referring to the degree of activation of central arousal systems, with wakefulness representing part of this continuum. However, we recognize that in the sleep field, arousal is often used more specifically to describe brief EEG desynchronization events and that we furthermore use micro-arousals as a broad term for these sleep arousals, which may make our wording confusing. To avoid ambiguity, we have revised the manuscript to replace this terminology with vigilance state, sleep–wake state, or state transitions, depending on the context.

(2.4) To interpret linear correlations between datasets, even for the ones with Rsquare values < 0.5, as 'predictive' represents an overextended interpretation that is not supported by available evidence. This concern is aggravated due to the use of AUCs that confound amplitudes and time courses. This is in particular the case for Figure 5d, 5k, or 5m…

We thank the reviewer for this important comment. We agree that the term predictive may overstate the interpretation of these correlations, particularly given the modest R² values in some analyses. Our intention was to highlight an association between heartrate and NE-related measures rather than imply strong predictive performance or causality. We have therefore revised the wording throughout the manuscript to avoid predictive language and instead refer to these relationships as associations or correlations.

Regarding the use of AUC, we selected summary metrics based on the physiological characteristics of the signal of interest. In cases where responses were characterized primarily by rapid shifts to a new level, amplitude measures were used. In contrast, AUC was chosen when both the magnitude and duration of the response were considered physiologically relevant.

Part 3. A substantial number of experimental and analytical points require clarification. Here is a list of a few examples; many observations noted here apply equally to other figure panels.

(3.1) Many figure panels leave it open about whether averages or representative data are shown, how many animals are included, and what kind of measures are plotted.

We thank the reviewer for this comment and agree that clarity in figure presentation is important. In response, we have carefully revised the figure legends throughout the manuscript to more explicitly state whether data shown are representative examples or group averages, clarify the number of animals included in each analysis, and specify the measures being plotted. We have also added “data are shown as mean ± SEM” where this information was previously not explicitly stated and clarified when n refers to the number of animals. We believe the revised figure legends now provide clearer guidance for interpretation, and further statistical details are available in the accompanying statistics table. It should also be noted that number of animals used for all experiments can be found in the Method section ‘Mice’.

(3.2) In case experiments were done in a paired manner (e.g., the LC stimulations), individual data points should be shown connected for the different conditions.

We thank the reviewer for this suggestion. While we agree that connecting individual data points is valuable for paired experimental designs, this visualization is not appropriate for the LC stimulation analyses presented here. Specifically, each condition reflects pooled stimulation events selected across animals rather than a single summary value per animal that can be directly matched across columns. As such, individual points in one condition do not map one-to-one onto points in the next condition, making connected visualizations potentially misleading.

Importantly, although the data are presented as pooled event-level measures, the paired structure of the experiment was accounted for in the statistical analyses, such that repeated measurements within animals and the paired nature of the design were included in the relevant comparisons.

(3.3) Numerous analyses involve heart rate measures from the neck EMG during wakefulness. However, it is not specified how, in this case, RR peaks could be detected within the high-activity EMG.

We thank the reviewer for this question. The procedure for heart-rate detection during wakefulness is described in detail in the Heart rate detection Methods section. Briefly, RR intervals were extracted from preprocessed neck EMG recordings using a previously validated approach for mouse sleep studies. To minimize contamination from movement-related EMG activity during wakefulness, R-peaks were not detected within periods extending 100 ms before and 250 ms after detected movement, as these segments were considered too noisy for reliable peak detection. Heart-rate estimates during these excluded periods were subsequently interpolated using surrounding valid RR intervals to preserve temporal continuity and enable analysis of HR dynamics before and after movement episodes.

(3.4) Figure 1b: PSD for RR intervals. The supplementary figure says that 11-minutelong NREMS or 5-minute-long NREMS periods were used. These are very rare events in mice, for which the average bout duration is around 2 min and the cycle length is 10 minutes. The methods do not explain how these bouts were chosen, how many of them were included, and why shorter bouts were not analyzed. Single cases or means?

We thank the reviewer for this important point and agree that additional clarification was warranted. The selection of long NREM (including microarousals) periods was motivated by methodological considerations related to spectral analysis of very slow oscillations rather than by an assumption that these bout lengths are representative of average NREM duration in mice. Because our analysis focused on very-low-frequency dynamics (~1 cycle every 50 s), sufficiently long continuous recordings are required to reliably estimate power at these frequencies and avoid fragmentation-related edge effects that disproportionately affect shorter bouts.

For this reason, shorter NREM episodes were excluded from the PSD analysis, as they do not provide sufficient duration to robustly capture slow-frequency components. We initially compared PSD estimates using NREM periods of at least 11 min (allowing ~10 VLF cycles) and 5 min duration to assess whether the shorter 5 min recordings introduced bias (Supplementary Fig. 1f). While longer periods resulted in higher overall power estimates, the frequency distribution remained highly similar between conditions. We therefore selected 300 s (5 min) as the inclusion criterion for the remainder of the study, as periods exceeding 10 min are uncommon in mice and would substantially limit analyses across experimental paradigms.

To further reduce bias related to bout duration, PSD estimates were weighted by NREM episode length, as longer bouts showed systematic effects on power estimates. For Figure 1, the analysis included 144 NREM-with-MA bouts across 7 animals, and data shown represent group means rather than single examples.

This description is also included in the Methods section/Data Analysis.

(3.5) Figure panel 1c,d: In Panel c, what is plotted?

As stated in the figure legend, it is the cross-correlation between NE and R-R. We have added more description in the figure legend.

A cross-correlation between two signals? If yes, how were these signals chosen per transition? Looks rather like they plot some time course across a transition. What is time point 0?

We thank the reviewer for pointing out that this analysis was insufficiently explained. The analysis shown represents a cross-correlation between the NE and RR signals, performed to assess their temporal relationship across different vigilance-state transitions. Specifically, for each transition type, NE and RR signals were extracted within the corresponding time windows and cross-correlated to determine the strength and timing of their interaction.

In this context, time point 0 (lag = 0) represents perfect temporal alignment between the two signals. Positive or negative lags indicate whether changes in one signal systematically precede or follow changes in the other. Due to methodological differences between the rapid electrical heart signal and the slower fluorescent NE signal, we intentionally avoided overinterpreting fine temporal lead–lag relationships. Rather, the aim of this analysis was to characterize the overall interaction between the two signals across transitions.

Because the relationship between NE and RR was predominantly inverse, negative cross-correlation values indicate that increases in NE are associated with decreases in RR (i.e., faster heart rate), and vice versa. Thus, this analysis served primarily to confirm and extend our other findings by quantifying the interaction between NE and heart-rate dynamics across vigilance-state transitions.

We have revised the descriptive sentence in the Results section to improve clarity and have added a more detailed description of the signals included directly in the figure panel.

(Results, page 7). “Here, cross-correlation analysis revealed a predominantly negative relationship between NE and RR across vigilance-state transitions, indicating that increases in NE were associated with reductions in RR (i.e., faster heart rate; Fig. 1c).”

(Figure 1 legend). “Cross correlation (how strongly and at what temporal offset the two signals covary) between NE and RR during transitions”.

In Panel d, what is measured here? Is the time point of the dotted line a NA trough or the moment of a transition? Show the data with connected lines. Heart rate calculation during wakefulness?

We thank the reviewer for these questions and apologize that this was not sufficiently clear. In panel d, the dotted line indicates the NE trough, not the moment of a vigilance state transition, as specified in both the figure and figure legend. The analysis is aligned to detected NE troughs during sleep and examines the subsequent physiological dynamics, including transitions into wakefulness.

We have updated the sentence in the result section to make it a bit more clear:

(Results, page 7): “To explore the NE-RR relationship across sleep-wake transitions, we examined four progressive sleep-to-wake transitions using the preceding NE trough as the time stamp (time 0):…”

Regarding the suggestion to connect data points, as noted in an earlier response, these analyses are based on event detection, where multiple events contribute from each animal. Thus, each condition represents pooled events across animals rather than a single matched value per animal, making connected-line visualizations inappropriate and potentially misleading. Importantly, the paired structure of the experimental design was accounted for in the statistical analyses.

Heart-rate detection during wakefulness was usually not possible due to movement artefacts in the EMG. In the detection, we excluded periods with movement. Our EMG-based detection approach is described in the Heart rate detection Methods section. Briefly, movement-contaminated periods were excluded from R-peak detection (100 ms before and 250 ms after detected movement), and RR intervals were subsequently interpolated using surrounding valid values to preserve temporal continuity. Because analyses were aligned to NE troughs occurring during sleep, we limited the temporal window to avoid excessive contamination from movement-related noise associated with subsequent wakefulness. We have slightly revised the text to make these points clearer.

Panel C is a cross correlation between NE and RR from the same traces included in the mean traces. x=0 is the NE through like the other figures.

Panel D is the mean traces of NE during transitions with the dotted line at x=0 being the NE though

Yes, but x = 0 for the cross-correlation is not the NE trough. It says something about how aligned NE and R-R are in time (see response further up in (3.5)).

(3.6) The sigma band should ideally be chosen between 10-15 Hz for better consistency with the literature.

We thank the reviewer for this suggestion and agree that consistency in the definition of frequency bands is important for comparison across studies. The sigma range used in the present manuscript was selected to match our previous publications and analyses, thereby allowing direct comparison with our earlier findings on LC-mediated regulation of sleep spindles and infraslow sleep dynamics. Maintaining the same band definition also ensured consistency across the datasets analyzed in this study.

We acknowledge that having a shared definition of sigma power would improve alignment and facilitate comparisons across laboratories. At present, there remains some variability in the exact frequency boundaries used for spindle and sigma analyses across studies and species, although there are ongoing efforts within the sleep field to improve standardization. Importantly, we do not expect that modest adjustments of the sigma-band boundaries would materially affect the conclusions of the present study, as the spindle-related activity of interest lies well within the selected frequency range and the observed effects are broad rather than restricted to a narrow frequency bin.

Moving forward, we aim to follow emerging consensus recommendations where appropriate. To clarify this point for readers, we have added a statement in the Methods section explaining that the sigma band was chosen to maintain consistency with our previous publications.

(Methods: EEG power, page 29). “The sigma band was defined as 8 - 15 Hz to maintain consistency with our previous publications. Modest differences in sigma-band boundaries are not expected to affect the main conclusions.”

(3.7) What are 'extreme LC stimulation frequencies'. The only information available is that stimulations were done for 2s at 20 Hz.

We thank the reviewer for pointing out that this wording was unclear. By “extreme LC stimulation frequencies”, we did not refer to the within-stimulation pulse frequency (which remained constant at 20 Hz for 2 s across all conditions). Rather, we referred to the effective frequency of LC activation at the infraslow timescale, which was progressively increased through the closed-loop stimulation paradigm.

Specifically, stimulations were triggered when NE levels crossed increasingly permissive thresholds during the descending phase of the endogenous NE signal. As thresholds increased over time (from −15 ΔF/F (%) to +5 ΔF/F (%)), stimulations occurred progressively earlier in the infraslow cycle, thereby compressing the oscillatory period and increasing the effective frequency of LC recruitment while preserving the endogenous temporal structure of NE dynamics.

Thus, “extreme stimulation frequencies” refers to the highest rate of repeated LC activations achieved through the closed-loop paradigm, where stimulations became increasingly frequent at the infraslow level rather than changes in the 20 Hz pulse train itself. To avoid confusion, we have revised the wording throughout the manuscript to refer more explicitly to faster infraslow LC activation frequencies or increased infraslow stimulation frequency.

We have added more information in the result section to highlight this better.

(Results, page 8). “We employed a closed-loop paradigm, where LC stimulations (2 s 20 Hz (10 ms) blue laser pulses with a light intensity of 5 mW) were triggered when NE levels fell below increasing thresholds (-15, -10, -5, 0 and 5 ΔF/F (%), Fig. 2a-b, Methods). This approach enabled controlled compression of the infraslow NE cycle by triggering LC activation during the descending phase of the endogenous NE signal, thereby increasing the effective oscillatory frequency while preserving the temporal structure of physiological LC–NE dynamics. This strategy allowed us to test whether heart-rate responses continue to track LC-driven NE fluctuations as the infraslow rhythm becomes progressively faster.”

(3.8) Could the discrepancy between panels 4c, left and right, be due to limited sample size? The whole figure lacks indications of sample numbers, making interpretation difficult.

We thank the reviewer for this comment. We assume the reviewer is referring to the apparent discrepancy between the NE response and heart-rate response following LC suppression (Fig. 4c–d), where LC inhibition induced a clear reduction in NE levels, whereas mean heart-rate responses were less pronounced.

The figure legends state sample sizes and event numbers. Specifically, for these analyses, n = 8 animals (4 Arch, 4 YFP) were included, comprising 48 Arch events and 38 YFP events, our interpretation is that this discrepancy reflects a biological observation rather than a failed manipulation. Specifically, LC suppression robustly reduced NE levels, confirming the effectiveness of the optogenetic intervention, whereas HR did not exhibit a similarly consistent group-level response. However, as highlighted by the correlation analyses, variability in RR responses remained associated with the magnitude of NE suppression, suggesting that heart-rate dynamics still reflected noradrenergic modulation at the individual-response level despite the absence of a strong mean effect.

(3.9) It would be great if Figure 4 j could be more explicitly illustrated. For example, behavioral traces that lead to higher NFR and corresponding changes in RR AUC should be shown for animals with large and small effect sizes. The sample size seems excessively low. Can this explain the difference in slopes compared to Figure 4e, right panel?

We thank the reviewer for this suggestion. To improve the interpretation of Figure 4j, we have now added representative examples illustrating animals with high and low memory performance and their respective RR responses following LC suppression.

We agree that the sample size for this analysis is limited. As noted in the Methods, Figure 4j is based on a secondary analysis of a previously published dataset (Kjaerby, Andersen et al., 2022), where heart-rate measures were retrospectively extracted from EMG recordings. Due to noise-related limitations in RR detection, reliable cardiac measures could not be obtained from all animals, reducing the number of subjects available for this analysis. For this reason, we have deliberately avoided direct statistical comparisons between Arch and YFP animals and instead limited our conclusions to the observed association between RR responses and memory performance across animals.

Regarding the difference in slope compared with Figure 4e, the two analyses are based on different levels of aggregation and address different questions. Figure 4e examines the relationship between NE and RR responses across individual LC suppression events, resulting in multiple observations per animal. In contrast, Figure 4j uses a single mean RR response and a single behavioral outcome per animal. Furthermore, Arch and YFP animals were pooled in Figure 4j to maximize statistical power and because the dataset was not sufficiently powered for direct group comparisons. Consequently, the slopes are not expected to be directly comparable between the two figures.

(3.10) It would be important to show anatomical validation of viral expression in THcre animals and optic fiber positioning.

We thank the reviewer for this comment. Anatomical validation of viral expression in TH-Cre animals and optic fibre placement was performed for these experiments and has been reported previously in Kjaerby et al. Nature Neuroscience paper, from which this dataset was derived. Specifically, viral targeting and fibre positioning were histologically verified as part of the original experimental validation. This is mentioned in the Method section ‘Surgery’: Viral expression and injection sites were validated through immunostaining of perfused brain slices from the experimental animals (see Kjaerby et al. (3) for more information).

Version 2

Abstract

Recent work shows that the brain’s arousal system remains active during sleep, with rhythmic locus coeruleus (LC) activity shaping sleep architecture and supporting memory consolidation. The LC releases norepinephrine (NE) in infraslow (∼0.02 Hz) bouts that gate NREM sleep spindles. Here, we demonstrate that heart rate (HR) fluctuations during NREM are tightly phase-locked to these NE rhythms, identifying the LC as a key driver of very-low-frequency HR variability (VLF-HRV), an understudied autonomic signal. Using optogenetics, transient LC inhibition blunts HR slowing, whereas LC activation produces rapid HR acceleration, directly linking LC output to cardiac control during sleep. We further show that infraslow HR variability is a cross-species marker of spindle-dependent memory processing. In mice, the amplitude of HR decelerations during NREM correlate with spindle activity and subsequent memory performance. Remarkably, human sleepers show the same pattern: stronger VLF-HR fluctuations during NREM correspond to increased spindle expression and better overnight memory retention. These findings reveal a mechanistic pathway through which LC activity modulates autonomic physiology during sleep and identify infraslow HR variability as a non-invasive marker of brainstem function and memory-promoting sleep. Because LC degeneration occurs early in neurodegenerative disease, sleep-derived HR metrics may provide a scalable indicator of emerging neuromodulatory dysfunction.

Introduction

Heart rate variability (HRV) is a well-established, noninvasive biomarker of autonomic nervous system function and cardiovascular health. Reduced HRV is associated with increased morbidity and mortality and is frequently observed in neurological disorders such as Alzheimer’s disease, where it may reflect central autonomic network dysfunction (Natarajan et al., 2020; Zulli et al., 2005). Yet, despite its widespread use, decoding of heart rate signals and the central drivers of HRV remain elusive. One potential orchestrator of brain-body communication is the locus coeruleus-norepinephrine (LC-NE) system; LC is a brainstem nucleus that is part of the brain’s arousal systems. A recent discovery of an infraslow phasic pattern (∼30-50 s) of LC-NE activity during non-rapid eye movement (NREM) sleep has linked LC-NE activity with a wide range of critical cortical functions, including memory consolidation as well as glymphatic clearance (Hauglund et al., 2025; Kjaerby et al., 2022; Osorio-Forero et al., 2021). The ascending phase of NE fluctuations is often coupled to brief periods of intrinsic arousals known as micro-arousals (MAs) (Andrillon et al., 2011). In contrast, the declining phase of NE is associated with an increase in sleep spindles, which are discrete oscillatory events lasting 1 to 2 seconds in the sigma band (8–15 Hz) implicated in memory consolidation (Andrillon et al., 2011; Fernandez and Lüthi, 2020; Mednick et al., 2013). The LC-generated infraslow phasic fluctuations of NE lead to a microstructural organization of NREM sleep shifting between arousal and spindle-rich segments (Kjaerby et al., 2022; Osorio-Forero et al., 2021). Additionally, infraslow NE activity is involved in organization of ripple-dependent memory consolidation (Chang et al., 2025) and elicit vasomotion-dependent cerebrospinal flow (Hauglund et al., 2025), indicating a broader functional role of infraslow phasic NE fluctuations in sleep restoration. With age, LC-NE function deteriorates along with restorative NREM sleep and memory performance (Grudzien et al., 2007; Mann et al., 1980). Yet, the LC brain structure and physiological markers are difficult to access in humans. Thus, there is a need to explore more approachable biomarkers for sleep-dependent memory consolidation.

Sleep spindles, along with their role in memory consolidation, also show tight coupling with the peripheral autonomic nervous system (Chen et al., 2020; Naji et al., 2019a). Specifically, phasic bursts in heart rate (HR) during NREM sleep are preceded by surges in delta power (1-4 Hz) and sleep spindles, and this temporal coupling has been linked to memory improvements across sleep (Chen et al., 2020; Naji et al., 2019a). It has previously been reported that HR fluctuates at similar infraslow frequencies as NE fluctuations and sigma power in mice (Lecci et al., 2017; Osorio-Forero et al., 2021) and that phasic HR fluctuations correlate with infraslow changes in pupil diameter (Carro-Domínguez et al., 2025), a proxy for changes in NE levels (Murphy et al., 2014; Reimer et al., 2016). Importantly, optogenetic manipulation of the LC has demonstrated that infraslow LC activity coordinates sleep spindle clustering and heart rate fluctuations during NREM sleep (Osorio-Forero et al., 2021). Together, these findings support a functional coupling between the central LC–NE system and peripheral cardiac dynamics, further supported by findings that HR increases accompany MAs during NREM sleep (Carro-Domínguez et al., 2025; Osorio-Forero et al., 2025).

The sympathetic and parasympathetic autonomic nervous system are involved in HRV, which is conventionally analyzed across three primary frequency bands: high frequency (HF) HRV, low frequency (LF) HRV, and very low frequency (VLF) HRV (Berntson et al., 1997). Due to the frequency overlap with VLF HRV, we wondered if central infraslow NE dynamics could be linked to this poorly understood HRV indicator. Furthermore, are infraslow NE fluctuations directly reflected by HRV under different physiological states or does LC–HR coupling scales differently with LC output? Specifically, if infraslow NE oscillations display faster frequencies - as occurs during sleep fragmentation - will cardiac dynamics exhibit corresponding changes?

Conversely, given that stronger infraslow NE dynamics correlate with memory consolidation through their regulation of sleep spindles, could the peripheral autonomic signatures provide an accessible cross-species biomarker of spindle-dependent memory consolidation? Addressing these questions could help bridge mechanistic insights into LC-mediated sleep regulation with established HRV metrics used in human physiology.

In this study, we combined real-time fluorescent imaging of NE with simultaneous heart rate recordings in mice to test whether phasic infraslow LC–NE activity during NREM sleep is reflected in HRV dynamics. We further examined whether these brain–heart relationships change when LC–NE oscillatory activity is experimentally shifted toward faster frequencies using titrated optogenetic stimulation during sleep. Finally, we investigated whether cardiac dynamics associated with these neuromodulatory rhythms relate to spindle regulation and memory consolidation in both mice and humans.

Results

Norepinephrine and heart rate correlate during NREM sleep

Separate areas of research has demonstrated that both NE and HR signals fluctuate every 30-50s. Here, we used causal experimental methods to explore whether these independent infraslow fluctuations may be coupled and how they are impacted by different sleep transitions (Kjaerby et al., 2022; Osorio-Forero et al., 2025, 2021; Tanoue et al., 2022; Watson et al., 1979).

Toward this goal, we expressed the fluorescent NE biosensor GRAB-NE2m(3.1) in the medial prefrontal cortex (mPFC) of adult C57BL/6 mice and recorded extracellular NE using fiber photometry simultaneously with EEG and EMG recordings covering multiple sleep–wake episodes (Fig. 1a). mPFC was selected as a representative cortical readout of LC-mediated NE dynamics, as infraslow NE fluctuations are coordinated across widespread brain regions. R-peaks of the QRS complex of the heart were extracted from the EMG (for details on HR/RR extraction, see the Method section ‘Heart rate detection’ and Supplementary Fig. 1a-c) and the inter-beat-interval (RR) was calculated (Sörnmo and Laguna, 2005) (Supplementary Fig. 1d-e). A shorter RR interval reflects faster HR and vice-versa. HR during the NREM period was on average 493 bpm (SD ± 59) while REM HR was 519 bpm (SD ± 67) both of which was consistent with the expected 450-600 bpm, based on the literature (D’Souza et al., 2021; Ho et al., 2011; Sakata et al., 2005).

Norepinephrine and heart rate correlate during NREM sleep transitions.

a. Left: Experimental setup visualizing the fiber photometric acquisition of extracellular norepinephrine (NE) levels in the medial prefrontal cortex, EEG, and EMG recordings. Right: Example trace of NE, EEG, EMG, and RR intervals (RR) during NREM sleep (blue shade), microarousals (MAs) <15 s (yellow shade), and wake (gray shade). RR ranged from 0.0980 to 0.1439 (417-612 BPM). b. Power spectral density (PSD) of RR and NE from 0-0.1 Hz. c. Cross correlation (how strongly and at what temporal offset the two signals covary) between NE and RR during transitions from NREM sleep to NREM sleep (blue), short MAs (beige), long MAs (orange), and wake (pink). These color codes were used for the rest of the figure. d. Left: mean trace with standard error of the mean (SEM) of NE with yellow background shade for parts of the trace that are quantified through area under the curve (AUC) estimations. Right: AUC for time 0 to 25 s after the NE trough with AUC of -25 to 0 as baseline correction. e. Left: mean trace with SEM of RR during sleep transitions. RR ranged from 0.105 to 0.13 (462-571 BPM). Right: RR AUC for time 0 to 25 after the NE trough with AUC of -25 to 0 as baseline correction. f. Correlation between RR AUC and NE AUC across all sleep transitions. g. Mean and SEM of EEG power bands surrounding the NE trough event marker during sleep transitions. h. AUC for time 0 to 25 after the NE trough for EEG power bands with AUC of - 25 to 0 as baseline correction. i. Correlations between the AUC of RR and the EEG power bands. n=7 mice, NREM events = 245, NREM to short MA events = 174, NREM to long MA events = 75, NREM to wake events = 66. *p < 0.05., **p < 0.01., and ***p < 0.001. Data is shown as mean±SEM. All frequency domain analysis figures have visualizations based on weighted estimates (see methods). For a more detailed overview of the statistics, see section 1c-i in ‘Statistics’ under Supplementary Materials.

From the simultaneous recording of NE and EMG-derived RR, example traces suggested that there was an inverse correlation between NE and RR, where RR would decrease (rapid burst in HR) around the same time NE would increase (Fig. 1a). To highlight the potential temporal relationship between NE and RR, we conducted a frequency-domain analysis of the infraslow fluctuations in both signals (Fig. 1b, Supplementary Fig. 1e-f).

To explore the NE-RR relationship across sleep-wake transitions, we examined four progressive sleep-to-wake transitions using the preceding NE trough as the time stamp (time 0): continued NREM (lowest arousal), NREM to short MA (movement ≤ 4 s) transitions, NREM to long MA (movement 5 ≤ 15 s), and NREM to wake (movement > 15 s, highest vigilance state). Multiple transition events were extracted from each animal’s recording and pooled across animals for subsequent analysis. Cross-correlation analysis revealed a predominantly negative relationship between NE and RR across vigilance-state transitions, indicating that increases in NE were associated with reductions in RR (i.e., faster heart rate; Fig. 1c). Additionally, we found that the magnitude of NE ascent corresponded to progressively greater shifts in vigilance state during sleep (Fig. 1d). Similarly, when comparing the RR, we found a corresponding drop in values (faster HR) across state transitions (Fig. 1e). When correlating the NE and RR changes across the sleep transitions, a negative correlation was found (Fig. 1f), indicating that phasic fluctuations in NE and peripheral autonomic activity were highly correlated. These comparisons were performed using area under the curve (AUC) estimates of NE and R-R responses. Importantly, comparing AUC and peak amplitude measures showed a similar overall relationship, and the coupling between NE and RR remained significant when only peak amplitudes were considered (Suppl. Fig. 1g), indicating that the findings are not solely driven by the response duration aspect of AUC. The somewhat stronger relationship observed with AUC-based measures may reflect rapid saturation of heart-rate responses across vigilance-state transitions, making response persistence an informative component of the physiological signal.

To appropriately categorize the different vigilance events, we calculated the power of delta (1 - 4 Hz), theta (4 - 8 Hz), sigma (8 - 15 Hz), beta (15 - 30 Hz), gamma (80 - 100 Hz) bands from the spectrum of EEG signals surrounding NE throughs (Fig. 1g). Power in all bands, except gamma, decreased as the vigilance state increased (Fig. 1h). Inversely, as vigilance state increased, gamma also increased as reported before (Osorio-Forero et al., 2025). Lastly, changes in band power were correlated with the decline in RR across all sleep transitions. All EEG bands showed a positive correlation with RR, except for gamma, which was negatively correlated (Fig. 1i). Thus, these results indicate that NE-locked RR changes predicted EEG-defined vigilance states.

In conclusion, changes in cardiac rhythm are tightly coupled with the infraslow phasic NE fluctuations during NREM sleep, and HR accelerations may serve as a proxy for estimating temporal arousability dynamics during sleep. Interestingly, the systematic periodicity of this heart-brain coupling falls within the very low frequency (VLF) band (0-0.15 Hz for mice (Pizzo et al., 2022)), previously noted in the heart rate variability literature (Berntson et al., 1997; Tobaldini et al., 2013).

Closed-loop optogenetic activation of the locus coeruleus during NREM sleep increases heart rate

After finding the inverse correlation between NE and RR in natural sleep transitions, we next sought to characterize the causal influence of LC activity on HR dynamics, using a closed-loop optogenetic approach to modulate NE oscillatory frequency during NREM sleep. To this end, we designed a stimulation paradigm to modulate NE oscillatory frequency during NREM sleep. Specifically, we optogenetically stimulated the LC bilaterally in TH-Cre transgenic mice through Cre-dependent Channelrhodopsin (ChR2) activation. We employed a closed-loop paradigm, where LC stimulations (2 s 20 Hz (10 ms) blue laser pulses with a light intensity of 5 mW) were triggered when NE levels fell below increasing thresholds (-15, -10, -5, 0 and 5 ΔF/F (%), Fig. 2a-b, Methods). This approach enabled controlled compression of the infraslow NE cycle by triggering LC activation during the descending phase of the endogenous NE signal, thereby increasing the effective oscillatory frequency while preserving sleep continuity. This strategy allowed us to test whether heart-rate responses continue to track LC-driven NE fluctuations as the infraslow rhythm becomes progressively faster.

Closed-loop optogenetic activation of locus coeruleus during NREM sleep increases heart rate.

a. Experimental overview. Top: Channelrhodopsin (ChR2) was used to enable Locus Coeruleus (LC) stimulation through light activation. NE release was recorded with fiber photometry during simultaneous EEG/EMG recordings. Bottom: LC stimulation was activated when the real-time ΔF/F (%) calculations of NE levels fell below the pre-specified thresholds of -15, -10, -5, 0, and 5 ΔF/F (%) difference of the mean ΔF/F (%) from the last two minutes. b. Example trace of NE, EEG, EMG, and RR intervals (RR) from the first threshold (-15) period during NREM sleep (blue shade), microarousals (MAs) <15 s (yellow shade), and laser stimulation (light pink shade). RR ranged from 0.0940 to 0.1310 (458-638 BPM). c.-d. Mean traces with standard error of the mean (SEM) of NE (left) and RR (right) surrounding the onset of a laser burst with beige (pre-stimulation) and green (post-stimulation) background shade for all five thresholds: -15 (beige), -10 (light pink), -5 (pink), 0 (dark pink), and 5 (wine red) ΔF/F (%) difference. These color codes remain consistent throughout the figure. RR ranged from 0.1190 to 0.1370 (438 - 504 BPM). e.-d. Distributions of area under the curve (AUC) of NE (e.) and RR (d.) either 20-0 s before stimulation (left) or 0-20 s after stimulation (right) for all five thresholds. Baseline estimates were subtracted from the observed values at 20-40 s before and after stimulation respectively. g. Pre-stimulation AUC minus the post-stimulation AUC for NE (left) and RR (right) for all thresholds. h. Left: correlation of the pre-stimulation AUC minus the post-stimulation AUC for NE and RR across all thresholds. Right: correlation of the pre-stimulation AUC minus the post-stimulation AUC for NE and RR across thresholds -15, -10, and -5. i. From left to right: 1) Power spectral density (PSD) for RR between 0-0.2 Hz across all thresholds. 2) AUC of the RR PSD across all thresholds in the very low frequency (VLF) range. 3) Frequency at the peak of the PSD across thresholds for RR in the VLF range. 4) Power value at the peak of the PSD across thresholds for RR in the VLF range. Datapoint darkness is proportional to bout length. j. Example traces for R-R intervals during different LC stimulations. n=10 mice (6 ChR2). Threshold -15 events = 108 (97 ChR2), Threshold -10 events = 260 (205 ChR2), Threshold -5 events = 777 (481 ChR2), Threshold 0 events = 1444 (934 ChR2), and Threshold 5 events = 1148 (771 ChR2). Number of ChR2 bouts for frequency domain analysis were, from baseline to highest threshold: 212, 44, 41, 30, 37, and 43. Data is shown as mean±SEM. *p < 0.05., **p < 0.01., and ***p < 0.001. All frequency domain analysis figures have visualizations based on weighted estimates (see Methods). For a more detailed overview of the statistics, see section 2e-i in ‘Statistics’ under Supplementary Materials.

Mean traces of NE and RR were aligned to LC stimulation onset (Fig. 2c-d, for number of laser stimulations see Fig. 2 legend or Methods). As apparent from the traces, the stimulation became more frequent with higher NE thresholds, the NE baseline increased, while the RR baseline decreased (faster HR). The magnitude of pre-stimulation NE descent and post-stimulation NE ascent was reduced as the thresholds increased, indicating less pronounced NE dynamics as LC stimulation became more frequent (Fig. 2e, for direct comparison with YFP control, see Suppl. Fig. 2c-d). A reversed pattern was observed for the RRs (Fig. 2f). These laser-induced changes were consistent with subcortical arousal responses through power reductions for sigma and beta (Supplementary Fig. 3a-c) (Lüthi and Nedergaard, 2025). An inverse NE–RR relationship was present in both ChR2 (Fig. 2h) and YFP (Suppl. Fig. 3f) groups but was significantly stronger in ChR2 animals than in YFP controls (Suppl. Fig. 3h). When exploring the shifts in NE and RR surrounding the LC stimulations, the negative association between NE and RR plateaued beyond the 3rd stimulation threshold (also known as ‘5’, Fig. 2g-h). These findings indicate a causal relationship in which LC-mediated NE modulation affects HR; however, this relationship breaks down at higher infraslow LC activation frequencies. Since pre-stimulation NE baseline levels progressively became higher in the ChR2 condition compared with YFP controls (Fig. 2c+e, Supplementary Fig. 2a-f), it demonstrates that they arise from stimulation-dependent modulation of noradrenergic tone rather than nonspecific signal drift. As a result, the pre-stimulation state at higher thresholds differed between ChR2 and control conditions, which should be considered when interpreting the immediate effects of laser stimulation across thresholds. To understand if there was a mechanistic link between the PSD overlap between NE and RR observed in Fig. 1c, the PSD was calculated for RR across stimulation thresholds (Fig. 2i). As LC stimulation became more frequent (higher thresholds), the peak frequency of the VLF range increased, indicating that the VLF band might be driven by LC activations (Fig. 2i). Surprisingly, the overall amount of power for the PSD also increased without changes to the peak power level, suggesting that a wider range of fluctuations, and thus a wider frequency distribution, emerge as LC stimulation increases (Fig. 2i).

In conclusion, we demonstrate a causal relationship between LC and HR, showing that LC activations drive HR bursts and that HR reliably tracks infraslow NE oscillations under physiological conditions. However, as the frequency of infraslow NE fluctuations increases beyond the normal range, this relationship progressively weakens, indicating that HR primarily reflects LC–NE activity during intact, restorative infraslow sleep regimes.

Heart rate decelerations in response to sustained high norepinephrine levels

Next, we wanted to explore the unexpected widening of R-R frequency distribution in response to more frequent LC stimulation (Fig. 2i) in more detail. Although the mean trace of the RRs at higher LC stimulation thresholds did not reveal such dynamics, the variability of RRs was elevated compared to lower thresholds (Fig. 3a-b). Mean HR also increased as LC stimulation became more frequent, and this correlated with the variability of the RRs (Fig. 3c-d). Thus, HR fluctuations became more frequent with more stimulation of LC. This was also evident from individual raw RR traces (Fig. 3e).

Heart rate decelerations in response to sustained high cortical norepinephrine levels.

a. Mean RR intervals (RRs) surrounded by raw RR traces for the five thresholds: -15 (beige), -10 (light pink), -5 (pink), 0 (dark pink), and 5 (wine red) ΔF/F (%) difference. Color codes in this figure remain consistent. b. Standard deviation (SD) of the RRs +/- 30s around the LC stimulation (left). Sample entropy of RRs 10 s after LC stimulation (right). c. Mean RRs +/- 30s around the LC stimulation (left). Mean heart rate (HR) measured in beats per minute (BPM) +/- 30s around the LC stimulation (right). d. Correlation between HR and SD of RRs across stimulation thresholds. e. RR example traces surrounding laser onset. f. Mean RR traces surrounding HR decelerations. g. RR value at HR decelerations. h. HR at HR decelerations. i. RR amplitude leading up to HR decelerations. j. Correlation between HR and RR value at HR decelerations across thresholds. k. Norepinephrine (NE) mean-trace surrounding HR decelerations across thresholds (left). NE mean-trace surrounding HR decelerations at thresholds 0 and 5 (right). l. Sigma power mean-trace surrounding HR decelerations across thresholds (left). Sigma power mean-trace surrounding HR decelerations at thresholds 0 and 5 (right). m. NE amplitude leading up to HR decelerations. n. NE mean values leading up to HR decelerations. o. Correlation between mean NE values and RR values at HR decelerations across thresholds. p. Sigma power amplitude leading up to HR decelerations. q. Correlation between sigma power amplitude and RR values at HR decelerations across thresholds. Color codes for laser thresholds: -15 (beige), -10 (light pink), -5 (pink), 0 (dark pink), and 5 (wine red). n=6. Threshold -15 events = 97 ChR2, Threshold - 10 events = 205, Threshold -5 events = 481 ChR2, Threshold 0 events = 934 ChR2, and Threshold 5 events = 771 ChR2. Data is shown as mean±SEM. *p < 0.05., **p < 0.01., and ***p < 0.001. For a more detailed overview of the statistics, see section 3b-q in ‘Statistics’ under Supplementary Materials.

To quantify the HR decelerations that happened after LC activation, we took the maximal RR value (so slowest HR) 2-7 s after LC stimulation and baseline corrected the value to the mean RR during the 8–10 s period preceding the RR peak to obtain their amplitude. We found that the amplitude only differed from 0 in the higher stimulation thresholds (0 and 5) and correlated with mean HR (Fig. 3 f-j). Thus, it seems that at high LC stimulation, where HR is elevated, HR decelerations occur. This may reflect compensatory decelerations in response to sustained elevated sympathetic activity and thus HR (Airaksine et al., 2001). To understand the potential mechanism behind the HR decelerations, we explored NE and sigma power surrounding the HR decelerations (Fig. 3k-l). As LC stimulation frequency increased, general NE levels elevated, while infraslow NE fluctuations shrank (Fig. 3m-n). Notably, larger HR decelerations correlated with a higher preceding mean NE value, further supporting the interpretation that HR decelerations occurred when NE levels were high and sustained over time (Fig. 3o). Lastly, we observed that as LC activations increased, the sigma power amplitude in the seconds leading up to the HR decelerations shifted from negative to positive, indicating that HR decelerations were associated with sigma power increases (Fig. 3p). The sigma amplitude furthermore positively correlated with the RR value at the HR peak (Fig. 3q). This indicates that HR decelerations, despite high NE levels, are still associated with neuronal synchronization in the sigma frequency range, since NE can still display subtle fast declines (Fig. 3k).

In conclusion, frequent LC activations do not lead to sustained elevation in HR; rather, the heart exhibits compensatory HR decelerations, leading to large dynamic HR shifts following sigma changes.

Heart rate responses to locus coeruleus suppression reflect memory consolidation

To complement the experiments increasing infraslow NE frequency, we next tested how slowing the infraslow NE rhythm affected the association between RR and NE by expressing a Cre-dependent inhibitory opsin, Archaerhodopsin (Arch), in the LC of TH-Cre transgenic mice, with YFP serving as a control. We measured NE levels through an optic fiber implanted in mPFC in addition to EEG/EMG measurements (Fig. 4a-b). When exploring the NE and RR response to 2 min (4 min stimulus-interval) LC suppression during NREM sleep, we observed a gradual NE decline sustained throughout the laser period that was not observed in the YFP condition, while HR did not deviate between Arch and YFP (Fig. 4c-e). Notably, despite the absence of a robust group-level HR effect, NE and RR responses remained negatively correlated across trials (Fig. 4e), indicating that HR dynamics continued to track the magnitude of noradrenergic suppression at the individual-response level. A similar negative relationship was observed in YFP controls (Fig. 4e), and the strength of the NE–RR association did not differ significantly between Arch and YFP animals (Suppl. File, Statistics), suggesting that LC suppression did not substantially alter the underlying coupling between these measures. To better understand how LC suppression impacted the NE and RR oscillations, a frequency-domain analysis was conducted for NE and RR for YFP and Arch mice (Fig. 4f). When exploring the PSD for NE and RR (VLF), the analysis revealed no shifts in infraslow dynamics between the Arch or YFP condition for either NE or RR (Fig. 4f-g). This likely reflects the relatively mild and intermittent LC suppression paradigm, which was designed to remain within physiological ranges and therefore did not globally reorganize infraslow sleep dynamics.

HR modulation to LC suppression improves memory consolidation.

a. Experimental overview: Archaerhodopsin (Arch) or Yellow fluorescent protein (YFP) was injected into the locus coeruleus (LC) while GRAB-NE2m(3.1) was expressed in the medial prefrontal cortex (mPFC). Arch was stimulated through green light activation, leading to decreased norepinephrine (NE) release in the mPFC. NE release was recorded with fiber photometry during simultaneous EEG/EMG recordings. Mean HR across the recordings for the Arch group was 525 BPM (+/- 60 SD) or 0.1143 (+/- 0.0131 SD) s between beats and for the YFP group 517 BPM (+/- 61 SD) or 0.1161 (+/- 0.0137 SD) s between beats. b. Example trace of NE, EEG, EMG, and RR intervals (RR) during NREM sleep (blue shade), microarousals (MAs) <15 s (yellow shade), REM sleep (green shade), and wake (gray shade). RR ranged from 0.0949 to 0.1453 (413-632 BPM). c.-d. Mean trace with standard error of the mean (SEM) of NE (left) and RR (right) surrounding the laser onset event marker with purple shade to mark the area quantified through area under the curve (AUC) calculations for Arch-(blue) and YFP- (green) related animals. RR ranged from 0.1109 to 0.1220 (492-541 BPM). e. AUC of the NE trace (left) and RR trace (middle) under LC suppression with the two measures correlated (right) for Arch- (blue) and YFP- (green) related animals. f. Power spectral density (PSD) of NE (left) and RR (right) for YFP mice (green) and Arch mice (blue). g. From left to right: 1-2) AUC of NE PSD and RR PSD in the very low frequency (VLF) range (blue shade), 3-4) frequency at the peak of the NE and RR PSD, and 5-6) power at the peak of the NE and RR PSD for YFP mice (green) and Arch mice (blue). Datapoint darkness is proportional to bout length. h. Mean trace with SEM of theta power (left), sigma power (middle), and beta power (right) surrounding the laser onset event marker. Purple shade marks the area quantified through AUC calculations for YFP (green) and Arch (blue) animals. i. AUC for theta (left), sigma (middle), and beta (right) during laser activation for Arch- (blue) and YFP- (green) related animals along with their correlations with RR AUC for the Arch condition. j. Left: Overview of the Novel Object Recognition task. Mice were habituated to the environment before neural recordings and were then tested in the same environment with one object being replaced with a novel one. Middle: Representative RR interval traces from a high novel-to-familiar ratio (NFR) mouse and a low NFR mouse during inhibition. Right: NFR correlated with RR AUC, and RR AUC correlated with NE AUC for Arch- (blue) and YFP- (green) related mice. n=8 (4 Arch). Arch events = 48, YFP events = 38. PSD bouts: Arch = 66, YFP = 1. 43. Data is shown as mean±SEM. *p < 0.05., **p < 0.01., and ***p < 0.001. All frequency domain analysis figures have visualizations based on weighted estimates (see methods). For a more detailed overview of the statistics, see section 4e-j in ‘Statistics’ under Supplementary Materials.

Lastly, we were interested in understanding the relationship between HR during LC suppression and memory consolidation. As sleep spindle density during NE decline correlate with memory performance (Kjaerby et al., 2022), we wanted to understand if changes in HR would correlate with post-sleep performance and thus if the HR/NE coupling during this phase is important for memory consolidation. We confirmed a sigma power upregulation and general neuronal synchronization during LC suppression. Furthermore, we found a correlation between RR responses to LC suppression and sigma power, suggesting that a stronger HR reduction response is linked to higher spindle power; similar correlations were also observed in the theta and beta frequency ranges (Fig. 4h-i). To test whether HR changes were associated with memory outcomes, we performed a novel object recognition task. The mice were habituated to an environment with two objects prior to the LC suppression paradigm and were subsequently re-introduced to the environment after the recording, but now with one known object exchanged for a new one. By measuring approaches to the new vs. old object, the novel-to-familiar ratio was extracted and used as a behavioral measure of memory consolidation (Fig. 4j). Interestingly, across pooled Arch and YFP animals, larger RR increases following LC suppression were associated with better subsequent memory performance. Additionally, RR and NE responses to LC suppression were negatively correlated indicating that animals showing stronger NE reductions also exhibited larger RR changes. Together, these findings suggest that heart-rate dynamics covary with noradrenergic responses during sleep and may reflect physiological processes relevant for sleep-dependent memory consolidation.

Sigma power surges prior to heart rate burst predict next day memory performance

To further understand the cross-species relationship between shifts in autonomic activity and memory consolidation, we explored sigma power dynamics surrounding phasic HR accelerations, also known as HR bursts (HRBs, increases in HR larger than 2 or 2.2 SD of the mean in windows of 180 or 24 s for humans or mice respectively) (Chen et al., 2022b, 2022a, 2020; Naji et al., 2019a, 2019b), in both humans and mice. In mice, sigma power has been previously shown to inversely correlate with NE fluctuations and predict memory performance (Kjaerby et al., 2022). As expected, NE increases in mice preceded phasic HRBs and peaks at the time of the HRB (Fig. 5b-c). Additionally, the amplitude of the NE increases associated with the strength of the HRBs (Fig. 5d). We showed that sigma power dropped directly prior to HRBs, consistent with our expectation given the negative correlation between sigma power and NE in prior work (Kjaerby et al., 2022) (Fig. 5e-g). A similar analysis was performed in a human dataset, where participants underwent memory testing and EEG/ECG recordings overnight. While the relationship between sigma power and RR intervals was conserved across species, we observed a temporal shift in the human data (Fig. 5h). Here, the surge in sigma power preceding the HRBs was confined to the 5-s interval prior to the HRB (Fig. 5i-j), while the sigma for mice peaked in the 15-10 s range prior to the event. These temporal differences may reflect species-specific physiology differences in cardiac timescales, which has also been previously reported (Bergel et al., 2025; Carro-Domínguez et al., 2025; Lecci et al., 2017). When analyzing the relationship between the pre-HRB sigma power and the slowing of the heart following the HRB in humans, we observe a positive correlation, suggesting the importance of the pre-HRB sigma power for HR baseline recovery (Fig. 5k). Lastly, we measured improvements in hippocampal-dependent, episodic memory through the word-pair association (WPA) task before and after a night of sleep (Fig. 5l). Sleep-dependent memory improvement (post-pre-difference) was associated with pre-HRB sigma power (baseline-corrected, p = 0.05), suggesting that surges in spindle activity linked to HRBs were important for memory consolidation (Fig. 5m).

Pre ‘Heart Rate Burst’ Sigma surges linked to improved memory.

a. Example trace showing norepinephrine (NE), sigma power, and RR-interval (RR) during NREM sleep (blue shade) and heart rate burst (HRB) events (gray lines). b.-d. Mean trace with standard error of the mean (SEM) of RR (left), NE (right), and sigma power (below) surrounding the HRB events. e-g. Correlation between the NE and RR amplitude (left), Sigma power and RR amplitude (middle), and Sigma and NE (right) amplitude. h. Example trace of human RR and sigma power during NREM stage 2 sleep (pink shade) during HRB events (gray lines). i. Mean trace of RR intervals surrounding HRB events. Purple shade represents the time during which the RRrecovery amplitude was measured. j. Mean trace of sigma power intervals surrounding HRB events. Purple shade represents the sigma power interval quantified in following sub-figures. k. Correlation between sigma power pre-HRB and RR amplitude following the HRB 1. l. Overview of the word-paired association (WPA) memory task. m. Correlation between memory improvement and baseline corrected pre-HRB sigma power. Mice: n=7, HRB events = 920. n. Schematic diagram showing the relationship between phasic locus coeruleus and heart rate events, which are followed by norepinephrine descends and sigma power upregulations correlating with memory performance. Hypothesis figure: Humans: n=28, HRB events = 4753. Data is shown as mean±SEM. *p < 0.05., **p < 0.01., and ***p < 0.001. For a more detailed overview of the statistics, see section 5d-m in ‘Statistics’ under Supplementary Materials.

In conclusion, mice and humans displayed evidence of conserved autonomic-central coupling, reflected in coordinated HR and sigma power dynamics surrounding phasic cardiac events.

However, the temporal relationship between these events differed across species, likely reflecting differences in sleep and cardiovascular physiology. In mice, causal manipulation of LC activity demonstrated that HR dynamics covary with LC-mediated noradrenergic and spindle-related processes linked to memory consolidation. In humans, sigma power preceding HR bursts was associated with both post-HRB heart-rate slowing and sleep-dependent memory improvement, suggesting that autonomic–central coupling surrounding HR events may provide a translational marker of restorative sleep processes (Fig. 5n).

Discussion

Variability in HR is a noninvasive biomarker of autonomic nervous system function and is frequently disrupted in aging and Alzheimer’s disease. Here, by combining real-time measurements of mPFC NE dynamics with simultaneous HR recordings in freely sleeping mice, we demonstrate that HR closely tracks the infraslow phasic activity of the LC–NE system. Specifically, we show that infraslow NE oscillations are reflected in HRV within the VLF range, providing mechanistic evidence that this previously poorly understood HRV component indexes central noradrenergic dynamics. When infraslow NE activity shifts toward faster frequencies, as observed during sleep fragmentation, HR dynamics follow these changes only within a limited range, after which compensatory mechanisms begin to obscure the coupling, suggesting a physiological boundary to brain–heart synchrony. Importantly, we further demonstrate that HR predicts the memory-relevant interaction between NE activity and sleep spindle organization in both mice and humans, identifying HRV as an accessible peripheral marker of sleep-dependent memory consolidation. Together, these findings bridge LC-mediated regulation of sleep microstructure with established HRV metrics and support VLF-HRV as a translational readout of restorative sleep processes and central noradrenergic function.

Heart rate is modulated by infraslow norepinephrine fluctuations during NREM sleep

Utilizing HR detection and fiber photometry, we found that NE and HR correlate across NREM sleep transitions. To determine causality, we utilized TH-Cre mice with ChR2/YFP injected in the LC to activate the nucleus. We found that LC stimulations were associated with increases in HR, implying that NE levels mediate HR changes.

Many studies have linked HR to NE (Fawaz and Simaan, 1963; Sundaram et al., 1991; Tanoue et al., 2022; Watson et al., 1979). Many rely on plasma levels of NE, which, while linked to central NE (Gurguis and Uhde, 1998), has low temporal resolution, making causal interpretations harder. In recent years, the use of biosensors and fiber photometry allows for very reliable estimate of the temporal dynamics of NE changes making association to HR more precise (Osorio-Forero et al., 2021). The periodicity of phasic fluctuations of LC and HR overlaps and corresponds with previous findings from HR studies, namely those centered on exploring HRV. HRV is a well-established metric within the health sciences literature (Berntson et al., 1997), with greater HRV consistently associated with a broad spectrum of positive health and cognitive outcomes, many of which are linked to frontal lobe functioning (Forte et al., 2019). HRV is conventionally analyzed across three primary frequency bands: high frequency (HF), low frequency (LF), and very low frequency (VLF) (Berntson et al., 1997). While HF HRV and LF HRV and their relation to autonomous functioning have been characterized (Berntson et al., 1997; Malik et al., 1996; Shaffer et al., 2014), the physiological mechanisms underlying VLF HRV remain poorly understood (Bonaduce et al., 1994; Brämer et al., 2019; Kitney, 1980). The frequency range for VLF in humans has been reported to be ≤0.04 Hz (Berntson et al., 1997; Tobaldini et al., 2013) but have in mice been extended to ≤0.15 Hz (Pizzo et al., 2022) due to the faster HR dynamics.

Here, we show that VLF HRV power overlaps with NE infraslow phasic fluctuations (0-0.1 Hz) in rodents, and recent evidence from pupil response in humans suggests that this overlap is conserved cross-species (Carro-Domínguez et al., 2025). Additionally, stimulating the LC pushed VLF HRV towards higher frequencies, suggesting that one potential driving force of VLF HRV is LC-mediated NE infraslow oscillations. Prior findings demonstrate that LC is linked to the autonomic nervous system. Stimulation of LC projections decrease parasympathetic cardiac vagal activity (Wang et al., 2014) and also influences sympathetic output through direct projections to the preganglionic cells in the sympathetic nervous system (Karemaker, 2017; Nygren and Olson, 1977; Samuels and Szabadi, 2008). While the mechanisms generating VLF HRV are not well defined (Armour, 2003; Shaffer et al., 2014; Wang et al., 2014), there is a clear parasympathetic component (Taylor et al., 1998). This combined with the ability of pharmacological blockage of the parasympathetic system to block infraslow oscillations of HR (Osorio-Forero et al., 2021) and pupil diameter during sleep (Yüzgeç et al., 2018) led us to expect a slowing of HR during LC suppression due to parasympathetic disinhibition. Interestingly, we found no consistent modulation in HR during LC suppression, suggesting that the LC-HR connection may also partly be driven by sympathetic outflow. Previous research had indicated that LF power may also represent sympathetic activity, but this interpretation has been challenged due to the mixed contribution of both autonomic branches (Houle and Billman, 1999; Japundzic et al., 1990; Reyes et al., 2013). LF and HF ratio (LF/HF) were traditionally thought to reflect balance between sympathetic and parasympathetic activities (i.e., the sympatho-vagal balance), though currently considered as an oversimplification of non-linear integration of autonomic signals (Billman, 2013; Pagani et al., 1986). Our findings implicate the VLF may offer a precise marker for central arousal states, given its overlaps with infraslow phasic fluctuations of LC-NE levels. These comparisons were performed important for the expression of VLF oscillations, growing evidence suggests that VLF dynamics reflect broader interactions between the heart and autonomic nervous system rather than simple sympathetic or parasympathetic control alone (Armour, 2003; Shaffer et al., 2014). Within this framework, infraslow LC–NE dynamics may represent one central contributor to these slow cardiac fluctuations during sleep either directly through sympathetic activation or indirectly by inhibition of cardiac vagal activity. Furthermore, heart rate and HRV may be a promising highly accessible read-out of LC activity, which has important clinical implications.

Heart rate as a cross-species memory marker

Spindle density, implicated in sleep-dependent memory consolidation, increases during periods of NE decline (Kjaerby et al., 2022). Specifically, the infraslow descent of NE creates spindle-rich periods. Now we report a correlation between memory consolidation and coupled HR-NE decrease during LC suppression, suggesting that HR can be used as a biomarker of memory consolidation in sleep. This finding was extended to humans, where the sigma power surges prior to heart rate bursts (HRBs) predicted sleep-dependent memory improvement. In mice, we saw that the HRBs co-occurred with a peak in NE, which had been on the rise prior to HR changes. Since NE changes cause HR alterations, this suggests that HRBs are triggered by increases in NE signaling in mice. Thus, a similar NE increase may exist in humans (Carro-Domínguez et al., 2025).

Our findings also suggest that the pre-HRB surge in sigma power could play a role in memory consolidation. Previous work in humans demonstrated that the strength of infraslow sigma oscillations correlates with sleep-dependent memory consolidation in humans (Lecci et al., 2017). Thus, given that HRB follows the infraslow NE rhythm, the specific timing of the sigma power relative to the HRB might be important for memory consolidation. These findings replicate past findings (Naji et al., 2019a) in daytime sleep and further generalize them to overnight sleep and thus overnight memory changes.

Interestingly, when comparing mouse data to human data, we see temporal differences in the sigma power leading up to the HRB. The mouse sigma peaks around 10 s earlier and has more oscillatory dynamics as opposed to the stark rise in sigma power pre-HRB in humans. This pattern has previously been reported (Bergel et al., 2025; Carro-Domínguez et al., 2025; Lecci et al., 2017) and has been attributed to species-specific differences in anatomy and physiology in the neural coupling to the heart (Lecci et al., 2017). These species-specific differences could involve intrinsic cardiac neuron (ICN) properties and cardiac neural circuitry. A comparative study shows that mouse ICNs exhibit more phasic firing and direct cholinergic projections to the sinoatrial node, enabling faster and more oscillatory heart-brain coupling, while human ICNs are predominantly tonic, supporting slower, more sustained modulation (Tompkins et al., 2024). These anatomical and physiological differences likely contribute to the earlier and more dynamic sigma peak observed in mice compared to humans and further research efforts to map cross-species heart-brain interactions may reveal critical differences that can be used in translational techniques.

Importantly, infraslow fluctuations of sigma power in human sleep have received growing attention. Recent work demonstrates that the infraslow fluctuation of sigma power segments N2 sleep into functional phases associated with arousal and memory-related sleep markers (Dimitriades et al., 2024). Complementary findings using pupillometry show that pupil diameter fluctuates on similar infraslow timescales during NREM sleep and is inversely related to spindle clustering, providing indirect evidence that arousal-related noradrenergic dynamics shape human sleep microstructure (Carro-Domínguez et al., 2025). However, LC activity during human sleep remains inferred rather than directly measured, and systematic perturbation studies linking LC output to spindle–autonomic coupling in humans are currently lacking. Together, these observations underscore both the promise and the current limitations of cross-species comparisons of infraslow sleep dynamics.

In summary, these findings imply that HR serves as a surrogate marker for NE in humans as well as a biomarker for memory consolidating processes during sleep. Loss of LC neurons occurs with general aging and is an early marker of Alzheimer’s disease (Grudzien et al., 2007; Mann et al., 1980). Furthermore, maintaining LC neural density prevents neurodegeneration (Wilson et al., 2013). Given the reported reduction in HRV in aging and Alzheimer’s disease (Natarajan et al., 2020; Zulli et al., 2005), suppressed LC responses might be responsible for the blunted autonomic cardiac modulation during sleep. The link to HRV could be further investigated by neuroimaging studies quantifying LC volume (atrophy) and determines whether the extent of LC loss could be a possible predictor of future cognitive decline through its relationship to sleep spindles. Indeed, evidence suggests that the LC activity during sleep is compromised (e.g., increased arousals, fragmented sleep, elevated cortisol) and linked to memory failure in both aging (Mander et al., 2017, 2013) and Alzheimer’s disease (Lim et al., 2013).

Locus-coeruleus activation drives compensatory heart rate decelerations via autonomic feedback mechanisms

A surprising finding was that diminishing the power of NE fluctuations by repetitive LC activation did not lead to less power in the HR fluctuations; rather, the opposite happened. This was caused by large decelerations in HR in conditions with repetitive LC activation and high HR. We propose that the deceleration in HR during repeated LC stimulation may be explained by a dynamic balance between heightened sympathetic control due to increased LC activity and compensatory baroreflex responses. This is likely occurring as NE activates alpha-1 receptors on the smooth muscle in blood vessels, leading to vasoconstriction and increased blood pressure, triggering the baroreflex (Tanoue et al., 2002; Webb, 2003). This baroreflex signals to the nucleus of the solitary tract, which activates inhibitory neurons. These neurons suppress the activity of the rostral ventrolateral medulla, ultimately resulting in reflex parasympathetic activation with sympathetic inhibition lowering the HR (Aicher et al., 2000; Lanfranchi and Somers, 2002). An interesting finding was that the heart rate decelerations at high NE levels were associated with increases in neuronal synchronization (i.e., increased sigma power), which seemed to be closely linked to HR decelerations and smaller NE fluctuations. It is possible that the LC receives autoinhibitory feedback from the baroreflex responses, but that this is not reflected in our cortical NE recordings.

Overall these findings suggest that infraslow phasic fluctuations of LC-NE can drive large HR fluctuations consistent with the VLF range. However, when NE levels reach artificially high and sustained levels, compensatory mechanisms such as baroreceptor reflex activation may intervene. This adaptive feedback could serve as a safety valve, reducing the coupling between central NE activity and cardiac oscillations to prevent excessive or destabilizing cardiovascular responses.

Conclusion

We show that LC-mediated NE release plays a role in generating VLF HRV. HR appears to be more sensitive to stimulation than inhibition of the LC, indicating a potential role for NE in modulating cardiac function. These results may point toward some contribution of sympathetic activity in addition to the well-known parasympathetic influence, though further investigation is needed to determine the precise autonomic mechanisms involved. Furthermore, we demonstrate that shifting NE fluctuations during NREM sleep toward faster frequencies, the LC-HR coupling starts to disassociate due to obscuring compensatory mechanisms. Lastly, we conclude that HR changes through their known association with sigma power reflect memory consolidation processes in mice and humans. This research opens doors to unexplored possibilities of using HR as a non-invasive, cost-effective biomarker not only for sleep quality and LC-NE function, but potentially also for conditions of cognitive decline associated with sleep disturbances.

Methods

Mice

This study included female and male mice, which were group-housed according to sex with unlimited access to food and water. The housing environment had a 12-hour light/dark cycle at 21°C with 40-60% humidity. The data for all mouse-based experiments were acquired in connection to past publications (Kjaerby et al., 2022). For experiments associated with sleep transitions, seven C57BL/6 (wild-type) male mice were used. The heterozygous TH-Cre transgenic mice used for optogenetic experiments were bred on a C57BL/6 background. Ten mice were included in the LC activation experiments, four of which were controls (2 males/2 females) and six of which were ChR2 (4 males/2 females), and nine were involved in the LC suppression experiments. Of the nine mice, four were put in the control condition (4 females) and four Arch (3 females/1 male). One Arch mouse was excluded because it took over one-third of the recording period to enter NREM. The number of mice included in the analysis was lower than in previous publications due to the exclusion of animals with poor EMG signal quality. All experiments received approval from the Danish Animal Experiments Inspectorate and were supervised by the University of Copenhagen’s Institutional Animal Care and Use Committee (IACUC), in accordance with the European Communities Council Directive of 22 September 2010 (2010/63/EU) on the protection of animals used for scientific purposes.

Surgery

Mice underwent surgery at 7-15 weeks of age. General anesthesia was induced with 5% isoflurane and maintained at 1-3% isoflurane. Additionally, 0.05 mg/kg of preoperative buprenorphine was administered for general analgesia, and 0.03 mg/kg of lidocaine was injected at the incision site. A stereotactic frame was used to ensure stability and alignment throughout the surgery. The fur was trimmed from the upper part of the skull, and a cut was made between the ears. The skull was cleaned with ethanol, and an electric drill (Tech2000, RAM Microtorque) was used to make 3-5 burr holes in the skull following the stereotactic coordinates relative to the bregma.

In the mPFC burr hole, AAV9-hSyn-GRAB-NE2m(3.1) (see Table 1 for supplier information) was injected under the neuronal hSyn promoter (stereotactic coordinates: A/P +1.7 mm, M/L −0.3 mm, and D/V −2.00, −2.25, −2.50, and −2.75 mm, with 125 nL of virus injected at each level). Additionally, for mice included in the optogenetic experiments, an AAV5/2 (Adeno-associated virus 5/2) encoding floxed Arch (ssAAV5/2-shortCAG-dlox-Arch3.0-eGFP-dlox-WPRE-SV40p(A)) or ChR2 (ssAAV5/2-hEF1a-dlox-hChR2(H134R)-eYFP(rev)-dlox-WPRE-hGHp(A)) was injected bilaterally into the LC of the transgenic TH-Cre mice (stereotactic coordinates: A/P −5.5 mm, M/L ±0.9 mm, and D/V −3.75 mm, with 300 nL of virus injected). YFP (AAV5-EF1a-DIO-eYFP) was injected in control animals. For both virus injections, the needle was kept in place for 7 minutes after the injection (speed: 100 nL/min) before slow withdrawal.

For EMG acquisition, two wires from W3 Wires International were placed into the musculus trapezius. For EEG acquisition, two 0.8 mm stainless steel screws from NeuroTek with low impedance were screwed into holes above the cerebellum (reference point) and the frontal cortex. Contralateral to the EEG screws, a mono fiber-optic cannula (400 μm, 0.48 NA, Doric Lenses), attached to a metal ferrule (2.5 mm diameter), was placed in the mPFC at D/V −2.50 mm. For mice that had received Arch, ChR2 or YFP injections in the LC, dual fiberoptic cannulas (200 μm, 0.22 NA, Doric Lenses) were implanted bilaterally above the LC at D/V −3.65 mm. Connectors linked to the EEG screws and EMG wires, along with the cannulas, were fixed to the head with SuperBond dental cement.

Mice received 5 mg/kg of carprofen (s.c.) prior to waking up and two days after as post-op care. They were given at least two weeks post-operation for recovery and virus expression before experimentation. The surgeries were conducted following established institutional guidelines. Viral expression and injection sites were validated through immunostaining of perfused brain slices from the experimental animals (see Kjaerby et al. (Kjaerby et al., 2022) for more information).

Fiber photometry

Two light-emitting diodes (LEDs) with wavelengths of 465 nm and 405 nm (Doric Lenses, Tucker Davis Technologies) were connected to a minicube (Doric Lenses) using attenuator patch cords (400-μm core, NA = 0.48, Doric Lenses). The fiber-optic patch cords were secured to the fiber implants on the animals’ heads through zirconia sleeves. The minicube, equipped with dichroic mirrors and cleanup filters, enabled the detection of fluorophores by matching the relevant excitation and emission spectra. The LEDs were controlled by LED drivers (Thorlabs, Doric, Tucker Davis Technologies) and interfaced with an RZ-5 or RZ10-X real-time processor (Tucker-Davis Technologies). In the mPFC, both the 465 nm and 405 nm excitation lights were transmitted through a single patch cord to activate GRAB-NE2m(3.1) fluorescence, with the 405 nm light serving as an isosbestic control wavelength to adjust for bleaching and movement-related signal fluctuations. The 465 nm and 405 nm excitations were modulated sinusoidally at 531 Hz and 211 Hz, respectively. The two modulated signals generated by the LEDs were each retrieved using standard synchronous demodulation techniques, which were implemented on the RZ-5/RZ10-X real-time processor at a sampling rate of 1000 Hz. Signal processing was managed using Synapse software (Tucker-Davis Technologies), which also synchronized the fluorescent signals with video and EEG/EMG signals via TTL pulses. The resulting data were then exported for further analysis in MATLAB R2023b (MathWorks).

To calculate ΔF/F, the 405 nm signal was fitted to the 465 nm signal using a first-degree polynomial regression (MATLAB polyfit function). The slope and intercept obtained from the regression were used to generate a scaled 405 nm control signal. This fitted control signal was then smoothed using a moving average filter (MeanFilter) with an order of 1000 to reduce noise. The ΔF/F was calculated by normalizing the 465 nm signal against the smoothed control signal, subtracting the fitted 405 nm signal from the 465 nm signal, and dividing by the fitted 405 nm signal. The resulting normalized signal was expressed as a percentage to represent the change in fluorescence relative to baseline levels.

Locus coeruleus optogenetics

To induce LC excitation, two-second stimulations consisting of 10 ms pulses at 20 Hz (resulting in a 40 ms break between pulses) at a 5-mW intensity were applied bilaterally to the LC using a 430-490 nm laser (Shanghai Laser & Optics Century Co., Ltd., model: BL473-200FC). A two-hour baseline without stimulation was implemented before and after the five-hour stimulation paradigm. The simulations were conducted in a closed loop, where stimulation was triggered if the smoothed, real-time calculated NE ΔF/F (%) crossed the threshold based on an average ΔF/F (%) value of the last two minutes. The threshold was initially set to −15 ΔF/F (%) but increased by 5 every hour until reaching the final level of +5 ΔF/F (%). Throughout the baseline and stimulation paradigm, fiber photometry recordings from the mPFC, along with EMG and EEG, were acquired. Recordings took place during the light phase to encourage natural sleep behavior.

To induce LC inhibition, two-minute Arch stimulations at 5 mW intensity, followed by a four-minute break, were applied bilaterally to the LC using a 532 nm laser (Changchun New Industries Optoelectronics Tech. Co., Ltd., model: 16030476). Simultaneously, GRAB-NE2m(3.1) recordings were made at the mPFC site, along with the previously described EEG and EMG measurements. The two-hour recordings took place during the light phase when the mice were expected to exhibit natural sleep/wake behavior undisturbed. The mice were recorded with an infrared camera.

Novel object recognition task

The NOR task was performed on 12-20 weeks old mice involved in the LC suppression experiments in an open-field arena (PVC, custom-made, 40 x 42 x 40 cm). The mice were habituated to the environment for 10 minutes as part of the encoding phase, where the mice familiarized themselves with two objects. Following this, the animals underwent the LC suppression paradigm and were then given five minutes in the NOR task arena for the recall phase to determine the novel-to-familar ratio. In the recall phase, one of the two known objects had been exchanged for an unknown object and the ratio of approaches to the novel vs. familiar object was recorded. Natural sleep was encouraged between the encoding and recall phases by performing the task during the light phase in an environment where the mice had been habituated for over 12 hours prior to the experiment. The mice were tracked using Ethovision XT 11.5 (Noldus).

EMG signal preprocessing

The R-peak detection algorithm was applied to EMG recordings, that were priorly subjected to filtration in frequency domain using in-house pipeline in MATLAB. First, all the signals were subjected to removal of frequencies associated with amplitude current using a bandstop infinite impulse response (IRR) Butterworth 4th order filter between frequencies of 59 and 61 Hz. In case of recordings acquired in US, the cut-off frequencies for the notch filter were of 60 to 64 Hz, and the filter order of 2 was used due to different frequency span of the nuisance signal. Subsequently, a high-pass IIR Butterworth filter with 0.6 steepness was applied to the frequencies <2.5Hz to gradually remove fluctuation of the EMG signal baseline from the recordings. This was followed with low-pass filtration >80Hz, using the same filter but of 0.9 steepness. The steepness’s of the filters were chosen to minimally induce artefacts at the border conditions, and either gradually (high-pass) or rapidly (low-pass) remove the signals that affect detecting R peaks due to either increased fluctuations of the baseline (i.e., insufficient grounding or slow motion) or high frequency noise (relatively high spectral power of muscle signals). In a few recordings, both low-pass and high-pass frequencies were adapted to be slightly above 4Hz (but not higher than 5 Hz) and 45Hz respectively, to better remove the signal baseline fluctuations or higher amplitude high-frequencies due to presence of artefacts significantly affecting the R peaks detection. This was, however, applied in recordings of low signal-to-noise ratio in the frequency range of HR signals, and was carefully evaluated for appropriateness in all cases, so that the bandwidth of the signals within the range of heart pulsations was preserved. For that purpose, at each step of the signal preprocessing, periodograms of the filtered and unfiltered signals were computed and compared to evaluate the effects of the applied filtrations. After filtration, all signals were saved in the .mat format and were used for further R peaks detection.

Heart rate detection

R-peak detection in mice was based on preprocessed EMG recordings to obtain HR without needing additionally invasive procedures. EMG is acquired in most sleep studies to aid sleep scoring, and the method has previously been validated in Lecci et al. (Lecci et al., 2017) and subsequently used for other publications (Osorio-Forero et al., 2025, 2021). User settings for this function included adjusting the movement threshold and peak prominence to ensure optimal detection for different traces. The MATLAB function first determines the movement threshold based on the overall mean and SD of the trace. It then runs peak detection in windows of two seconds of data at a time. The threshold for the peaks is based on the mean and SD within each window to ensure peak detection even in cases of power loss. If movement is detected within a section, data that falls 100 ms before and 250 ms after the movement is not considered for peak detection due to noise and is also excluded from the peak threshold calculation within the window.

When all windows are analyzed, the timestamps at which R-peaks have been found are processed through a diff function to determine the time between peaks. The original time vector is maintained to ensure that the times of the RR are preserved for event extraction. To reduce any artifacts in the data, RR that suggested a HR faster than 800 BPM or slower than 400 BPM were removed, as previous findings suggest that the HR typically falls within 450-650 BPM throughout a 24-hour cycle (D’Souza et al., 2021; Ho et al., 2011; Sakata et al., 2005). Lastly, the data and associated time vector were interpolated to 64 Hz using the p-chip method, as in Chen et al. (Chen et al., 2023), since this method was found to have the lowest error when analyzing the frequency domain (Benchekroun et al., 2023). Interpolation was performed to obtain a consistent sampling rate, but also to get RR interval estimates during periods of movement based on the values leading up to and after the movement periods.

EEG power

For the mouse data, the power within each EEG band was determined by applying the spectrogram function to extract the short-time Fourier transformation of the EEG in one-second windows for frequencies between 0 and 100 Hz. To obtain frequency power bands, the absolute log of the short-time Fourier-transformed data was taken, and the mean values within each frequency band (slow oscillations: 0.5 - 1 Hz, delta: 1 - 4 Hz, theta: 4 - 8 Hz, sigma: 8 - 15 Hz, beta: 15 - 30 Hz, gamma (high): 80 - 100 Hz) were extracted. The sigma band was defined as 8 - 15 Hz to maintain consistency with our previous publications. Modest differences in sigma-band boundaries are not expected to affect the main conclusions.

Mouse sleep scoring

The mice were recorded in chambers using ViewPoint Behavior Technology, with EEG and EMG electrodes connected to a commutator (Plastics One, Bilaney, SL12C) through cables. Habituation in the recording environment lasted at least 12 hours before recordings, during which the EEG and EMG cables were connected. The mice were attached to fiber optic implants immediately before the recording, which lasted for 2–9 hours during their light phase. The EEG and EMG signals were amplified using a model 3500, 16-channel AC amplifier from National Instruments. A bandpass filter was applied to the EEG at 1-100 Hz and to the EMG at 10-100 Hz. Both signals had an added notch filter at 50 Hz to reduce power line noise. The EMG and EEG were digitized through a Multifunction I/O DAQ device from National Instruments (USB-6343) and sampled at 512 Hz. An infrared camera from FLIR Systems was used to record the mice to aid in scoring vigilance states.

Sleep scoring for mice was performed manually through visual inspection of the EEG and EMG traces, along with the video, to divide the traces into 5-second and, later, 1-second windows. Scoring was performed using SleepScore software from ViewPoint Behavior Technology. Vigilance states included ’wake’ (high-amplitude EMG and high-frequency, low-amplitude EEG), ’NREM sleep’ (low EMG amplitude and low-frequency, high-amplitude EEG), and ’REM sleep’ (low EMG amplitude and high-frequency, low-amplitude EEG). Later in the analysis pipeline, wake periods between 5-15 s were classified as long MAs, and wake periods below 5 s were scored as short microarousals.

Event marker selection

All time-domain analyses were based on the timestamps of event markers in the data. To determine the NE trough timestamps during the sleep transition analysis, similar approaches were used for the different stages. Across all stages, the NE trace was inverted, and peak detection was performed on the data. For the NE troughs during NREM, a peak prominence criterion of 0.7 to 3 was set. For NREM to MA (short and long), only data from 30 s prior to the transition were considered, and the peak prominence criterion was set to 0.8 to 1.8. Additionally, there had to be at least 15 s between troughs, with the one closest to the transition being favored. Lastly, for transitions to wake, data from 15 s prior to the scored transitions were considered. The peak prominence criterion was set to 0.1 to 0.4, and the peak height was set to a minimum of the mean of the trace minus five. After automatic trough detection, manual verification was performed across all transitions to ensure optimal trough detection.

The event marker for the data presented in Fig. 2 and 3a-b was the onset of the stimulation burst. As the laser had an online measure of NE and turned on for 2 s at 20 Hz, the event marker was defined as the point when the laser turned on after being off for over 100 ms. These timestamps were divided into groups based on when the threshold shifted. For the rest of Fig. 3, the RR peak value 2-7 s after onset of stimulation burst was utilized as the event marker for HR decelerations. The event marker for the data presented in Fig. 4 was set to when the laser turned on. For both LC excitation data, the exclusion criteria required that the animal be in NREM for 30 s before and after the laser turning on to be included and for LC suppression that the animal was in NREM for 30 s before and 60 s after laser onset.

For figure 5, the event marker was HRBs. These events have previously been reported in humans, where they are defined as when the RR intervals drop below 2 SD of the mean in a 3 min window (Chen et al., 2020; Naji et al., 2019a). This event marker was for mice adjusted to be 2.2 SD from the mean in 24 s windows. This adjustment was done based on differences in dynamics. The time window was adjusted based on the differences in mean HR between humans and mice.

Data analysis

Fiber photometry, EMG, and EEG data were extracted from all mice, and RR was calculated for each recording using the R-peak detection algorithm. For analysis pertaining to the sleep transition data, the EEG data were excluded from further analysis for two animals in the sleep transitions analysis due to noise, resulting in 245 events for NREM to NREM transitions (191 with EEG), 174 events for NREM to short MA transitions (133 with EEG), 75 events for NREM to long MAs (54 with EEG), and 66 events for NREM to wake transitions (42 with EEG) across the seven mice.

For the LC activation-related analysis, 108 events were found for Threshold -15 (11 of these being YFP), 260 events for Threshold -10 (55 of these YFP), 777 events for Threshold -5 (296 of these YFP), 1,444 events for Threshold 0 (510 of these YFP), and 1,148 events for Threshold 5 (377 of these YFP) across ten animals (four being YFP). For the LC suppression experiment, 48 events were found for the four Arch mice and 38 for the four YFP mice. All data preprocessing was performed using MATLAB. For the mouse data related to HRB events, 1654 events were found across the 7 mice also analyzed in connection with sleep transitions.

For all four experiments, the event variables were generated, and data from RR, NE, and the EEG frequency bands were extracted from 60 s before and after each event. Each data type was averaged across all events from all mice within each relevant group for each experiment, leading to the standard error being based on the number of events rather than the number of animals.

Across all experiments and data types, the AUC was calculated as a measure of change between the different conditions. For the sleep transitions experiments, the AUC from -25 to 0 s (zero being the event time) was used as a baseline and was subtracted from the AUC from 0 to 25 s after the event. For the ChR2/YFP experiment, three types of AUC measures were used: the pre-stimulation AUC was defined as the data from -40 to -20 s before the event (baseline) subtracted from the data from -20 to 0 s before the event. Similarly, the post-stimulation AUC was defined as the data from 20 to 40 s before the event (baseline) subtracted from the data from 0 to 20 s after the event. The pre-stimulation minus the post-stimulation AUC was defined as the difference between the two measures outlined above. Lastly, the AUC for the Arch/YFP-related experiments were defined as -40 to -5 s before the event marker (baseline) subtracted from 5 to 40 s after the event marker.

The data presented in Fig. 3 was based on the same +/- 60 s data files surrounding the LC stimulation onset as seen in Fig. 2, though only ChR2 animals were included in this analysis, as it focused on HR changes in response to stimulation. In Fig. 3a, mean traces across thresholds were shown and 97 raw traces were randomly extracted for each threshold to visualize variability, as this was the lowest number of total events present within a threshold. In Fig 3b, the RR, BPM and RR SD estimate was based on the full 2 min data file, while the sample entropy (m = 2) only included the 10 s after laser onset (Richman et al., 2004). The following analysis was based on the HR deceleration event marker. Here, the RR amplitude was calculated as the RR at HR decelerations minus the mean RR 8-10 s prior. Similarly, the NE amplitude was determined by subtracting the lowest NE value 5-10 s before the HR decelerations from the NE value at the HR decelerations. The mean NE values also reflected the NE values in the 10 s leading up to the HR decelerations. Lastly, the sigma amplitude was determined as the mean sigma 2-3 s before the HR decelerations subtracted from the sigma value at the HR decelerations.

For the analysis related to HRB events, amplitudes were analyzed rather than AUC to use similar measures as in the human data. For the mouse data, the RR amplitude was defined as the RR value at the HRB minus the mean across the 15 to 10 s leading up to the event. The NE amplitude was calculated as the peak value in the 2.5 s interval around the HRB event minus the minimum in the 15 to 10 s interval preceding the event. Lastly, the sigma amplitude was defined as the smallest observation in the 2.5 s interval around the HRB event minus the largest value found between 15 to 5 s range preceding the event.

The frequency-domain analysis was based on the non-parametric Welch approach, and similar procedures were used on NREM data across all experiments. The data underwent detrending with a 5th degree polynomial fitted to the raw data before analysis as in Ferreira et al. (Ferreira et al., 2017). Across experiments, frequencies between 0-0.1 Hz with a frequency resolution of 0.0002 Hz were analyzed for NE, and frequencies between 0-0.4 Hz with a resolution of 0.002 Hz were calculated for RR. Across all frequency-domain analyses, consecutive periods of minimally 300 s of NREM sleep (with MAs) were included for RR/NE, as this was the minimal NREM period duration suggested to capture VLF (Shaffer and Ginsberg, 2017). However, as the ability to interpret VLF power within short term recordings has previously been questioned (Thayer et al., 2022), we also extracted NREM sleep periods of minimum 11 minutes (allowing 10 cycles of VLF 0-0.015 Hz) to visualize an overlap in the frequency distribution (Supplementary Fig. 1f). Given the large similarities between the frequency distributions, we chose to utilize the minimum of 300 s NREM as the data selection criteria, as it is uncommon for mice to remain in NREM sleep for over 10 min, making frequency-domain analyses across different experimental paradigms challenging. Each bout of data was run through Welch separately, and its AUC, peak power value, and frequency at peak power were extracted along with information about the length of the period (bout length). Bout length was subsequently used to weigh the final mean/SEM estimates, which was utilized both for statistical analysis and visualization generation. SEM was based on the number of bouts. The PSD presented in Fig. 1 and Supplementary Fig. 1 were based on the same data from the mice used to analyze sleep transitions and held 144 bouts. For the LC activation data, the PSD was made separately for ChR2/YFP and for each stimulation threshold including the baseline before and after stimulation. The number of bouts for the ChR2 data (from baseline to threshold 5 were 212, 44, 41, 30, 37, and 43. For YFP, these numbers were 147, 24, 28, 21, 27, and 28. For the LC suppression data, the PSD was split by Arch/YFP, which consisted of 66 and 43 bouts respectively.

Statistical analysis

Linear mixed-effects models (LMMs) were utilized across all figures to estimate the effects using R. One exception was made for models related to memory estimates, as only one estimate was available per night per subject. Thus, a standard linear model was employed instead. The choice of applying LMMs to the data was made, as the model type allowed for event-based analysis by adding a random intercept for subject ID. By doing so, a hierarchy was created in the data to ensure that issues with independence did not occur (Yu et al., 2022). When continuous predictors were present in the model, the predictor was added as a random slope and compared to the base model (with only a random intercept) to determine which model was best using the Akaike and Bayesian information criteria. If the random slope model had a lower Akaike/Bayesian information criterion, converged, and did not have a singular fit, the statistics were summarized based on that model.

Models correlating sleep transitions with an array of dependent variables were built by setting the four sleep transitions as independent variable factors, which were re-leveled to ensure that the lowest level of arousal (NREM to NREM) was set as the baseline. For models associated with Fig. 2-3, the thresholds were similarly set as a factor, with -15 as the baseline threshold, except for the frequency-domain analysis, where the baseline was set as the intercept. The thresholds and conditions (ChR2 or YFP) were both set as predictors in Fig. 2 and had an interaction effect to explore the differences between the two conditions at different thresholds. In Fig. 3, the thresholds were the main predictor utilized, as only ChR2 data was included. Lastly, models for Fig. 4 had a similar predictor setup with condition (YFP or Arch) as the predictor. All LMMs had SubjectID as a random intercept.

After fitting the LMMs, they were assessed for any violations of the fixed effects assumptions through visual inspection of the residual distributions. While LMMs have been found to be robust to assumption violations (Schielzeth et al., 2020), bootstrapping of 1000 samples was performed to obtain robust estimates of the model in cases where one or more assumptions had been violated. For models where autocorrelation of residuals was present, block bootstrapping was chosen over normal bootstrapping. The block bootstrapping resampled the data in groups based on subject ID, which was the recommended solution to issues with autocorrelations of residuals. Additionally, when relevant to the analysis, weights were added to the LMMs and factored into the bootstrapping to ensure the bout length was appropriately weighed in the model.

For models with a continuous independent variable, the marginal and conditional R squared values were reported in the figure and was also the measure that was bootstrapped (Efron, 1983; Nakagawa and Schielzeth, 2012). In cases where the predictor had more than two levels or there were multiple predictors, the estimates of the pairwise comparisons were bootstrapped. Lastly, in cases where only a single predictor with two levels was present, the difference between model estimates for the two levels was bootstrapped. All pairwise comparisons were adjusted for multiple comparisons using Turkey’s method.

Participants

28 human participants (Mage=20.46 ± 2.73 years, 17 females) were included in the analysis of HRB. For information regarding demographics, see the supplementary table. Participants had no history of psychological, neurological or chronic illnesses and were previously included in another publication (Chen et al., 2021). All participants provided informed consent approved by both the Western Institutional Review Board and the Human Research Review Board at the University of California, Riverside. Individuals were excluded from participation if they exhibited irregular sleep/wake cycles, had diagnosed sleep disorders, personal or familial histories of psychopathology, substance abuse or dependence, a history of epilepsy or loss of consciousness exceeding two minutes, current psychotropic medication use, or cardiac or respiratory illnesses potentially influencing cerebral metabolism. Eligibility was determined during an in-person psychiatric assessment conducted by trained research personnel. Additionally, participants underwent a medical evaluation, including a detailed medical history review and physical examination by a staff physician, to confirm overall physical health. Participants were instructed to avoid caffeine, alcohol, and all stimulants for at least 24 hours prior to and during the experimental day. Compliance with consistent sleep patterns for one week prior to each experimental session was verified using sleep diaries completed daily and wrist-based activity monitors (Actiwatch Spectrum, Philips Respironics). Participants were required to achieve a minimum of 7 hours of sleep per night during this period, including the night immediately before the study day. Monetary compensation and/or course credits were provided to participants in acknowledgment of their time and participation.

Word-paired association task

Human participants underwent the word-paired association task, which consisted of an encoding, criterion and testing phase. In the encoding phase, participants were presented with 60 word-pairs one at a time for 1 s with an inter-trial interval of 1s. During the criterion phase, the participants were probed with one of the words from a pair and were subsequently presented with the missing word for 2 s followed by a 100 ms inter-trial interval. This process continued until the participants obtained 70% accuracy overall. Both the encoding and criterion phase occurred in the morning on the first day. Lastly, the testing phase was conducted three times: shortly after the criterion phase, in the evening before participants slept in the lab and the following morning. At each test, participants were tested on their memory for 20 of the half word-pairs. Participants were shown one word and asked to recall the missing word. No feedback was provided during the testing phase and there was a 1 s inter-trial interval between pairs. Performance was measured as the difference score between the first test and the final test approximately 24 hours later.

Heart rate detection in humans

In humans, ECG data were recorded using a modified Lead II Einthoven configuration at a sampling rate of 1,000 Hz. Heart rate variability (HRV) analysis was conducted on the R-wave intervals across the entire sleep-wake period using Kubios HRV Analysis Software 2.2 (Biosignal Analysis and Medical Imaging Group, University of Kuopio), following Task Force guidelines (Task Force of the European Society of Cardiology the North American Society of Pacing Electrophysiology, 1996). The R-wave peaks within the ECG QRS complex were automatically detected by Kubios software and subsequently verified visually by trained technicians. Any incorrectly identified R-peaks were manually adjusted, and missing beats were interpolated using cubic spline interpolation. After computing inter-beat intervals, a third-order polynomial detrending filter was applied to eliminate trend-related components in the data. Artifacts were further corrected using Kubios software’s built-in automatic medium-level artifact correction filter.

Human sleep recordings

EEG data were collected using a 32-channel electrode cap (EASEYCAP GmbH) equipped with Ag/AgCl electrodes arranged according to the international 10-20 system (Jasper, 1958). Among these electrodes, 22 were utilized for scalp EEG recordings, while the remaining electrodes were dedicated to ECG, electromyogram (EMG), electrooculogram (EOG), ground, an online common reference channel located at FCz (retained post-referencing), and mastoid references (A1 and A2). EEG signals were recorded at a sampling rate of 1,000 Hz and subsequently referenced offline to the contralateral mastoid electrode (A1 and A2).

Data preprocessing was performed using BrainVision Analyzer 2.0 software (BrainProducts). Eight scalp electrodes (F3, F4, C3, C4, P3, P4, O1, O2), along with EMG and EOG recordings, were selected for scoring nighttime sleep stages. EEG and EOG signals were filtered with high-pass and low-pass settings of 0.3 Hz and 35 Hz, respectively. Raw data were visually staged in 30-second epochs into Wake, Stage 1, Stage 2, Slow Wave Sleep (SWS), and REM sleep, according to the criteria of Rechtschaffen and Kales, using the custom MATLAB-based scoring toolbox, HUME. Following staging, epochs containing artifacts and arousals were visually identified and removed prior to spectral analysis.

Human data analysis

HRBs were used as the event marker for human data and was defined as when the RR intervals drop below 2 SD of the mean in a 3 min window (Chen et al., 2020; Naji et al., 2019a). This resulted in 4753 events across the 28 subjects during stage 2 NREM sleep. For these events, the RRRecovery amplitude reflected the mean RR values 5 to 10 s after the HRB minus the RR value at the event-time. The sigma power (12-15 Hz) trace was extracted and its value in the 5 to 0 s preceding the event was used to correlate to the RRRecovery. To correlate the sigma values to memory estimates, we followed the same procedure as past papers(Naji et al., 2019a). Sigma power was extracted in +/- 20 s around the HRB and then segmented into four-time bins: 10 to 5 s preceding the event, 5 to 0 s preceding the event, 0 to 5 s following the event, and 5 to 10 s following the event. For each participant, sigma power was averaged across all events within each bin, resulting in one mean value per bin. To correlate sigma power before the HRB and memory performance, baseline sigma power was subtracted from the sigma power 5 s before the HRB, following the equation below. This baseline represents mean sigma power during NREM sleep outside of 20 s windows surrounding event extracted data. All statistical tests followed the same principles outlines in the ‘Statistical Analysis’ section. Sigma (bl. corr.)=Sigma power−baselineSigma power+baseline